---
title: "Get started with bmm"
output: bookdown::html_document2
author:
  - Gidon Frischkorn
  - Ven Popov
pkgdown:
  as_is: true
---

```{=html}
<style type="text/css">
div.main-container {
max-width: 850px !important;
}

p {
margin-top: 1.5em ;
margin-bottom: 1.5em ;
}
.author{
    display: none;
}
</style>
```

```{r, include = FALSE}
options(crayon.enabled = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "jpeg",
  dpi = 100,
  fig.asp = 0.8,
  fig.align = "center"
)
fansi::set_knit_hooks(knitr::knit_hooks, which = c("output", "message", "error"))
options(width = 300)
```

# What bmm does

`bmm` fits cognitive measurement models to behavioral data: models whose
parameters stand for cognitive processes, such as the precision of a memory
representation or the sensitivity in a recognition task. You write the model
for each parameter with the formula syntax of `brms`, the R package for
Bayesian regression, and `bmm` translates the measurement model into a
distribution that `brms` can pass to Stan, the sampler behind it. In other
words, you get hierarchical Bayesian estimates of interpretable parameters
with the same formula interface you already use for regression.

This page takes you from an installed package to a first fitted model. It has
three parts: checking that your machine can fit models, finding the model that
matches your task, and fitting one model end to end.

# Install and check your machine

The released version is on CRAN, the development version on GitHub:

```r
install.packages("bmm")

# development version
if (!requireNamespace("remotes")) {
  install.packages("remotes")
}
remotes::install_github("popov-lab/bmm", upgrade = "never")
```

`upgrade = "never"` keeps `remotes` from updating the packages `bmm` depends
on. Updating `Rcpp`, `StanHeaders` or `RcppParallel` underneath an installed
`rstan` can leave `rstan` unable to load, in particular on Windows. This
matters even if `cmdstanr` does the sampling, because `brms` uses `rstan` to
store every fit.

Fitting a model needs more than the package: a C++ compiler, and a Stan
backend, which is the software that compiles the model and samples from it.
There are two backends, `cmdstanr` with CmdStan and `rstan`; we recommend
`cmdstanr`. These are the parts that fail most often on a fresh machine, and
the error usually shows up only when the first model refuses to compile.
Therefore, run `bmm_setup()` before anything else:

```r
library(bmm)
bmm_setup()
```

The function checks each requirement, prints `PASS`, `FAIL` or `SKIP` per
check, and gives one fix for every check that failed. It installs nothing
itself. With the default `smoke_test = TRUE`, it finally compiles and samples
a small mixture model, which is the only check that shows the whole chain
works. This is what the report looked like on the machine that built this
page; we skipped the smoke test here, so its line reads `SKIP` instead of
`PASS`:

```{r setup-report}
library(bmm)
bmm_setup(smoke_test = FALSE)
```

If a check fails, apply the fix it prints, restart R, and run `bmm_setup()`
again. Once every check passes, you are ready to fit models.

# Which model fits your task

You come to `bmm` with data from a task, so this section is organized by task.
Each entry links to the article that walks through the models `bmm` provides
for the task, and a table lists every model and version with the data columns
it needs and the parameters it estimates. In the data columns, the name before
the colon is the argument of the model function that you set to the name of
your column. Parameters listed as fixed by default are held at a constant
value; `?modelname` gives that value and what the parameter does. The tables are
generated from the package, so they show the models of the version this page
was built from.

## Continuous reproduction

Participants reproduce a continuous feature, such as a color or an
orientation, and the response error on the circle is the dependent variable.
The [continuous reproduction article](https://popov-lab.github.io/bmm/articles/bmm_vwm_crt.html)
describes the task and compares the models.

```{r, echo = FALSE}
knitr::kable(bmm:::model_overview(group = "Continuous reproduction"))
```

## Categorical recall and n-AFC decisions

Participants pick one response from a set of categories, for example the
correct item, an item from another list position, or an item that was not
studied. The [M3 article](https://popov-lab.github.io/bmm/articles/bmm_m3.html)
shows how to set up the response categories and the activation model.

```{r, echo = FALSE}
knitr::kable(bmm:::model_overview(group = "Categorical recall and n-AFC decisions"))
```

## Detection, recognition and confidence judgments

Participants decide whether a signal was present or an item was studied, rate
their confidence, give Remember/Know judgments, pick the target among several
alternatives, or rank the alternatives. Signal detection theory separates
sensitivity from response bias for each of these response formats, and `bmm`
has one model per format. The models take response counts per subject and
condition. `sdt_yn()` also accepts one row per trial, as the first fit below
shows; the rating, ranking and Remember/Know models need one count column per
response category, which `aggregate_sdt_cdp_data()` builds for `sdt_cdp()`
from trial-level data whose confidence runs on one old/new scale. The
[signal detection article](https://popov-lab.github.io/bmm/articles/bmm_sdt.html)
walks through the yes/no, rating, m-AFC and ranking models, and the
[dual-process and meta-d′ article](https://popov-lab.github.io/bmm/articles/bmm_sdt_dualprocess_metad.html)
through the `dpsdt` and `metad` versions of `sdt_rating()` and through
`sdt_cdp()`.

```{r, echo = FALSE}
knitr::kable(bmm:::model_overview(group = "Detection, recognition and confidence judgments"))
```

The models differ in three respects. The response format is what the
participant produces. The `dist` argument names the noise distribution assumed
for the evidence on which the decision is based; the first entry in the table
is the default of that model, and `?modelname` explains the others. The last
column names the parameter that lets the signal distribution differ in spread
from the noise distribution, if the model has one. This table is built from the
package, so a model appears once it is part of the version this page was built
from.

```{r sdt-table, echo = FALSE}
sdt_models <- list(
  list(format = "yes/no detection, old/new recognition", model = "sdt_yn",
       uv = "`sdratio`"),
  list(format = "confidence ratings", model = "sdt_rating",
       uv = "`sdratio`"),
  list(format = "m-alternative forced choice", model = "sdt_mafc",
       uv = "none"),
  list(format = "ranking", model = "sdt_ranking",
       uv = "`sdratio` (normal noise only)"),
  list(format = "old/new confidence with Remember/Know judgments",
       model = "sdt_cdp", uv = "`sigmar` (recollection)")
)
available <- supported_models(print_call = FALSE)
rows <- lapply(sdt_models, function(m) {
  if (!m$model %in% available) return(NULL)
  f <- formals(get(m$model, asNamespace("bmm")))
  versions <- if (!is.null(f$version)) eval(f$version) else NULL
  model <- paste0("`", m$model, "()`")
  if (!is.null(versions)) {
    model <- paste0(model, ", versions ", paste0("`", versions, "`", collapse = ", "))
  }
  data.frame(
    `Response format` = m$format,
    Model = model,
    `dist` = paste(eval(f$dist), collapse = ", "),
    `Unequal-variance parameter` = m$uv,
    check.names = FALSE
  )
})
knitr::kable(do.call(rbind, rows))
```

## Choices and response times

Participants choose between two options and the response time is recorded on
every trial. The [DDM article](https://popov-lab.github.io/bmm/articles/bmm_ddm.html)
and the [censored shifted Wald article](https://popov-lab.github.io/bmm/articles/bmm_cswald.html)
cover trial-level data. The [EZ-diffusion article](https://popov-lab.github.io/bmm/articles/bmm_ezdm.html)
covers the case where you only have the mean and variance of the response
times and the accuracy per condition. Fast guesses and other contaminant
responses are the topic of the
[response time contamination article](https://popov-lab.github.io/bmm/articles/bmm_rt_contamination.html).

```{r, echo = FALSE}
knitr::kable(bmm:::model_overview(group = "Choices and response times"))
```

## Processing-tree models

Participants give categorical responses, and you describe the discrete
processes that lead to each response as a multinomial processing tree, for
example the two-high-threshold model of recognition or the pair-clustering
model of free recall. You write each tree's branch probabilities as
expressions, and `mpt()` fits them to response counts. The
[MPT article](https://popov-lab.github.io/bmm/articles/bmm_mpt.html) covers
specifying trees, importing model files from MPTinR and TreeBUGS, and
constraining parameters.

```{r, echo = FALSE}
knitr::kable(bmm:::model_overview(group = "Processing-tree models"))
```

```{r other-groups, results = "asis", echo = FALSE}
known <- c(
  "Continuous reproduction", "Categorical recall and n-AFC decisions",
  "Detection, recognition and confidence judgments", "Choices and response times",
  "Processing-tree models"
)
registry <- bmm:::model_registry()
other <- setdiff(unique(registry$group), known)
if (length(other) > 0) {
  cat("## Other models\n\n")
  cat("These models were added after this page was written:\n\n")
  print(knitr::kable(bmm:::model_overview(group = other)))
}
```

You can list the models in R at any time with `supported_models()`, and
`?modelname` (for example `?sdt_yn`) documents the data a model expects, its
parameters and its default priors, that is, what the model assumes about each
parameter before it sees the data.

# Fit your first model

We fit a yes/no signal detection model to the recognition data of Broeder and
Schuetz (2009, Experiment 3), which ships with the package as
`broeder_schuetz_2009_e3` (see `?broeder_schuetz_2009_e3` for the source). 40
participants judged items as old or new under five base-rate conditions. A
higher proportion of old items makes participants more willing to say "old",
so the conditions shift the response criterion while sensitivity should stay
the same.

## The data

`sdt_yn()` expects one row per subject, condition and stimulus type, with the
number of "old" responses and the number of trials in that cell:

```{r}
head(broeder_schuetz_2009_e3)
```

- `stimulus` codes the stimulus type: `1` for old items, `0` for new items.
- `n_old` counts the "old" responses in the cell. For old items these are the
  hits, for new items the false alarms.
- `n_trials` is the number of items in the cell, which differs between cells
  because the base rate manipulates it.
- `condition` runs from the most conservative (`br1`) to the most liberal
  (`br5`) base rate. The base rate is set through the number of old and new
  items: in `br1` a participant saw 6 old and 54 new items, in `br5` 54 old
  and 6 new.

## The model

A model object names the columns the model needs. Nothing is estimated yet;
this only tells `bmm` where to look:

```{r}
model <- sdt_yn(
  response = "n_old",
  stimulus = "stimulus",
  n_trials = "n_trials"
)
```

If your data has one row per trial instead of counts, you can pass it as is:
the column named in `response` then holds `0` or `1`, the column named in
`n_trials` holds `1` on every row, and `stimulus` codes the item of that
trial as before. We leave `dist` at its default, `"normal"`.

The model has two parameters. `d` is the sensitivity: the distance between
the evidence distributions of old and new items. `criterion` is the response
bias: the point on the evidence axis above which participants say "old". A
third parameter, `sdratio`, allows the two distributions to differ in spread,
and stays fixed to equal variances unless you add a formula for it.

## The formula

In `bmm`, every parameter gets its own formula. This is the main difference
from `brms`, where you write one formula with the response on the left-hand
side, such as `n_old ~ condition`. `bmf()` is short for `bmmformula()`.
Concretely, we let sensitivity vary between participants, and we estimate one
criterion per base-rate condition, again with participant-level variation:

```{r}
formula <- bmf(
  d ~ 1 + (1 | id),
  criterion ~ 0 + condition + (1 | id)
)
```

`(1 | id)` lets a parameter vary between participants, as in the mixed-model
syntax of `lme4` and `brms`. `0 + condition` drops the intercept, so each
coefficient is the criterion of one condition rather than a difference from a
reference condition. The
[bmmformula article](https://popov-lab.github.io/bmm/articles/bmm_bmmformula.html)
explains the syntax in full, including how to fix a parameter to a constant.

## The fit

`bmm()` takes the formula, the data and the model. Its `file` argument caches
the fit on disk, creating the folder if it does not exist; this is how this
page avoids refitting the model every time it is built, and deleting the
file forces a refit. Everything else goes to `brms::brm()`, so you can set the
number of chains, cores and iterations as you would there. A chain is one run
of the sampler. Its first iterations, the warmup, tune the sampler and are
discarded; the remaining iterations are the posterior draws from which every
estimate below is computed. The defaults are 4 chains of 2000 iterations,
the first 1000 of each as warmup, which gives 4000 posterior draws:

```{r fit, results = "hide", message = FALSE}
fit <- bmm(
  formula = formula,
  data = broeder_schuetz_2009_e3,
  model = model,
  cores = 4,
  backend = "cmdstanr",
  file = "assets/bmmfit_sdt_yn_getting_started"
)
```

## Reading the results

The summary lists the estimates for each parameter, together with the
convergence diagnostics you know from `brms`:

```{r, comment = NA}
summary(fit)
```

Each row is named after a parameter and the predictor it belongs to. The
first block, headed "Multilevel Hyperparameters", gives the standard
deviations of the participant-level effects. The second block gives the
population-level estimates: `d_Intercept` is the sensitivity averaged over
participants, and the five `criterion` coefficients are the criteria of the
base-rate conditions. Both parameters use an identity link here, so the
estimates are on the natural scale of the model. `Estimate` is the posterior
mean, `Est.Error` the posterior standard deviation, and `l-95% CI` and
`u-95% CI` bound the 95% credible interval, the range that holds the
parameter with 95% probability given the data and the priors. The last block
lists `sdratio` at its fixed value of 0, which the `Formula` block shows as
`sdratio = 0`; the parameter is on a log scale, so 0 means a ratio of 1, that
is, equal variances. In the summary above, the criterion decreases from `br1`
to `br5`. That is, participants say "old" more readily the more old items
they expect, which is what the manipulation was meant to do, while
sensitivity is one number for all conditions because we modeled it that way.

Before you interpret an estimate, check that every `Rhat` is below 1.01 and
that `Bulk_ESS` and `Tail_ESS`, the effective sample sizes, are above 400. If
they are not, raise the number of iterations, for example with
`bmm(..., iter = 4000)`, or look at the
[Stan guidance on runtime warnings](https://mc-stan.org/misc/warnings.html).

A posterior predictive check compares the observed counts with counts
simulated from the fitted model. The dark line is the distribution of the
observed counts, each light line the distribution of one simulated data set.
If the model describes the data, the dark line sits inside the bundle of light
ones:

```{r pp-check, message = FALSE}
pp_check(fit, ndraws = 50)
```

# Where to go next

- If you know `brms`, [From brms to bmm](https://popov-lab.github.io/bmm/articles/bmm_brms_to_bmm.html)
  fits one model in both packages and explains each difference in the code.
- The [bmmformula article](https://popov-lab.github.io/bmm/articles/bmm_bmmformula.html)
  covers the formula syntax: fixed parameters, non-linear predictors, and how
  parameters on a link scale are read.
- The task sections above link to one article per model family, each with a
  worked example on package data.
- `?bmm` documents the fitting function, and
  [Extracting model information](https://popov-lab.github.io/bmm/articles/bmm_extract_info.html)
  shows how to get the default priors, the Stan code and the Stan data of a
  model before you fit it.
- If a model you need is missing, or something does not work as this page
  says, open an [issue on GitHub](https://github.com/popov-lab/bmm/issues).
