---
title: "Get started with bmm"
output: bookdown::html_document2
author:
  - Gidon Frischkorn
  - Ven Popov
pkgdown:
  as_is: true
---

```{=html}
<style type="text/css">
div.main-container {
max-width: 850px !important;
}

p {
margin-top: 1.5em ;
margin-bottom: 1.5em ;
}
.author{
    display: none;
}
</style>
```

```{r, include = FALSE}
options(crayon.enabled = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "jpeg",
  dpi = 100,
  fig.asp = 0.8,
  fig.align = "center"
)
fansi::set_knit_hooks(knitr::knit_hooks, which = c("output", "message", "error"))
options(width = 300)
```

# What bmm does

`bmm` fits cognitive measurement models to behavioral data. You write the
model for each parameter with the formula syntax of `brms`, and `bmm`
translates the measurement model into a distribution that `brms` and Stan can
sample from. In other words, you get hierarchical Bayesian estimates of
interpretable parameters, such as memory precision or the sensitivity in a
recognition task, with the same formula interface you already use for
regression.

This page takes you from an installed package to a first fitted model. It has
three parts: checking that your machine can fit models, finding the model that
matches your task, and fitting one model end to end.

# Install and check your machine

The released version is on CRAN, the development version on GitHub:

```r
install.packages("bmm")

# development version
if (!requireNamespace("remotes")) {
  install.packages("remotes")
}
remotes::install_github("popov-lab/bmm", upgrade = "never")
```

`upgrade = "never"` keeps `remotes` from updating the packages `bmm` depends
on. Updating `Rcpp`, `StanHeaders` or `RcppParallel` underneath an installed
`rstan` can leave `rstan` unable to load, in particular on Windows, and `brms`
needs `rstan` to store the results of every fit, also with the `cmdstanr`
backend.

Fitting a model needs more than the package: a C++ compiler, and a Stan
backend, either `cmdstanr` with CmdStan or `rstan`. These are the parts that
fail most often on a fresh machine, and the error usually shows up only when
the first model refuses to compile. Therefore, run `bmm_setup()` before
anything else:

```r
library(bmm)
bmm_setup()
```

The function checks each requirement, prints `PASS` or `FAIL` per check, and
gives one fix for every check that failed. It installs nothing itself. With the
default `smoke_test = TRUE`, it finally compiles and samples a small mixture
model, which is the only check that shows the whole chain works. This is what
the report looked like on the machine that built this page, without the
smoke test:

```{r setup-report}
library(bmm)
bmm_setup(smoke_test = FALSE)
```

If a check fails, apply the fix it prints, restart R, and run `bmm_setup()`
again. Once every check passes, you are ready to fit models.

# Which model fits your task

You come to `bmm` with data from a task, so this section is organised by task.
Each entry lists the models `bmm` provides for it and links to the article that
walks through them. The lists are generated from the installed package, so they
always show the models of the version you are running.

## Continuous reproduction

Participants reproduce a continuous feature, such as a color or an
orientation, and the response error on the circle is the dependent variable.
The [continuous reproduction article](https://popov-lab.github.io/bmm/articles/bmm_vwm_crt.html)
describes the task and compares the models.

```{r, results = "asis", echo = FALSE}
print_pretty_models_md(group = "Continuous reproduction")
```

## Categorical recall and n-AFC decisions

Participants pick one response from a set of categories, for example the
correct item, an item from another list position, or an item that was not
studied. The [M3 article](https://popov-lab.github.io/bmm/articles/bmm_m3.html)
shows how to set up the response categories and the activation model.

```{r, results = "asis", echo = FALSE}
print_pretty_models_md(group = "Categorical recall and n-AFC decisions")
```

## Detection, recognition and confidence judgements

Participants decide whether a signal was present or an item was studied, rate
their confidence, pick the target among several alternatives, or rank the
alternatives. Signal detection theory separates sensitivity from response
bias for each of these response formats, and `bmm` has one model per format.
The models expect aggregated counts per subject and condition, which the help
page of each model describes.

```{r, results = "asis", echo = FALSE}
print_pretty_models_md(group = "Detection, recognition and confidence judgements")
```

The models differ in the response format they take, in the noise distributions
they offer through the `dist` argument, and in whether the signal distribution
may have a different spread from the noise distribution. This table is built
from the installed package, so a model appears once it is part of the version
you are running.

```{r sdt-table, echo = FALSE}
sdt_models <- list(
  list(format = "yes/no detection, old/new recognition", model = "sdt_yn",
       uv = "`sdratio`"),
  list(format = "confidence ratings", model = "sdt_rating",
       uv = "`sdratio`"),
  list(format = "m-alternative forced choice", model = "sdt_mafc",
       uv = "none"),
  list(format = "ranking", model = "sdt_ranking",
       uv = "`sdratio` (normal noise only)"),
  list(format = "old/new confidence with remember/know judgements",
       model = "sdt_cdp", uv = "`sigmar` (recollection)")
)
available <- supported_models(print_call = FALSE)
rows <- lapply(sdt_models, function(m) {
  if (!m$model %in% available) return(NULL)
  f <- formals(get(m$model, asNamespace("bmm")))
  versions <- if (!is.null(f$version)) eval(f$version) else NULL
  model <- paste0("`", m$model, "()`")
  if (!is.null(versions)) {
    model <- paste0(model, ", versions ", paste0("`", versions, "`", collapse = ", "))
  }
  data.frame(
    `Response format` = m$format,
    Model = model,
    `dist` = paste(eval(f$dist), collapse = ", "),
    `Unequal variance` = m$uv,
    check.names = FALSE
  )
})
knitr::kable(do.call(rbind, rows))
```

## Choices and response times

Participants choose between two options and the response time is recorded on
every trial. The [DDM article](https://popov-lab.github.io/bmm/articles/bmm_ddm.html)
and the [censored shifted Wald article](https://popov-lab.github.io/bmm/articles/bmm_cswald.html)
cover trial-level data. The [EZ-diffusion article](https://popov-lab.github.io/bmm/articles/bmm_ezdm.html)
covers the case where you only have the mean and variance of the response
times and the accuracy per condition. Fast guesses and other contaminant
responses are the topic of the
[response time contamination article](https://popov-lab.github.io/bmm/articles/bmm_rt_contamination.html).

```{r, results = "asis", echo = FALSE}
print_pretty_models_md(group = "Choices and response times")
```

```{r other-groups, results = "asis", echo = FALSE}
known <- c(
  "Continuous reproduction", "Categorical recall and n-AFC decisions",
  "Detection, recognition and confidence judgements", "Choices and response times"
)
registry <- bmm:::model_registry()
other <- setdiff(unique(registry$group), known)
if (length(other) > 0) {
  cat("## Other models\n\n")
  cat("These models were added after this page was written:\n\n")
  print_pretty_models_md(group = other)
}
```

You can print the same list in R at any time with `supported_models()`, and
`?modelname` (for example `?sdt_yn`) documents the data a model expects, its
parameters and its default priors.

# Fit your first model

We fit a yes/no signal detection model to the recognition data of Broeder and
Schuetz (2009, Experiment 3), which ships with the package as
`broeder_schuetz_2009_e3` (see `?broeder_schuetz_2009_e3` for the source). 40
participants judged items as old or new under five base-rate conditions. A
higher proportion of old items makes participants more willing to say "old",
so the conditions shift the response criterion while sensitivity should stay
the same.

## The data

`sdt_yn()` expects one row per subject, condition and stimulus type, with the
number of "old" responses and the number of trials in that cell:

```{r}
head(broeder_schuetz_2009_e3)
```

- `stimulus` codes the stimulus type: `1` for old items, `0` for new items.
- `n_old` counts the "old" responses in the cell. For old items these are the
  hits, for new items the false alarms.
- `n_trials` is the number of items in the cell, which differs between cells
  because the base rate manipulates it.
- `condition` runs from the most conservative (`br1`) to the most liberal
  (`br5`) base rate.

If your data has one row per trial, you can pass it as is: the response column
then holds `0` or `1` and the trial column holds `1` on every row.

## The model

A model object names the columns the model needs. Nothing is estimated yet;
this only tells `bmm` where to look:

```{r}
model <- sdt_yn(
  response = "n_old",
  stimulus = "stimulus",
  n_trials = "n_trials"
)
```

The model has two parameters. `d` is the sensitivity: the distance between
the evidence distributions of old and new items. `criterion` is the response
bias: the point on the evidence axis above which participants say "old". A
third parameter, `sdratio`, allows the two distributions to differ in spread,
and stays fixed to equal variances unless you add a formula for it.

## The formula

In `bmm`, every parameter gets its own formula. This is the main difference
from `brms`, where the response is on the left-hand side. Concretely, we let
sensitivity vary between participants, and we estimate one criterion per
base-rate condition, again with participant-level variation:

```{r}
formula <- bmf(
  d ~ 1 + (1 | id),
  criterion ~ 0 + condition + (1 | id)
)
```

`0 + condition` drops the intercept, so each coefficient is the criterion of
one condition rather than a difference from a reference condition. The
[bmmformula article](https://popov-lab.github.io/bmm/articles/bmm_bmmformula.html)
explains the syntax in full, including how to fix a parameter to a constant.

## The fit

`bmm()` takes the formula, the data and the model. Everything else goes to
`brms::brm()`, so you can set the number of chains, cores and iterations as
you would there. `file` caches the fit, which is how this page avoids
refitting the model every time it is built:

```{r fit, results = "hide", message = FALSE}
fit <- bmm(
  formula = formula,
  data = broeder_schuetz_2009_e3,
  model = model,
  cores = 4,
  backend = "cmdstanr",
  file = "assets/bmmfit_sdt_yn_getting_started"
)
```

## Reading the results

The summary lists the estimates for each parameter, together with the
convergence diagnostics you know from `brms`:

```{r, comment = NA}
summary(fit)
```

The first block gives the standard deviations of the participant-level
effects. The second block gives the population-level estimates: `d_Intercept`
is the sensitivity averaged over participants, and the five `criterion`
coefficients are the criteria of the base-rate conditions. Both parameters
use an identity link here, so the estimates are on the natural scale of the
model. In the summary above, the criterion decreases from `br1` to `br5`.
That is, participants say "old" more readily the more old items they expect,
which is what the manipulation was meant to do, while sensitivity is one
number for all conditions because we modelled it that way.

Before you interpret an estimate, check that all `Rhat` values are close to 1
and that the effective sample sizes are in the hundreds or higher. If they are
not, run more iterations or look at the
[Stan guidance on runtime warnings](https://mc-stan.org/misc/warnings.html).

A posterior predictive check compares the observed counts with counts
simulated from the fitted model. If the model describes the data, the observed
distribution sits inside the simulated ones:

```{r pp-check, message = FALSE}
pp_check(fit, ndraws = 50)
```

# Where to go next

- The [bmmformula article](https://popov-lab.github.io/bmm/articles/bmm_bmmformula.html)
  covers the formula syntax: fixed parameters, non-linear predictors, and how
  parameters on a link scale are read.
- The task sections above link to one article per model family, each with a
  worked example on package data.
- `?bmm` documents the fitting function, and
  [Extracting model information](https://popov-lab.github.io/bmm/articles/bmm_extract_info.html)
  shows how to get the default priors, the Stan code and the Stan data of a
  model before you fit it.
- If a model you need is missing, or something does not work as this page
  says, open an [issue on GitHub](https://github.com/popov-lab/bmm/issues).
