---
title: "Visualizing conditional effects of model parameters"
output: bookdown::html_document2
author:
  - Gidon Frischkorn
  - Ven Popov
bibliography: REFERENCES.bib
header-includes:
  - \usepackage{amsmath}
vignette: >
  %\VignetteIndexEntry{Visualizing conditional effects of model parameters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
pkgdown:
  as_is: true
---
<style type="text/css">
div.main-container {
max-width: 800px !important;
}

p {
margin-top: 1.5em ;
margin-bottom: 1.5em ;
}
.author{
    display: none;
}
</style>

```{r, include = FALSE}
options(crayon.enabled = TRUE, width = 800)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "jpeg",
  dpi = 100
)
fansi::set_knit_hooks(knitr::knit_hooks, which = c("output","message","error"))
```

# Overview

After fitting a model with `bmm()`, you often want to visualize how predictors
affect model parameters. If you are familiar with `brms`, you may know the
`conditional_effects()` function for this purpose. However, calling
`brms::conditional_effects()` directly on a `bmmfit` object requires you to know
which brms arguments to use for each parameter, and returns results on the
sampling scale without link-function back-transformation.

The `conditional_effects()` method for `bmmfit` objects makes this easier:

- **Specify parameters by name**: Just pass `par = "kappa"` or `par = "c"` and
  the method takes care of the rest.
- **Choose your scale**: View results on the natural parameter scale
  (`scale = "native"`, the default) or on the MCMC sampling scale
  (`scale = "sampling"`).
- **Mixture proportions as probabilities**: For `mixture3p` models, the softmax
  transformation is applied jointly so that mixing proportions (`thetat`,
  `thetant`) are displayed as proper probabilities that sum to 1.

# Basic usage

We illustrate the function using the SDM model fitted to the
`oberauer_lin_2017` data set. First, we fit a model where the memory strength
parameter `c` varies by `set_size`, while the precision parameter `kappa` is
estimated as an intercept only:

```{r fit-sdm, eval = FALSE}
library(bmm)

fit_sdm <- bmm(
  formula = bmf(c ~ 0 + set_size, kappa ~ 1),
  data = oberauer_lin_2017,
  model = sdm(resp_error = "dev_rad")
)
```

To compute and plot the conditional effect of `set_size` on `c`:

```{r ce-basic, eval = FALSE}
ce_c <- conditional_effects(fit_sdm, par = "c")
plot(ce_c)
```

By default, the results are shown on the **native scale**. Since `c` uses a log
link, this means the exponentiated values (always positive). To see the results
on the **sampling scale** (log scale) instead:

```{r ce-sampling, eval = FALSE}
ce_c_log <- conditional_effects(fit_sdm, par = "c", scale = "sampling")
plot(ce_c_log)
```

# Viewing all estimated parameters at once

If you omit the `par` argument, effects are computed for all estimated
parameters. Fixed parameters (like `mu` in the SDM) are excluded automatically:

```{r ce-all, eval = FALSE}
ce_all <- conditional_effects(fit_sdm)
plot(ce_all)
```

The resulting effect names are prefixed with the parameter name (e.g.,
`"c.set_size"`, `"kappa.1"`) so you can distinguish them.

# Mixture proportions in mixture3p

For `mixture3p` models, the mixing proportions `thetat` (target) and `thetant`
(non-target) are estimated on an unconstrained scale internally. When you use
`scale = "native"` (the default), the softmax transformation is applied jointly,
and you see proper probabilities between 0 and 1:

```{r ce-softmax, eval = FALSE}
fit_mix3p <- bmm(
  formula = bmf(kappa ~ 0 + setsize, thetat ~ 0 + setsize),
  data = zhang_luck_2008,
  model = mixture3p(
    resp_error = "response_error",
    nt_features = paste0("col_lure", 1:5),
    set_size = "setsize"
  )
)

# P(target response) on probability scale
ce_thetat <- conditional_effects(fit_mix3p, par = "thetat")
plot(ce_thetat)

# Precision on native (exponentiated) scale
ce_kappa <- conditional_effects(fit_mix3p, par = "kappa")
plot(ce_kappa)
```

# Passing additional arguments

All extra arguments are forwarded to `brms::conditional_effects()`. Some useful
options:

```{r ce-advanced, eval = FALSE}
# Select a specific predictor to plot
ce_specific <- conditional_effects(fit_sdm, par = "c", effects = "set_size")

# Adjust credible interval width
ce_narrow <- conditional_effects(fit_sdm, par = "c", prob = 0.80)

# Add spaghetti lines showing individual posterior draws
ce_spag <- conditional_effects(fit_sdm, par = "c", spaghetti = TRUE, ndraws = 50)
plot(ce_spag)
```

See `?brms::conditional_effects` for the full list of available options.

# Quick reference

| What you want | How to do it |
|:--------------|:-------------|
| Single parameter | `conditional_effects(fit, par = "kappa")` |
| All parameters | `conditional_effects(fit)` |
| Native scale (default) | `scale = "native"` |
| Sampling scale | `scale = "sampling"` |
| Specific predictor | `effects = "set_size"` |
| Narrower CI | `prob = 0.80` |
| Spaghetti plot | `spaghetti = TRUE, ndraws = 50` |
