---
title: "Signal Detection Theory models"
output: bookdown::html_document2
author:
  - Gidon Frischkorn
  - Ven Popov
bibliography: REFERENCES.bib
header-includes:
  - \usepackage{amsmath}
vignette: >
  %\VignetteIndexEntry{Signal Detection Theory models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
pkgdown:
  as_is: true
---
<style type="text/css">
div.main-container {
max-width: 800px !important;
}

p {
margin-top: 1.5em ;
margin-bottom: 1.5em ;
}
.author{
    display: none;
}
</style>

```{r, include = FALSE}
options(crayon.enabled = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  dev = "jpeg",
  dpi = 100,
  fig.width = 5,
  fig.asp = 0.8,
  out.width = "80%",
  fig.align = "center"
)
fansi::set_knit_hooks(knitr::knit_hooks, which = c("output", "message", "error"))
```

```{r setup, message = FALSE}
library(bmm)
library(ggplot2)
```

# Signal detection theory in bmm

Signal detection theory (SDT) separates an observer's *sensitivity* — how well
two classes of stimuli (signal vs. noise, old vs. new) can be told apart — from
their *response bias* — where they place the decision criterion
[@Green_Swets_1966; @DeCarlo_1998]. `bmm` ships four SDT measurement models, one
per response format. Each is a self-contained model with its own data
requirements; they are not modes of a single function.

| Constructor | Response format | Key parameters | ROC defined? |
|-------------|-----------------|----------------|--------------|
| `sdt_binary()` | counts of "signal"/"old" responses | `dprime`, `criterion`, `sdratio` | yes |
| `sdt_rating()` | counts per confidence-rating category | `dprime`, `criterion`, threshold params, `sdratio` | yes |
| `sdt_mafc()` | counts of correct m-AFC responses | `dprime` | no |
| `sdt_ranking()` | counts per rank position | `dprime`, `sdratio` | no |

All four assume a latent evidence variable whose noise distribution is selected
with the `dist` argument: `"normal"` (Gaussian SDT), `"logistic"`,
`"gumbel_min"` (extreme-value SDT), or `"gumbel_max"`.

Two of the models — `sdt_binary()` and `sdt_rating()` — have an explicit
response criterion, so a **receiver operating characteristic (ROC)** curve and
its **area under the curve (AUC)** are defined. The m-AFC and ranking models
have no criterion: accuracy already integrates over the decision rule, so
`roc_sdt()` and `auc_sdt()` deliberately error for them. Posterior predictive
checks via `pp_check()` are available for all four.

This article walks through fitting each model and then through the
post-processing toolkit: `roc_sdt()`, `roc_observed()`, `auc_sdt()`,
`latent_sdt()`, the `plot()` methods (including the z-transformed ROC via
`scale = "z"`), and `pp_check()`.

# SDT in brief

All SDT models share one idea. On each trial the observer obtains a noisy sample
of *evidence* from a latent variable $X$. Noise items and signal items generate
evidence from two distributions that overlap; the observer cannot see which
distribution a sample came from, only its value. For the Gaussian (`"normal"`)
model the two distributions are

$$
X \mid \text{noise} \sim \mathrm{Normal}(0,\, 1),
\qquad
X \mid \text{signal} \sim \mathrm{Normal}(d',\, s^2),
$$

where **sensitivity** $d'$ is the mean separation expressed in units of the noise
standard deviation, and $s$ is the signal-to-noise SD ratio (the `sdratio`
parameter, fixed to $s = 1$ unless estimated). In a binary task the observer
responds "signal" whenever the evidence exceeds a **criterion** $k$. The two
response probabilities are then tail areas of the two distributions:

$$
\text{FA} = P(X > k \mid \text{noise}) = 1 - \Phi(k),
\qquad
\text{Hit} = P(X > k \mid \text{signal}) = 1 - \Phi\!\left(\tfrac{k - d'}{s}\right).
$$

```{r sdt-distributions, echo = FALSE, fig.cap = "Latent evidence model. Noise and signal distributions overlap; responses above the criterion are labelled as signal. The shaded right tails are the false-alarm rate (noise) and the hit rate (signal); their separation is the sensitivity d-prime."}
crit <- 0.75
dpr  <- 1.5
x    <- seq(-3.5, 5.5, length.out = 600)
dens <- rbind(
  data.frame(x = x, density = dnorm(x, 0, 1),   dist = "Noise"),
  data.frame(x = x, density = dnorm(x, dpr, 1), dist = "Signal")
)
ggplot(dens, aes(x, density, color = dist, fill = dist)) +
  geom_area(data = subset(dens, x > crit), alpha = 0.3,
            position = "identity", color = NA) +
  geom_line(linewidth = 0.8) +
  geom_vline(xintercept = crit, linetype = "dashed") +
  annotate("segment", x = 0, xend = dpr, y = 0.43, yend = 0.43,
           arrow = arrow(ends = "both", length = unit(0.12, "cm"))) +
  annotate("text", x = dpr / 2 - 0.2, y = 0.465, label = "italic(d)*minute",
           parse = TRUE) +
  annotate("text", x = crit + 0.15, y = 0.55, label = "criterion (k)",
           hjust = 0) +
  scale_color_brewer(palette = "Set1") +
  scale_fill_brewer(palette = "Set1") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  coord_cartesian(ylim = c(0, 0.56)) +
  labs(x = "Latent evidence", y = "Density", color = NULL, fill = NULL) +
  theme_classic() +
  theme(legend.position = "top")
```

Sweeping the criterion from strict to lenient traces every (false-alarm, hit)
pair the observer could produce — the **receiver operating characteristic
(ROC)**. When the two distributions have equal variance ($s = 1$) the ROC is
symmetric about the minor diagonal; when the signal distribution is wider
($s > 1$, as is typical for recognition memory) the ROC becomes asymmetric. This
is why a single (hit, false-alarm) point cannot identify $s$: separating $d'$
from $s$ requires *several* operating points along the curve.

```{r sdt-roc-geometry, echo = FALSE, fig.cap = "ROC geometry. Equal-variance models give a symmetric curve; unequal variance (wider signal distribution) bends it asymmetrically. The dashed line is chance (d' = 0)."}
far <- seq(1e-4, 1 - 1e-4, length.out = 400)
roc <- function(far, d, s) 1 - pnorm((qnorm(1 - far) - d) / s)
curves <- rbind(
  data.frame(FA = far, Hit = roc(far, 1.5, 1.0), Model = "Equal variance (s = 1)"),
  data.frame(FA = far, Hit = roc(far, 1.5, 1.6), Model = "Unequal variance (s = 1.6)")
)
ggplot(curves, aes(FA, Hit, color = Model)) +
  geom_abline(linetype = "dashed", color = "grey60") +
  geom_line(linewidth = 1) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  scale_color_brewer(palette = "Dark2") +
  labs(x = "False-alarm rate", y = "Hit rate", color = NULL) +
  theme_classic() +
  theme(legend.position = "top")
```

The four response formats below reuse this latent model and differ only in the
decision rule mapped onto it. The `dist` argument selects the shape of the latent
distributions: `"normal"` is the textbook Gaussian model; `"logistic"` is the
numerically convenient logistic analogue; `"gumbel_min"` is the extreme-value
(EVSDT) model that often fits recognition data well and yields a logistic ROC;
`"gumbel_max"` is its mirror image. Gaussian is the safe default unless theory or
prior fits motivate an extreme-value form.

# Binary SDT

In a binary old/new recognition task each cell of the design contributes a count
of "old" responses out of a number of trials, separately for old (`signal`) and
new (`noise`) items. The latent decision variable places "old" responses above a
criterion; sensitivity `dprime` is the distance between the signal and noise
distributions and `criterion` is the location of the boundary.

## A single condition: equal-variance SDT

We simulate one condition with `rsdt_binary()` and fit an equal-variance model.

```{r}
set.seed(123)
dat <- rsdt_binary(n_per_cell = 100, n_subjects = 30,
                   dprime = 1.5, criterion = 0.2)

model <- sdt_binary(response = "n_old", stimulus = "stimulus",
                    n_trials = "n_trials")

fit_ev <- bmm(
  formula = bmf(dprime ~ 1 + (1 | id), criterion ~ 1 + (1 | id)),
  data = dat,
  model = model,
  backend = "cmdstanr",
  cores = 4,
  refresh = 0,
  silent = 2,
  file = "assets/bmmfit_sdt_binary_ev_vignette"
)
summary(fit_ev)
```

The population estimates recover the simulated `dprime = 1.5` and
`criterion = 0.2`.

By default `sdratio` is fixed to 0, i.e. the signal and noise distributions have
equal variance (`exp(0) = 1`). With a single (hit, false-alarm) pair per subject
the signal/noise variance ratio is not identified, so leaving it fixed is the
right default.

## Multiple criteria: identifying the unequal-variance ratio

When the criterion is shifted across conditions — for example by a base-rate or
payoff manipulation — while sensitivity is held constant, the several
(false-alarm, hit) points trace out a *curve* rather than a single point. That
curvature is what identifies the unequal-variance ratio `sdratio` — the
signal-to-noise SD ratio $s$, whose reciprocal is the slope of the zROC
[@DeCarlo_1998]. The bundled `broeder_schuetz_2009_e3` data
[@Broeder_Schuetz_2009] has exactly this structure: 40 subjects tested under five
base-rate conditions (`br1` conservative … `br5` liberal).

```{r}
model <- sdt_binary(response = "n_old", stimulus = "stimulus",
                    n_trials = "n_trials")

fit_uv <- bmm(
  formula = bmf(
    dprime    ~ 1 + (1 | id),
    criterion ~ 0 + condition + (1 | id),
    sdratio   ~ 1
  ),
  data = broeder_schuetz_2009_e3,
  model = model,
  backend = "cmdstanr",
  cores = 4,
  refresh = 0,
  silent = 2,
  file = "assets/bmmfit_sdt_binary_uv_vignette"
)
summary(fit_uv)
```

Here `dprime` and `sdratio` are shared across conditions while `criterion` varies
by `condition` — the five criteria are five operating points on a single ROC.
`sdratio` is estimated on the log scale, so exponentiate it for the SD ratio on
the natural scale:

```{r}
exp(brms::fixef(fit_uv)["sdratio_Intercept", c("Estimate", "Q2.5", "Q97.5")])
```

A value above 1 means the signal distribution is wider than the noise
distribution, the classic recognition-memory pattern.

## ROC and AUC

`roc_sdt()` returns the model-implied ROC from the posterior. For a binary fit it
traces the smooth analytical curve from the posterior of `dprime` (and `sdratio`
when estimated). When it detects a predictor that varies the criterion only — as
`condition` does above — it additionally returns the model-implied operating
points (one per criterion level) in a `points` attribute. `roc_observed()` gives
the empirical operating points straight from the response counts, and the
`plot()` method overlays everything.

```{r}
roc <- roc_sdt(fit_uv)
obs <- roc_observed(fit_uv)

plot(roc, observed = obs)
```

The smooth ribbon is the posterior ROC, the filled points with crosshairs are the
model-implied operating points per base-rate condition, and the open circles are
the observed points. A curvilinear ROC that is asymmetric about the minor
diagonal is the signature of unequal variance (`sdratio > 1`).

`auc_sdt()` returns the posterior area under the curve. For *equal-variance*
Gaussian and Gumbel-min fits it uses the closed forms $\Phi(d'/\sqrt{2})$ and
$\mathrm{logistic}(g')$ respectively; every other case — unequal variance, the
logistic distribution, and all rating fits — is integrated numerically. The
returned AUC is always the area under the full curve (one value per curve, not
the area spanned by the discrete points).

```{r}
auc <- auc_sdt(fit_uv)
plot(auc)
attr(auc, "summary")
```

## z-ROC and latent distributions

Passing `scale = "z"` to the ROC `plot()` method reads the rates on the noise
distribution's quantile axis ($z = \Phi^{-1}(\text{rate})$ for the Gaussian
model). On this axis the model ROC is a straight line whose slope is
$1/\exp(\mathtt{sdratio})$ and whose intercept is $d'/\exp(\mathtt{sdratio})$, so
a slope below 1 is the visual signature of unequal variance and observed points
that bow away from a straight line flag misfit. (The $(0,0)$ and $(1,1)$
endpoints map to infinity and are dropped.)

```{r}
plot(roc, observed = obs, scale = "z")
```

`latent_sdt()` reconstructs the latent decision-variable distributions
themselves: the noise and signal evidence densities, separated by `dprime` and
scaled by `exp(sdratio)`, with the criterion drawn as a dashed line carrying a
credible band on its location. This is the picture the criterion and `dprime`
estimates describe.

```{r}
plot(latent_sdt(fit_uv))
```

## Posterior predictive checks

`sdt_binary()` uses a custom binomial family, so `pp_check()` delegates to
`brms`. The "old"-response counts span a wide range whose ceiling differs across
cells (the base-rate conditions have different numbers of trials), so the bar
view is cluttered and hard to read. An overlaid empirical CDF split by stimulus
type is a cleaner check that the model reproduces the response distributions for
signal and noise items — and the ROC plot above already checks the operating
points directly:

```{r}
pp_check(fit_uv, type = "ecdf_overlay", group = "stimulus", ndraws = 100)
```

# Rating SDT

In a confidence-rating task each item is judged on an ordered scale (e.g. 1 =
"sure new" … 6 = "sure old"). The data are counts per rating category, one column
per category. Internally `bmm` fits this with `brms`' native multinomial family;
the K-1 confidence criteria are parameterized by a `threshold_type`
[@Selker_2019; @Paulewicz_Blaut_2020].

We use a real recognition ROC dataset. The `roc6` object in the
[MPTinR](https://cran.r-project.org/package=MPTinR) package collects 12 published
old/new recognition experiments as per-subject counts of old and new items across
a 6-point confidence scale. Here we take @Pratte_2010 (97 subjects, 240 trials per
stimulus class — narrow posteriors, clear effects) and reshape it to the format
`sdt_rating()` expects: one row per (subject × stimulus), with categories
`r1` = "sure new" … `r6` = "sure old".

```{r}
stopifnot(requireNamespace("MPTinR", quietly = TRUE))
data("roc6", package = "MPTinR")
pratte <- roc6[roc6$exp == "Pratte_2010", ]

rating_data <- rbind(
  data.frame(id = pratte$id, stimulus = 1L,
             r1 = pratte$OLD_3new, r2 = pratte$OLD_2new, r3 = pratte$OLD_1new,
             r4 = pratte$OLD_1old, r5 = pratte$OLD_2old, r6 = pratte$OLD_3old),
  data.frame(id = pratte$id, stimulus = 0L,
             r1 = pratte$NEW_3new, r2 = pratte$NEW_2new, r3 = pratte$NEW_1new,
             r4 = pratte$NEW_1old, r5 = pratte$NEW_2old, r6 = pratte$NEW_3old)
)

model <- sdt_rating(response = paste0("r", 1:6), stimulus = "stimulus")

# Recognition ROCs are classically asymmetric, so estimate unequal variance
fit_rating <- bmm(
  formula = bmf(dprime ~ 1 + (1 | id), criterion ~ 1 + (1 | id),
                spacing ~ 1, sdratio ~ 1),
  data = rating_data,
  model = model,
  backend = "cmdstanr",
  cores = 4,
  refresh = 0,
  silent = 2,
  chains = 4,
  iter = 2000,
  file = "assets/bmmfit_sdt_rating_vignette"
)
summary(fit_rating)
```

## Threshold parameterizations

A rating model is binary SDT with several criteria at once: for a $K$-point scale
the $K-1$ thresholds carve the latent evidence axis into $K$ ordered response
bins. The probability of each rating is the area of the corresponding bin under
the signal or noise distribution.

```{r sdt-rating-thresholds, echo = FALSE, fig.cap = "A 6-point rating scale places K-1 = 5 thresholds on the evidence axis, splitting it into six ordered confidence bins. threshold_type governs how these thresholds are spaced."}
thr   <- c(-1.5, -0.6, 0.1, 0.9, 1.9)
dpr   <- 1.4
edges <- c(-3.2, thr, 3.2)
mids  <- (head(edges, -1) + tail(edges, -1)) / 2
x     <- seq(-3.5, 5.5, length.out = 600)
dens <- rbind(
  data.frame(x = x, density = dnorm(x, 0, 1),   dist = "Noise"),
  data.frame(x = x, density = dnorm(x, dpr, 1), dist = "Signal")
)
ggplot(dens, aes(x, density, color = dist)) +
  geom_line(linewidth = 0.8) +
  geom_vline(xintercept = thr, linetype = "dashed", color = "grey50") +
  annotate("text", x = mids, y = 0.45, label = paste0("r", 1:6), size = 3) +
  scale_color_brewer(palette = "Set1") +
  labs(x = "Latent evidence", y = "Density", color = NULL) +
  theme_classic() +
  theme(legend.position = "top")
```

The `threshold_type` argument controls how those $K-1$ criteria are placed and
therefore how many parameters they cost:

- `"parsimonious"` (default) and `"equidistant"`: two parameters
  (`criterion` + `spacing`) — the thresholds are evenly spaced around the
  criterion, differing only in their canonical anchor positions [@Selker_2019].
- `"log_distance"` and `"log_ratio"`: K-2 `delta` parameters that allow uneven
  spacing while guaranteeing the thresholds stay ordered [@Paulewicz_Blaut_2020].
- `"softmax"`: a shared `spacing` plus K-3 allocation `delta` parameters.

Add `sdratio ~ 1` to the formula to estimate unequal variance.

## ROC, AUC, and observed ROC

For a rating model the K-1 thresholds define K-1 ROC points per posterior draw.
`roc_sdt()` returns those points with their posterior uncertainty; `plot()` shows
the mean operating points with crosshair credible intervals and a connecting
curve; `roc_observed()` gives the empirical points from the pooled counts.

```{r}
roc <- roc_sdt(fit_rating)
plot(roc, observed = roc_observed(fit_rating))

attr(auc_sdt(fit_rating), "summary")
```

## Posterior predictive checks

Because the rating model uses a multinomial family, `pp_check()` produces a
response-proportion profile (observed bars vs. posterior-predictive
point-ranges). For SDT the signal and noise rating distributions should be
checked separately, so `pp_check()` **defaults to faceting by `stimulus`**:

```{r}
pp_check(fit_rating, ndraws = 100)
```

Pass `group = NA` to pool both stimulus types into a single profile, or
`group = "<predictor>"` to facet by another design variable.

The predicted point-ranges track the observed bars closely at the extreme
categories but **deviate systematically for the interior categories** (r3–r5):
the misses are smooth and one-sided rather than random scatter. That pattern is
the fingerprint of *threshold misfit*. The default `"parsimonious"`
parameterization buys its economy (two threshold parameters) by fixing the K-1
criteria at regularly spaced canonical positions, and when the true confidence
thresholds are unevenly spaced — as they usually are in recognition data — those
two parameters cannot reproduce the full rating profile.

## Improving fit with flexible thresholds

The `"log_ratio"` parameterization frees the spacing with K-2 `delta` parameters
(for a 6-point scale: `delta1`, `delta2`, `delta4`, `delta5`; `delta3` is the
reference midpoint), so the thresholds can sit wherever the data require. Refit
with it and re-run the same check:

```{r}
model_lr <- sdt_rating(response = paste0("r", 1:6), stimulus = "stimulus",
                       threshold_type = "log_ratio")

fit_rating_lr <- bmm(
  formula = bmf(dprime ~ 1 + (1 | id), criterion ~ 1 + (1 | id),
                delta1 ~ 1, delta2 ~ 1, delta4 ~ 1, delta5 ~ 1,
                sdratio ~ 1),
  data = rating_data,
  model = model_lr,
  backend = "cmdstanr",
  cores = 4,
  chains = 4,
  iter = 2000,
  refresh = 0,
  silent = 2,
  file = "assets/bmmfit_sdt_rating_logratio_vignette"
)

pp_check(fit_rating_lr, ndraws = 100)
```

The interior categories are now captured: the flexible thresholds absorb the
systematic deviation that the parsimonious model could not. The improvement is
not free — it costs four threshold parameters (the `delta`s) instead of two — so
it should be weighed against the gain in fit.

For these models the posterior predictive check is the most practical guide to
that trade-off: a parameterization that leaves systematic, one-sided deviations
is too rigid, while one whose predicted point-ranges sit on the observed bars (as
here) is adequate. A *formal* predictive comparison is tempting, but the usual
`loo()` is **not** appropriate here: each observation is an aggregated count over
many trials, so leaving one out removes a large, highly influential chunk of data
and the PSIS-LOO approximation becomes unreliable — it flags most observations
with a high Pareto-$k$, and (because that reflects heavy-tailed importance weights
rather than Monte Carlo noise) more posterior draws will not fix it. The
principled alternative for aggregated data is k-fold cross-validation via
`brms::kfold()`, which refits on held-out folds and avoids importance sampling
altogether. `"log_distance"` behaves like `"log_ratio"`, and `"softmax"` offers an
intermediate option with a shared spacing plus fewer free deltas.

# m-AFC SDT

In an m-alternative forced-choice task the observer picks the target among `m`
alternatives; the data are counts of correct responses. There is no criterion and
no stimulus column — accuracy alone identifies `dprime` [@DeCarlo_2012]. A correct
response occurs when the target's evidence exceeds that of all $m-1$ distractors,
so `dprime` is referenced to the maximum of the $m-1$ noise samples and stays on
the same scale as binary $d'$: for $m = 2$ this is the closed form
$\Phi(d'/\sqrt{2})$, and for $m \geq 3$ `bmm` evaluates the accuracy by
quadrature. The set size `m` may be a constant or a data column (for mixed set
sizes).

```{r}
# rsdt_mafc() is vectorised over its arguments and returns a data frame with
# id, n_trials and n_correct; varying dprime by subject gives hierarchical data.
set.seed(123)
dat <- rsdt_mafc(dprime = rnorm(30, mean = 1.5, sd = 0.3), m = 4, n_trials = 100)

model <- sdt_mafc(response = "n_correct", n_trials = "n_trials", m = 4)

fit_mafc <- bmm(
  formula = bmf(dprime ~ 1 + (1 | id)),
  data = dat,
  model = model,
  backend = "cmdstanr",
  cores = 4,
  refresh = 0,
  silent = 2,
  file = "assets/bmmfit_sdt_mafc_vignette"
)
summary(fit_mafc)
```

ROC curves are undefined here (`roc_sdt(fit_mafc)` errors). Each subject
contributes one correct-response count, so a bar per count value is sparse and
noisy; an overlaid empirical CDF of the correct-response counts is a compact
check that the model captures the spread of accuracy across subjects:

```{r}
pp_check(fit_mafc, type = "ecdf_overlay", ndraws = 100)
```

# Ranking SDT

In a ranking task the observer orders `m` items by evidence and the data are
counts per rank position assigned to the target [@MeyerGrant_2025]. As with
m-AFC there is no criterion: the rank assigned to the target is an *order
statistic* of the latent evidence, so the probability of each rank follows from
the evidence distribution directly. With `dist = "gumbel_min"` these rank
probabilities have a closed form (the extreme-value assumption that makes ranking
analytically tractable); `dist = "normal"` is also available and evaluated by
quadrature — prefer `gumbel_min` for speed and the standard Thurstonian ranking
model, `normal` when you want Gaussian evidence comparable to the other SDT
models. The model uses the native multinomial family with one category per rank
position, and `m` may again be a constant or a column for mixed set sizes. The
bundled `meyer_grant_jakob_2025` data [@MeyerGrant_Jakob_2025] illustrate the
wide format.

```{r}
model <- sdt_ranking(
  response = c("rank1", "rank2", "rank3", "rank4", "rank5"),
  m = "set_size",
  dist = "gumbel_min"
)

fit_rank <- bmm(
  formula = bmf(dprime ~ 1 + (1 | id)),
  data = meyer_grant_jakob_2025,
  model = model,
  backend = "cmdstanr",
  cores = 4,
  refresh = 0,
  silent = 2,
  file = "assets/bmmfit_sdt_ranking_vignette"
)
summary(fit_rank)
```

ROC/AUC are undefined; `pp_check()` shows the rank-position profile. With mixed
set sizes, facet by the set-size column so each panel has a homogeneous
denominator:

```{r}
pp_check(fit_rank, group = "set_size", ndraws = 100)
```

# Post-processing reference

| Function | Models | Returns |
|----------|--------|---------|
| `roc_sdt()` | binary, rating | model-implied ROC draws (+ operating points for multi-criteria binary) |
| `roc_observed()` | binary, rating | empirical ROC points from the data |
| `auc_sdt()` | binary, rating | posterior AUC |
| `latent_sdt()` | binary, rating | noise/signal latent densities + criterion/threshold locations |
| `plot()` | `bmm_sdt_roc` (incl. `scale = "z"`), `bmm_sdt_auc`, `bmm_sdt_latent` | `ggplot2` ROC / z-ROC / AUC / latent-distribution figures |
| `pp_check()` | all four | posterior predictive check |

# References
