---
title: "Interpreting Model Results"
output: bookdown::html_document2
author:
  - Gidon Frischkorn
  - Ven Popov
bibliography: REFERENCES.bib
header-includes:
  - \usepackage{amsmath}
vignette: >
  %\VignetteIndexEntry{Interpreting Model Results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
pkgdown:
  as_is: true
---
<style type="text/css">
div.main-container {
max-width: 800px !important;
}

p {
margin-top: 1.5em ;
margin-bottom: 1.5em ;
}
.author{
    display: none;
}
</style>

```{r, include = FALSE}
options(crayon.enabled = TRUE, width = 800)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "jpeg",
  dpi = 100
)
fansi::set_knit_hooks(knitr::knit_hooks, which = c("output","message","error"))
```

# Overview

You have fitted a measurement model with `bmm()`. Now what? This article walks
through the steps that usually follow: checking that the model actually fits the
data, looking at what the parameter estimates tell you, and formally testing
whether your experimental manipulation had an effect.

We cover five topics:

1. Posterior predictive checks with `pp_check()`.
2. Visualizing predictor effects with `conditional_effects()`.
3. Testing specific hypotheses with `brms::hypothesis()` and
   `emmeans::emmeans()`.
4. Comparing different predictor structures for the same model via LOO
   cross-validation.
5. Comparing different measurement models on the same data.

Since `bmm` builds on `brms`, most post-estimation techniques you already know
from `brms` work on `bmmfit` objects without modification.


# Setup {#setup}

We use the `oberauer_lin_2017` data set, which contains continuous reproduction
responses from 19 participants at set sizes 1 through 8 [@Oberauer_Lin_2017].
We fit three models to this data:

- A **full SDM** [@Oberauer_2023] where memory strength `c` varies by set size.
- A **restricted SDM** where `c` is a single intercept (no set-size effect).
- A **two-component mixture model** [@Zhang_Luck_2008] for cross-model
  comparison.

```{r load-packages, message = FALSE}
library(bmm)
```

## Fitting the models

The full SDM lets `c` vary across set sizes, while `kappa` (precision) is
estimated as a single intercept. We pass `sample_prior = TRUE` so that
Savage-Dickey Bayes factors are available for `brms::hypothesis()` later, and
`save_pars = brms::save_pars(all = TRUE)` so that bridge sampling works for
`brms::bayes_factor()`.

```{r fit-sdm-full}
fit_sdm <- bmm(
  formula = bmf(c ~ 0 + set_size, kappa ~ 1),
  data = oberauer_lin_2017,
  model = sdm(resp_error = "dev_rad"),
  cores = 4,
  backend = "cmdstanr",
  sample_prior = TRUE,
  save_pars = brms::save_pars(all = TRUE),
  file = "assets/bmmfit_postestimation_sdm"
)
```

The restricted SDM estimates `c` as a single intercept, ignoring set size:

```{r fit-sdm-null}
fit_sdm_null <- bmm(
  formula = bmf(c ~ 1, kappa ~ 1),
  data = oberauer_lin_2017,
  model = sdm(resp_error = "dev_rad"),
  cores = 4,
  backend = "cmdstanr",
  sample_prior = TRUE,
  save_pars = brms::save_pars(all = TRUE),
  file = "assets/bmmfit_postestimation_sdm_null"
)
```

The mixture model decomposes responses into a von Mises target component and a
uniform guessing component:

```{r fit-mix2p}
fit_mix2p <- bmm(
  formula = bmf(thetat ~ 0 + set_size, kappa ~ 1),
  data = oberauer_lin_2017,
  model = mixture2p(resp_error = "dev_rad"),
  cores = 4,
  backend = "cmdstanr",
  save_pars = brms::save_pars(all = TRUE),
  file = "assets/bmmfit_postestimation_mix2p"
)
```


# Posterior predictive checks {#pp-check}

Before diving into parameter estimates, you should check that the model actually
describes the data reasonably well. Posterior predictive checks do this by
simulating data from the fitted model and comparing it to what was observed.

## Basic density overlay

The default `pp_check()` overlays the observed data density (dark line) on
densities from posterior predictive draws (light lines). If the two look
similar, the model is doing a reasonable job:

```{r pp-check-sdm, fig.width = 7, fig.height = 4}
pp_check(fit_sdm, ndraws = 100)
```

## Comparing models visually

Running `pp_check()` on both the full and restricted SDM gives you a quick sense
of whether dropping the set-size predictor matters. If the restricted model
produces visibly worse predictions (e.g., a wider or less peaked density), that
is already informative:

```{r pp-check-null, fig.width = 7, fig.height = 4}
pp_check(fit_sdm_null, ndraws = 100)
```

## Grouped checks

You can also split the check by an experimental variable to see whether the
model captures condition-level differences:

```{r pp-check-grouped, fig.width = 7, fig.height = 4}
pp_check(fit_sdm, ndraws = 100, group = "set_size",
         type = "dens_overlay")
```

If certain conditions (say, the highest set sizes) are poorly captured, that
points to misspecification for those conditions specifically.


# Visualizing conditional effects {#conditional-effects}

With the model checked, the next thing you probably want is a plot of how your
predictors affect each parameter.

## Why not just use `brms::conditional_effects()`?

You can, but it is a bit awkward. Calling `brms::conditional_effects()` on a
`bmmfit` requires you to figure out whether each parameter is a `dpar` or
`nlpar` in brms terms, and the results come back on the sampling scale without
back-transformation.

The `bmm` method wraps this so you do not have to worry about those details:

- Pass `par = "kappa"` or `par = "c"` and the method figures out the rest.
- Results are shown on the native parameter scale by default (e.g., exponentiated
  for log-linked parameters). Use `scale = "sampling"` if you want the raw
  sampling scale.
- For `mixture3p` models, the softmax transformation is applied jointly so that
  `thetat` and `thetant` are displayed as probabilities that sum to 1.

## Single parameter

To plot the conditional effect of `set_size` on memory strength `c`:

```{r ce-basic, fig.width = 7, fig.height = 4}
ce_c <- conditional_effects(fit_sdm, par = "c")
plot(ce_c)
```

Since `c` uses a log link, the native scale shows exponentiated values (always
positive). If you want the log scale instead:

```{r ce-sampling, fig.width = 7, fig.height = 4}
ce_c_log <- conditional_effects(fit_sdm, par = "c", scale = "sampling")
plot(ce_c_log)
```

## All parameters at once

Omitting `par` computes effects for every estimated parameter. Fixed parameters
(like `mu` in the SDM) are excluded automatically:

```{r ce-all, fig.width = 7, fig.height = 4}
ce_all <- conditional_effects(fit_sdm)
plot(ce_all)
```

Effect names are prefixed with the parameter (`"c.set_size"`, `"kappa.1"`) so
you can tell them apart.

## Credible intervals and spaghetti lines

Extra arguments are forwarded to `brms::conditional_effects()`. For instance, to
narrow the credible interval and overlay individual posterior draws:

```{r ce-advanced, fig.width = 7, fig.height = 4}
ce_spag <- conditional_effects(fit_sdm, par = "c",
                               prob = 0.80,
                               spaghetti = TRUE,
                               ndraws = 50)
plot(ce_spag)
```

See `?brms::conditional_effects` for all available options.

## Mixture models

For `mixture2p`, `thetat` is the mixture weight for target responses. On the
native scale the logit link is inverted, so you see probabilities directly:

```{r ce-mix2p, fig.width = 7, fig.height = 4}
ce_thetat <- conditional_effects(fit_mix2p, par = "thetat")
plot(ce_thetat)
```

For `mixture3p`, both `thetat` and `thetant` go through a joint softmax, so the
displayed probabilities sum to 1.


# Testing hypotheses on model parameters {#hypothesis-testing}

## `brms::hypothesis()` {#brms-hypothesis}

`brms::hypothesis()` lets you test directed hypotheses about model parameters.
You write hypotheses as character strings using the parameter names from the
model summary.

To see what those names are:

```{r summary}
summary(fit_sdm)
```

The fixed effects for `c` are called `c_set_size1` through `c_set_size8`.
Because `c` uses a log link, these are on the log scale. To test whether memory
strength at set size 1 exceeds that at set size 4:

```{r hypothesis-directed}
brms::hypothesis(fit_sdm, "c_set_size1 > c_set_size4")
```

The output includes the posterior probability for the hypothesis, an evidence
ratio (Bayes factor), and a 95% credible interval for the difference. Because we
passed `sample_prior = TRUE` when fitting, the evidence ratio is a Savage-Dickey
density ratio [@Wagenmakers_2010].

You can test several hypotheses at once. Here we check whether `c` decreases
monotonically across selected set sizes:

```{r hypothesis-multiple}
hypotheses <- c(
  "c_set_size1 > c_set_size2",
  "c_set_size2 > c_set_size4",
  "c_set_size4 > c_set_size8"
)
brms::hypothesis(fit_sdm, hypotheses)
```

### Native scale

Keep in mind that `brms::hypothesis()` operates on the sampling scale. For `c`
that means log scale. If you want to compare on the original scale, wrap the
expressions in the inverse link:

```{r hypothesis-native}
brms::hypothesis(fit_sdm, "exp(c_set_size1) - exp(c_set_size4) > 0")
```


## `emmeans` for estimated marginal means {#emmeans}

The `emmeans` package [@Lenth_2024] is a standard R tool for post-hoc
comparisons. `bmm` provides S3 methods so that `emmeans()` works directly on
`bmmfit` objects.

The main advantage over `brms::hypothesis()` is that you refer to parameters by
their model name (`"c"`, `"kappa"`) rather than having to know whether brms
treats them as `dpar` or `nlpar` internally.

### Marginal means for `c`

```{r emmeans-c, message = FALSE}
library(emmeans)

em_c <- emmeans(fit_sdm, ~ set_size, dpar = "c")
em_c
```

These are on the sampling scale (log scale for `c`).

### Pairwise comparisons

```{r emmeans-pairs}
pairs(em_c)
```

Each row is the posterior difference between two set-size levels. Where the 95%
CI excludes zero, there is evidence for a difference.

### Back-transformation

To get estimates on the natural scale, use `regrid()`:

```{r emmeans-regrid}
em_c_native <- regrid(em_c, transform = "log")
em_c_native
```

Pairwise comparisons after `regrid()` are on the natural scale:

```{r emmeans-pairs-native}
pairs(em_c_native)
```

### Custom contrasts

`emmeans` also supports custom contrasts. For instance, a polynomial contrast to
test for a linear trend:

```{r emmeans-contrasts}
contrast(em_c, method = "poly")
```


# Comparing predictor structures {#model-comparison}

Does adding the set-size predictor actually help? There are two ways to answer
this: LOO cross-validation, which asks which model predicts better out of
sample, and Bayes factors, which quantify relative evidence given the data and
priors.

## LOO-CV comparison

`brms::loo()` computes the expected log pointwise predictive density (ELPD).
`loo_compare()` ranks the models:

```{r loo-compare}
loo_sdm <- brms::loo(fit_sdm)
loo_sdm_null <- brms::loo(fit_sdm_null)
loo_compare(loo_sdm, loo_sdm_null)
```

The model with the higher ELPD (closer to zero) fits better out of sample. The
`elpd_diff` column is the difference relative to the best model, with a standard
error. A difference larger than about 4 standard errors is strong evidence.

Also check the Pareto $k$ diagnostics: observations with $k > 0.7$ have
unreliable LOO estimates and may need closer inspection.

```{r loo-diagnostics, fig.width = 7, fig.height = 4}
plot(loo_sdm, label_points = TRUE)
```

## Bayes factor comparison

`brms::bayes_factor()` computes the Bayes factor between two models via bridge
sampling. This requires that all Stan parameters are saved, which is why we
passed `save_pars = brms::save_pars(all = TRUE)` when fitting:

```{r bf-predictor}
brms::bayes_factor(fit_sdm, fit_sdm_null)
```

The Bayes factor tells you how much the data favor one model over the other,
given their respective priors. A $BF > 10$ is usually considered strong evidence
for the numerator model; $BF < 1/10$ favors the denominator.

Note that Bayes factors and LOO-CV answer slightly different questions. LOO-CV
is about predictive performance: which model will generalize better to new data?
Bayes factors weigh relative evidence for the two models given the priors. The
two will usually agree, but they can diverge when priors are very diffuse or
when the models differ mainly in how they handle extreme observations.


# Comparing different measurement models {#cross-model}

Sometimes the question is not which predictors to include but which measurement
model to use. Here we compare the SDM against the two-component mixture model,
both fit to the same data.

## LOO-CV across models

Both models are fit to the same response variable (`dev_rad`) and the same
observations, so their ELPD estimates are directly comparable:

```{r loo-cross-model}
loo_mix2p <- brms::loo(fit_mix2p)
loo_compare(loo_sdm, loo_mix2p)
```

## Bayes factor across models

The same `brms::bayes_factor()` call works for comparing different measurement
models:

```{r bf-cross-model}
brms::bayes_factor(fit_sdm, fit_mix2p)
```

Since both models were fit with `save_pars = brms::save_pars(all = TRUE)`,
bridge sampling can estimate the marginal likelihood for each. The resulting
Bayes factor tells you how much the data favor the SDM over the mixture model
(or vice versa), given their respective priors.

## Things to watch out for

A few caveats when comparing across models:

- Both models must use the same response variable. You cannot compare a
  continuous-response model against a categorical one this way.
- LOO-CV compares on the log-likelihood scale. Different families or link
  functions are fine as long as the response is the same.
- LOO-CV penalizes overfitting, so a more complex model is not automatically
  favored.
- Bayes factors depend on the priors. With very diffuse priors, the more complex
  model is penalized more heavily (the Lindley paradox). Make sure the priors on
  both models are reasonable before interpreting the Bayes factor.

@Oberauer_2023 discusses model comparison for visual working memory models in
more detail.


# Summary {#summary}

| Goal | Method | Function |
|:-----|:-------|:---------|
| Check model adequacy | Posterior predictive check | `pp_check()` |
| Visualize predictor effects | Conditional effects | `conditional_effects()` |
| Test a specific hypothesis | Savage-Dickey density ratio | `brms::hypothesis()` |
| Compare conditions (all pairwise) | Estimated marginal means | `emmeans::emmeans()` + `pairs()` |
| Compare predictor structures | LOO cross-validation | `brms::loo()` + `loo_compare()` |
| Compare predictor structures | Bayes factor | `brms::bayes_factor()` |
| Compare measurement models | LOO cross-validation | `brms::loo()` + `loo_compare()` |
| Compare measurement models | Bayes factor | `brms::bayes_factor()` |

In practice, these steps tend to go in order:

1. Run `pp_check()` to make sure the model fits.
2. Plot `conditional_effects()` to see what the estimates look like.
3. Test hypotheses with `brms::hypothesis()` or `emmeans()`.
4. If you are comparing models, use `brms::loo()`, `loo_compare()`, or
   `brms::bayes_factor()`.


# References
