---
title: "From brms to bmm"
output: bookdown::html_document2
author:
  - Gidon Frischkorn
  - Ven Popov
bibliography: REFERENCES.bib
pkgdown:
  as_is: true
---

```{=html}
<style type="text/css">
div.main-container {
max-width: 850px !important;
}

p {
margin-top: 1.5em ;
margin-bottom: 1.5em ;
}
.author{
    display: none;
}
</style>
```

```{r, include = FALSE}
options(crayon.enabled = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "jpeg",
  dpi = 100,
  fig.asp = 0.8,
  fig.align = "center"
)
fansi::set_knit_hooks(knitr::knit_hooks, which = c("output", "message", "error"))
options(width = 300)
```

# Who this article is for

You know `brms`. You have written `brm(y ~ x + (1 | id))`, read its summary, and
perhaps written a non-linear formula or two. Then you open a `bmm` article and
the formula has no response on its left-hand side, the model is an object of its
own, and the estimates come with links you did not choose. This article
translates between the two packages.

We fit one model twice: once in plain `brms`, once with `bmm`. The model is the
two-parameter mixture model for continuous reproduction tasks
[@Zhang_Luck_2008], which `brms` can express with its built-in mixture family.
Both fits run the same Stan model, so they give the same estimates. With that
established, the rest of the article explains each difference in the code: where
the response went, why every parameter gets its own formula, what
`0 + setsize` and `1 + setsize` each give you, which scale the estimates are on,
and why `bmm` sets informative default priors.

# The same model, written twice

## The data

`zhang_luck_2008` ships with `bmm`. Eight participants saw one, two, three or
six colored squares, and after a delay they reproduced the color of one of them
on a color wheel. The response error is the distance between the reported and
the true color, in radians:

```{r, message = FALSE}
library(bmm)
library(brms)

dat <- zhang_luck_2008
head(dat[, c("subID", "setsize", "response_error")])
```

The mixture model assumes that a response comes either from memory or from a
guess. Memory responses scatter around the true color with a von Mises
distribution, the normal distribution of the circle, whose precision is `kappa`.
Guesses fall anywhere on the circle with equal probability. The second parameter
is the probability that a response comes from memory. We let both parameters
differ between set sizes and between participants.

## The model in brms

In `brms`, this is a mixture of two von Mises distributions. Each component has
a location (`mu1`, `mu2`) and a precision (`kappa1`, `kappa2`), and `theta1` is
the weight of the first component. Turning this into the mixture model takes
three constraints. Both locations are fixed at 0, because the response error is
already relative to the target. The precision of the second component is fixed
at a value so small that the distribution is flat, which makes it the guessing
distribution. Constant priors do the fixing:

```{r brms-fit, results = "hide", message = FALSE}
brms_fit <- brm(
  formula = bf(
    response_error ~ 1,
    kappa1 ~ 0 + setsize + (1 | subID),
    theta1 ~ 0 + setsize + (1 | subID),
    mu2 ~ 1,
    kappa2 ~ 1
  ),
  family = mixture(von_mises, von_mises, order = "none"),
  prior = prior(constant(0), class = Intercept, dpar = mu1) +
    prior(constant(0), class = Intercept, dpar = mu2) +
    prior(constant(-100), class = Intercept, dpar = kappa2) +
    prior(normal(2, 1), class = b, dpar = kappa1) +
    prior(logistic(0, 1), class = b, dpar = theta1) +
    prior(exponential(1), class = sd, dpar = kappa1) +
    prior(exponential(1), class = sd, dpar = theta1),
  data = dat,
  cores = 4,
  control = list(adapt_delta = 0.95),
  backend = "cmdstanr",
  file = "assets/brmsfit_brms_to_bmm"
)
```

`kappa2` is fixed at -100 because `brms` estimates `kappa` on the log scale, and
`exp(-100)` is as close to zero as anyone needs. With only eight participants,
the participant effects on the mixture weight make the sampler take small steps
in places, so we raise `adapt_delta` from its default of 0.8 to avoid divergent
transitions; this is a `brms` argument and works the same in `bmm()`. The last
four priors are not
the `brms` defaults. They are the priors `bmm` uses for this model, and we pass
them here so that both fits are the same model. We come back to them in
[the section on priors](#why-the-default-priors-are-informative).

## The model in bmm

The `bmm` version needs a model object, one formula per parameter, and the data:

```{r bmm-fit, results = "hide", message = FALSE}
bmm_fit <- bmm(
  formula = bmf(
    kappa ~ 0 + setsize + (1 | subID),
    thetat ~ 0 + setsize + (1 | subID)
  ),
  data = dat,
  model = mixture2p(resp_error = "response_error"),
  cores = 4,
  control = list(adapt_delta = 0.95),
  backend = "cmdstanr",
  file = "assets/bmmfit_brms_to_bmm"
)
```

The constraints are gone from the code. `mixture2p()` knows that the locations
are 0, that the second component is the guessing distribution, and which priors
to use. What is left is what you want to know: how `kappa` and `thetat`, the
weight of the memory responses (the `t` stands for target), depend on set size.

## Same model, same estimates

`bmm` translates its formula into a `brmsformula`, and `brms` writes the Stan
model from that. You can look at the formula `bmm` generated:

```{r}
bmm_fit$formula
```

`kappa` and `thetat` are non-linear parameters that `bmm` passes on to `kappa1`
and `theta1`. `mu1`, `mu2` and `kappa2` are fixed with constant priors, as in
the `brms` call. Apart from the parameter names, the Stan code of the two fits
is the same program with the same priors, so the posteriors agree up to the
noise of the sampler:

```{r, message = FALSE}
brms_est <- fixef(brms_fit)
bmm_est <- fixef(bmm_fit)
ci <- function(est, row) {
  sprintf(
    "%.2f [%.2f, %.2f]",
    est[row, "Estimate"], est[row, "Q2.5"], est[row, "Q97.5"]
  )
}
set_sizes <- levels(dat$setsize)
comparison <- data.frame(
  parameter = rep(c("kappa", "thetat"), each = length(set_sizes)),
  set_size = rep(set_sizes, 2),
  brms = c(
    sapply(paste0("kappa1_setsize", set_sizes), ci, est = brms_est),
    sapply(paste0("theta1_setsize", set_sizes), ci, est = brms_est)
  ),
  bmm = c(
    sapply(paste0("kappa_setsize", set_sizes), ci, est = bmm_est),
    sapply(paste0("thetat_setsize", set_sizes), ci, est = bmm_est)
  )
)
knitr::kable(comparison, row.names = FALSE)
```

`bmm` does not add anything to the model. It writes the `brms` code, checks the
data, and picks priors that suit the model, and that is all.

# Where the response lives

In `brms`, the response is on the left-hand side of the main formula:
`response_error ~ 1`. In `bmm`, it is an argument of the model object:
`mixture2p(resp_error = "response_error")`. The formulas only ever have
parameters on their left-hand side.

The reason is that a measurement model usually needs more than one column, and
each column plays a different role. The three-parameter mixture model needs the
response error, the colors of the items that were not cued, and the set size.
A yes/no signal detection model needs the number of "old" responses, the number
of trials, and whether the item was old or new. In `brms`, these roles end up
spread over the response, addition terms such as `| trials()`, and predictors
inside non-linear formulas. In `bmm`, the model object names each column once,
by its role:

```r
mixture3p(
  resp_error = "response_error", nt_features = "col_lure",
  set_size = "setsize", regex = TRUE
)
sdt_yn(response = "n_old", stimulus = "stimulus", n_trials = "n_trials")
```

Knowing the role of each column lets `bmm` check the data before anything is
compiled. For instance, `mixture2p()` warns if the response error looks like
degrees rather than radians. It also lets `bmm` build the columns that the Stan
model needs from the ones you have, such as which of the non-target columns are
filled at each set size. The help page of each model, such as `?mixture2p`,
lists the columns it expects.

# One formula per parameter

Because the response lives in the model object, a `bmmformula` is a list of
regressions, one for each parameter of the model. Each one is a regular `brms`
formula, with population-level effects, group-level effects in the `lme4`
syntax, and non-linear terms if you need them. The
[bmmformula article](https://popov-lab.github.io/bmm/articles/bmm_bmmformula.html)
covers the syntax in full, including how to fix a parameter to a constant.

## `0 + setsize` or `1 + setsize`

Most `bmm` articles write `0 + setsize`, and you may be used to `1 + setsize`
(or just `setsize`, which is the same thing). Both describe the same regression
of each parameter on set size. They differ in what each coefficient means, and,
as we show below, in which default prior each coefficient gets.

With `0 + setsize`, the formula has no intercept, and there is one coefficient
per set size. Each coefficient is the parameter in that cell of the design:
`kappa_setsize1` is `kappa` at set size 1, `kappa_setsize6` is `kappa` at set
size 6, both on the log scale explained
[below](#which-scale-the-estimates-are-on). This is often easier when you start
out, because the summary shows the estimate for each condition directly, and
those are the numbers you want to plot or report. If you are not sure which
coding to use, start with this one.

With `1 + setsize`, the intercept is the parameter in the reference condition,
set size 1, and the other coefficients are differences from it.
`kappa_setsize6` is now how much log `kappa` at set size 6 differs from log
`kappa` at set size 1.
This coding is just as suitable, and it reads naturally when the difference
between conditions is your question. To get the parameter in a cell, you add
the coefficients:

```{r bmm-fit-ref, results = "hide", message = FALSE}
bmm_fit_ref <- bmm(
  formula = bmf(
    kappa ~ 1 + setsize + (1 | subID),
    thetat ~ 1 + setsize + (1 | subID)
  ),
  data = dat,
  model = mixture2p(resp_error = "response_error"),
  cores = 4,
  control = list(adapt_delta = 0.95),
  backend = "cmdstanr",
  file = "assets/bmmfit_brms_to_bmm_reference"
)
```

```{r}
fixef(bmm_fit_ref)[c("kappa_Intercept", "kappa_setsize6"), ]

draws_ref <- as_draws_df(bmm_fit_ref)
draws_cell <- as_draws_df(bmm_fit)
kappa6_ref <- draws_ref$b_kappa_Intercept + draws_ref$b_kappa_setsize6
kappa6_cell <- draws_cell$b_kappa_setsize6
rbind(
  `1 + setsize` = quantile(kappa6_ref, c(0.025, 0.5, 0.975)),
  `0 + setsize` = quantile(kappa6_cell, c(0.025, 0.5, 0.975))
)
```

Add the coefficients draw by draw, as above, and summarize afterwards. The draws
carry the uncertainty of each coefficient and how the coefficients covary, which
adding the two point estimates loses. The reverse also works: a difference
between set sizes is the difference of two cell coefficients of the
`0 + setsize` fit, again taken per draw.

Doing this by hand for every cell and parameter gets tedious.
`native_parameters()` does it for you. It evaluates every parameter in every
cell of the design, whatever the coding of the formula, and reports it on the
scale of the model (more on scales in the next section). Here are both fits
side by side:

```{r}
np_cell <- native_parameters(bmm_fit, re_formula = NA, summary = TRUE)
np_ref <- native_parameters(bmm_fit_ref, re_formula = NA, summary = TRUE)
fmt <- function(np) sprintf("%.2f [%.2f, %.2f]", np$Estimate, np$Q2.5, np$Q97.5)
np_cell <- np_cell[np_cell$parameter %in% c("kappa", "thetat"), ]
np_ref <- np_ref[np_ref$parameter %in% c("kappa", "thetat"), ]
knitr::kable(data.frame(
  parameter = np_cell$parameter,
  set_size = np_cell$setsize,
  `0 + setsize` = fmt(np_cell),
  `1 + setsize` = fmt(np_ref[match(paste(np_cell$parameter, np_cell$setsize),
                                   paste(np_ref$parameter, np_ref$setsize)), ]),
  check.names = FALSE
), row.names = FALSE)
```

Unlike the coefficients in the first table, these values are on the scale of
the model: `kappa` is the precision itself, not its log, and `thetat` is shown
as the probability of a memory response, not its logit. `re_formula = NA` sets
the participant effects to zero, which gives the parameters of a typical
participant; `?native_parameters` explains how this differs from the population
average under a non-identity link.

```{r, include = FALSE}
np_ref_matched <- np_ref[match(paste(np_cell$parameter, np_cell$setsize),
                               paste(np_ref$parameter, np_ref$setsize)), ]
rel_diff <- abs(np_ref_matched$Estimate / np_cell$Estimate - 1)
worst <- which.max(rel_diff)
dpr <- mixture2p(resp_error = "response_error")$default_priors
prior_mean <- function(p) as.numeric(sub("^\\w+\\(([-0-9.]+),.*$", "\\1", p))
prior_sd <- function(p) as.numeric(sub("^.*,\\s*([0-9.]+)\\)$", "\\1", p))
kappa_sds <- c(prior_sd(dpr$kappa$main), prior_sd(dpr$kappa$effects))
kappa_cell_prior <- sprintf(
  "normal(%g, %.2f)", prior_mean(dpr$kappa$main), sqrt(sum(kappa_sds^2))
)
```

The two codings give the same results. In every cell, the estimates of the two
fits differ by less than `r ceiling(max(rel_diff) * 1000) / 10`% of their value;
the largest gap is `r np_cell$parameter[worst]` at set size
`r np_cell$setsize[worst]`, `r sprintf("%.2f", np_cell$Estimate[worst])` against
`r sprintf("%.2f", np_ref_matched$Estimate[worst])`. Pick the coding whose
coefficients you want to read.

What the coding does change is the default prior of each coefficient. `bmm`
sets its default priors per coefficient, and the value of a parameter in one
cell is a different quantity from the difference between two cells: the first
lies somewhere in the plausible range of the parameter, the second is centered
on zero. With `0 + setsize`, every coefficient is a cell, so every one gets the
prior for the parameter itself, `r dpr$kappa$main` for `kappa` on the log scale
and `r dpr$thetat$main` for `thetat`. With `1 + setsize`, only the intercept,
the reference cell, gets that prior. The differences get the prior for effects,
`r dpr$kappa$effects` for `kappa` and `r dpr$thetat$effects` for `thetat`. Any
other cell is the intercept plus a difference, so its prior is the sum of two
priors. For `kappa`, the two normal priors add up to `r kappa_cell_prior`, whose
standard deviation is the square root of
`r kappa_sds[1]`² + `r kappa_sds[2]`²: wider than the `r dpr$kappa$main` that
the same cell gets with `0 + setsize`. With this data set, the table above shows
that the difference barely matters. With fewer trials per cell, the priors carry
more weight, and the two codings can give different estimates. So check
`default_prior()` for the coding you use, as shown in [the section on
priors](#why-the-default-priors-are-informative), and set your own priors if the
defaults do not match what you know.

The participant effects can be coded either way, too. `(1 | subID)`, as here,
gives every participant one shift for all set sizes. `(0 + setsize | subID)` and
`(1 + setsize | subID)` both let participants differ per set size, coded as cell
values or as differences from the reference cell.

# Which scale the estimates are on

The summary of a `bmm` fit lists the link of each parameter:

```{r, comment = NA}
summary(bmm_fit)
```

`kappa` must be positive, and `thetat` turns into a probability. Neither can be
predicted by a linear model directly, so `bmm` estimates them on an unbounded
scale and transforms them back. `kappa` has a log link, so the coefficients are
log precisions. `thetat` has a logit link: the weight of the guessing component
is fixed at 0 on the same scale, so a `thetat` of 0 means that memory and
guessing are equally likely, and the probability that a response comes from
memory is `plogis(thetat)`. Every coefficient in the summary, every prior, and every
value you fix a parameter to lives on this link scale.

To report a parameter on its own scale, transform the draws and summarize
afterwards, not the other way around. The mean of `exp()` of the draws is not
`exp()` of their mean. For set size 6:

```{r}
c(
  `exp(mean)` = exp(mean(kappa6_cell)),
  `mean(exp)` = mean(exp(kappa6_cell))
)
```

`native_parameters()`, used above, follows this order for every parameter. For
the mixture models, `k2sd()` converts `kappa` into the circular standard
deviation in radians, which is the measure of precision many papers report:

```{r}
kappa_draws <- native_parameters(bmm_fit, re_formula = NA, pars = "kappa")
sd_deg <- k2sd(kappa_draws$value) * 180 / pi
tapply(sd_deg, kappa_draws$setsize, median)
```

A coefficient on its own transforms to something else. With `1 + setsize`,
`exp(b_kappa_setsize6)` is the factor by which `kappa` at set size 6 is larger
or smaller than at set size 1, not `kappa` at set size 6. The help page of each
model lists the links of its parameters.

# Why the default priors are informative

`brms` uses flat priors on regression coefficients unless you set them. For the
`brms` version of this model, the defaults are:

```{r}
default_prior(
  bf(
    response_error ~ 1,
    kappa1 ~ 0 + setsize + (1 | subID),
    theta1 ~ 0 + setsize + (1 | subID),
    mu2 ~ 1,
    kappa2 ~ 1
  ),
  data = dat,
  family = mixture(von_mises, von_mises, order = "none")
) |>
  as.data.frame() |>
  transform(prior = ifelse(prior == "", "(flat)", prior)) |>
  subset(dpar %in% c("kappa1", "theta1") & coef == "" & group == "",
         select = c(dpar, class, prior)) |>
  knitr::kable(row.names = FALSE)
```

`class = b` covers every coefficient of `kappa1` and `theta1`, so each of them
has a flat prior. That suits a regression whose coefficients could take any
value. It does not suit a measurement model. Its parameters stand for cognitive processes, and we know
roughly which values are plausible before seeing any data. `bmm` puts this
knowledge into its default priors:

```{r}
default_prior(
  bmf(kappa ~ 0 + setsize + (1 | subID), thetat ~ 0 + setsize + (1 | subID)),
  data = dat,
  model = mixture2p(resp_error = "response_error")
) |>
  as.data.frame() |>
  subset(nlpar %in% c("kappa", "thetat") & coef == "" & group == "",
         select = c(nlpar, class, prior)) |>
  knitr::kable(row.names = FALSE)
```

The `normal(2, 1)` on the log of `kappa` places 95% of its mass between a
`kappa` of `r round(exp(2 - 1.96), 1)` and `r round(exp(2 + 1.96), 1)`, that
is, between a circular standard deviation of
`r round(k2sd(exp(2 - 1.96)) * 180 / pi)` and
`r round(k2sd(exp(2 + 1.96)) * 180 / pi)` degrees. Values in the thousands,
which would mean memory accurate to a fraction of a degree, get next to no
prior mass. The `logistic(0, 1)` on `thetat` is flat on the probability
scale: before seeing the data, every probability of a memory response between
0 and 1 is equally likely.

There is a second reason, which has to do with the mixture. When `kappa` comes
close to 0, the memory distribution becomes as flat as the guessing
distribution, and the data cannot tell the two apart. In that region, `thetat`
can take any value without changing the fit. A flat prior does nothing to keep
the sampler out of that region; a prior that holds `kappa` in a plausible range
does.

What this means for you:

- The more trials per participant and condition, the less the estimates
  depend on the priors. With few trials, the priors pull the estimates towards
  their center.
- Look at the priors before you fit, with `default_prior()` as above, and at
  the help page of the model, which lists them.
- If you know more than the defaults, or something different, set your own
  priors with `prior = ` in `bmm()`. They are `brms` priors on the `bmm`
  parameter names, with `nlpar` or `dpar` as `default_prior()` shows it, for
  example `prior(normal(1.5, 0.5), class = b, nlpar = kappa)`. The
  [article on extracting model information](https://popov-lab.github.io/bmm/articles/bmm_extract_info.html)
  shows the details.
- Bayes factors depend on the priors much more than estimates do. If you plan to
  compute them, choose the priors on purpose rather than taking the defaults.

# What stays brms

A `bmm` fit is a `brmsfit` with extra information, so the `brms` toolbox works
on it: `fixef()`, `ranef()`, `as_draws_df()`, `pp_check()`,
`conditional_effects()`, `hypothesis()`, `loo()` and the rest. As shown above,
`bmm_fit$formula` holds the `brmsformula` that `bmm` generated, and
`stancode(bmm_fit)` the Stan code. `summary(bmm_fit, backend = "brms")` prints
the `brms` summary instead of the `bmm` one. If you want to change a model
beyond what `bmm` offers, these give you the `brms` code to start from.

# Where to go next

- [Get started](https://popov-lab.github.io/bmm/articles/bmm.html) lists the
  models by task and walks through a first fit.
- The [bmmformula article](https://popov-lab.github.io/bmm/articles/bmm_bmmformula.html)
  covers the formula syntax.
- The [mixture models article](https://popov-lab.github.io/bmm/articles/bmm_mixture_models.html)
  fits the two- and three-parameter mixture models to a larger data set.
- [Extracting model information](https://popov-lab.github.io/bmm/articles/bmm_extract_info.html)
  shows how to get the default priors, the Stan code and the Stan data before
  you fit.

# References
