---
title: "Random-Utility Choice Models (utility)"
output: bookdown::html_document2
author:
  - Gidon Frischkorn
bibliography: REFERENCES.bib
header-includes:
  - \usepackage{amsmath}
vignette: >
  %\VignetteIndexEntry{Random-Utility Choice Models (utility)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
pkgdown:
  as_is: true
---

```{=html}
<style type="text/css">
div.main-container {
max-width: 800px !important;
}
p {
margin-top: 1.5em ;
margin-bottom: 1.5em ;
}
.author{
    display: none;
}
</style>
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE}
library(bmm)
```

## Scope

The `utility()` model fits **category-count choice** data: on each trial a person
chooses one of a fixed set of categories, and you record how often each category
was chosen. Each category carries a *systematic utility* built from either known
payoffs or observed values, and the category is selected through a random-utility
(softmax) or Luce-ratio (simple) rule.

This model does **not** cover per-option attribute or binary-lottery choice
(prospect theory in the sense of @nilsson_hierarchical_2011), where every option
has its own outcome-and-probability attributes. That design needs a different data
interface and is out of scope here.

## Two ways to build utility

A category's utility is a weighted sum of components. `utility()` gives you two
entry points, depending on whether the weights or the values are known:

* **`payoffs`** — a matrix of *known* coefficients (from the experimental design)
  multiplying *estimated* weights. This is the welfare-weight case: the payoff
  structure is fixed by the game, and you estimate how much each payoff component
  matters.
* **`value_cols`** — *observed* per-trial value columns multiplied by an estimated
  sensitivity `gamma`. This is the expected-utility case, optionally with a power
  value function (`rho`) and Prelec probability weighting (`alpha`).

The two are mutually exclusive.

## Welfare weights from a payoff matrix

Consider a nested social dilemma [@gross_nested_2025] with three options — keep the
resource, share it with the in-group, or share it universally. Each option has a
known payoff vector for self, in-group, and out-group. We want the **welfare
weights** `wi` (in-group) and `wo` (out-group), relative to the self-payoff, which
we fix as the numeraire.

```{r welfare-model}
model_welfare <- utility(
  resp_cats = c("nkeep", "ningroup", "nuniversal"),
  payoffs = rbind(
    nkeep      = c(b = 1.0, wi = 0.0, wo = 0.0),
    ningroup   = c(b = 0.5, wi = 1.0, wo = 0.0),
    nuniversal = c(b = 0.3, wi = 0.6, wo = 0.9)
  ),
  choice_rule = "simple",
  numeraire   = c(b = 1)
)
```

The rows are the response categories, the columns are the weight parameters, and
`b` is the numeraire — the self-payoff weight, fixed to 1 so the other weights are
interpretable on that scale. You never write the activation formulas by hand; the
model generates one per category from the payoff matrix:

```{r welfare-actfuns}
construct_utility_act_funs(model_welfare)
```

You only supply formulas for the parameters you want to predict. To fit the model,
predict each welfare weight (here by a fixed intercept):

```{r welfare-fit, eval = FALSE}
fit_welfare <- bmm(
  formula = bmf(wi ~ 1, wo ~ 1),
  data    = my_data,
  model   = model_welfare,
  # the simple (Luce) rule takes log(utility); keep utilities positive
  prior   = brms::set_prior("normal(1, 0.5)", nlpar = "wi", lb = 0) +
            brms::set_prior("normal(1, 0.5)", nlpar = "wo", lb = 0)
)
```

In a recovery simulation with true `wi = 1.2` and `wo = 0.8` (30 subjects, 100
trials), the model recovers both weights with 95% intervals covering the truth and
no divergent transitions.

### Positive activations for the simple rule

The `simple` (Luce) rule normalises raw utilities, `P(i) = U_i / \sum_j U_j`, which
`bmm` computes on the log scale. If an identity-linked weight turns a utility
negative during sampling, `log()` returns `NaN`. `utility()` warns you about this
and the fix is a lower bound on the weight priors (`lb = 0`), as above. Welfare
weights are naturally non-negative, so this bound is also substantively sensible.

## Payoffs that vary by trial

Payoffs are not always fixed across the whole experiment. If the payoff structure
changes trial-by-trial (say the in-group return depends on a manipulated pie
size), pass a **data.frame** whose cells are either constants or the *names of data
columns*:

```{r pertrial-model}
model_trial <- utility(
  resp_cats = c("nkeep", "ningroup", "nuniversal"),
  payoffs = data.frame(
    row.names = c("nkeep", "ningroup", "nuniversal"),
    b  = c(1.0, 0.5, 0.3),
    wi = c(0.0, "wi_ingroup", 0.6),   # column name -> per-trial payoff
    wo = c(0.0, 0.0, 0.9)
  ),
  choice_rule = "simple",
  numeraire   = c(b = 1)
)
construct_utility_act_funs(model_trial)
```

Here `wi_ingroup` is a column in the data that supplies the in-group payoff for the
`ningroup` category on each trial, while the other coefficients stay constant. The
estimated weights `wi` and `wo` are unchanged — only the *known* coefficients now
vary per trial.

## Expected utility with value columns

When options carry observed values rather than known payoffs, use `value_cols` to
map each category to its per-trial value column and estimate a single value
sensitivity `gamma`:

```{r eu-model}
model_eu <- utility(
  resp_cats  = c("corr", "other", "npl"),
  num_options = c(1, 1, 8),
  value_cols = c(corr = "V_corr", other = "V_other"),
  utility_fn = "power"      # gamma * V^rho
)
construct_utility_act_funs(model_eu)
```

`utility_fn = "power"` adds a curvature parameter `rho`, and
`weighting = "prelec"` adds a probability-weighting parameter `alpha` that
reweights the choice-set proportions `n_i / N`. Both `rho` and `alpha` use a log
link, so their priors are placed on the log scale — a `normal(0, 0.5)` prior on the
linear predictor, which is a median of 1 (linear utility / no weighting) on the
natural scale. You read them back as `exp(rho)` and `exp(alpha)`.

## Identifiability guards

Utility parameters are only identifiable under specific design conditions. Instead
of returning a flat posterior, `utility()` checks the design at fit time and raises
an informative message:

| Parameter | Requirement | If violated |
|-----------|-------------|-------------|
| `gamma` (value slope) | the value column must vary (≥ 2 distinct values) | error |
| `rho` (power utility) | the value range should span a ratio ≥ 3 | warning |
| `alpha` (Prelec) | the choice-set proportions `n_i / N` must vary across trials | error |
| identity-linked weights + `simple` rule | positive activations (`lb = 0` priors) | warning |

The constructor also rejects incoherent specifications early — for example, value
curvature or probability weighting on the `payoffs` path (where coefficients are
already known), or Prelec weighting with the `simple` rule.

## Theoretical note

The softmax rule is McFadden's multinomial-logit random-utility model
[@mcfadden_conditional_1974]: `P(i) \propto n_i \exp(a_i)` is the choice
probability of `U_i = a_i + \ln n_i + \varepsilon_i` with i.i.d. Gumbel noise, and
a hierarchical fit is a mixed logit [@train_discrete_2009]. The simple rule is
Luce's ratio rule [@luce_individual_1959] over raw utilities. These are genuine
random-utility choice models, but they are *not* a claim of von
Neumann–Morgenstern subjective utility: softmax is interval-scale, simple is
ratio-scale, and `gamma` measures how strongly value modulates activation rather
than a normative utility. The `utility()` model is built on the same
multinomial-logit likelihood as the [m3](bmm_m3.html) model.

## References
