---
title: "The Censored Shifted Wald Model (cswald)"
output: bookdown::html_document2
author:
  - Gidon Frischkorn
  - Ven Popov
bibliography: REFERENCES.bib
header-includes:
  - \usepackage{amsmath}
vignette: >
  %\VignetteIndexEntry{The Censored Shifted Wald Model (cswald)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
pkgdown:
  as_is: true
---

<style type="text/css">
div.main-container {
max-width: 800px !important;
}

p {
margin-top: 1.5em ;
margin-bottom: 1.5em ;
}
.author{
    display: none;
}
</style>

```{r, include = FALSE}
options(crayon.enabled = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "jpeg",
  dpi = 100
)
fansi::set_knit_hooks(knitr::knit_hooks, which = c("output","message","error"))
```

# Introduction

The Censored Shifted Wald (cswald) model is a measurement model for choice
response time (RT) tasks, particularly suited for tasks with high accuracy
(few errors). The model was introduced by @millerRelationCensoredShifted2018
as a simplified alternative to the full Wiener diffusion model
[@ratcliffDiffusionDecisionModel2008] that is easier to estimate while still
capturing essential features of the decision-making process.

The cswald model is ideal when:

- You have RT data from a two-choice decision task
- Accuracy is relatively high (error rates < 20%)
- You want to separate different components of the decision process
- You need a computationally efficient alternative to the full diffusion model

# Theoretical Background

## The Shifted Wald Distribution

The Wald distribution (also known as the inverse Gaussian distribution)
describes the first-passage time of a Wiener process (Brownian motion with
drift) to a single absorbing boundary. In the context of decision-making,
this models the time it takes for evidence to accumulate to a decision
threshold.

The shifted Wald distribution adds a non-decision time component to account
for processes like stimulus encoding and motor response execution:

$$
RT = T_{decision} + T_{ndt}
$$

where $T_{decision}$ follows a Wald distribution and $T_{ndt}$ is the
non-decision time.

The probability density function of the shifted Wald is:

$$
f(t | v, a, \tau) = \frac{a}{\sqrt{2\pi (t-\tau)^3}}
\exp\left(-\frac{(a - v(t-\tau))^2}{2(t-\tau)}\right)
$$

for $t > \tau$, where:

- $v$ is the drift rate (rate of evidence accumulation)
- $a$ is the boundary (decision threshold)
- $\tau$ is the non-decision time

## Censoring for Error Responses

In tasks with high accuracy, most responses are correct. The "censored"
aspect of the model refers to how error responses are handled. Instead of
modeling errors as arising from a separate accumulation process toward the
wrong boundary (as in the full diffusion model), errors are treated as
**censored** observations — trials where the correct response had not yet
been reached.

This censoring approach assumes that:

1. Correct responses follow the shifted Wald distribution
2. Error responses represent trials where the accumulation process was
   interrupted before reaching the correct boundary

The likelihood for a correct response is the Wald density:

$$
L_{correct}(t) = f_{Wald}(t - \tau | v, a)
$$

The likelihood for an error response is the survivor function (1 - CDF):

$$
L_{error}(t) = S_{Wald}(t - \tau | v, a) = 1 - F_{Wald}(t - \tau | v, a)
$$

## Relationship to the Diffusion Model

The cswald model can be seen as a simplification of the Wiener diffusion
model that makes certain assumptions:

1. **Single boundary**: Only the correct response boundary is explicitly
   modeled
2. **Unbiased starting point**: The starting point is assumed to be
   equidistant from both boundaries
3. **High accuracy**: The model works best when error rates are low

For tasks with substantial error rates (> 20%), the competing risks version
of the model (`version = "crisk"`) provides a more appropriate fit by
modeling both boundaries explicitly.

# Model Versions

The `bmm` package implements two versions of the cswald model:

## Simple Version (`version = "simple"`)

The simple version treats errors as censored correct responses. It estimates:

- **drift**: The drift rate (rate of evidence accumulation toward the
  correct response)
- **bound**: The distance from the starting point to the correct boundary
- **ndt**: Non-decision time
- **s**: Diffusion constant (usually fixed to 1 for scaling)

This version is best suited for tasks with error rates below 20%.

**Important note on boundary parameterization:** The `bound` parameter in the
simple version represents only the distance from the (unbiased) starting point
to the correct response boundary. This is **half** the total boundary
separation in the standard diffusion model parameterization. To convert the
simple version's `bound` estimate to the equivalent DDM or crisk boundary
separation, multiply by 2:

$$
a_{DDM} = 2 \times \text{bound}_{simple}
$$

## Competing Risks Version (`version = "crisk"`)

The competing risks version models both correct and error responses as
arising from racing accumulators toward opposite boundaries. It estimates:

- **drift**: The drift rate (can be positive or negative; negative drift
  indicates evidence accumulation toward the lower boundary)
- **bound**: Total boundary separation (full distance between both boundaries,
  consistent with DDM parameterization)
- **ndt**: Non-decision time
- **zr**: Relative starting point (proportion of boundary separation,
  0.5 = unbiased)
- **s**: Diffusion constant

This version is better suited for tasks with substantial error rates. The
`bound` parameter is directly comparable to the boundary separation in the
standard diffusion model.

## When to Use Each Version

### Use `version = "simple"` when:

- Error rates are low (< 20%)
- You want a computationally efficient model
- The focus is on correct response times
- Errors can reasonably be treated as censored observations

### Use `version = "crisk"` when:

- Error rates are substantial (> 20%)
- You want to model both response types explicitly
- Starting point bias might be present
- You need estimates comparable to full diffusion model parameters

## Comparison with the Wiener Diffusion Model

The `cswald` model in `bmm` is related to but distinct from the full Wiener
diffusion model available in `brms::wiener()`. Key differences:

| Feature | cswald (simple) | cswald (crisk) | brms::wiener |
|---------|-----------------|----------------|--------------|
| Boundaries | Single | Both | Both |
| Starting point | Fixed (unbiased) | Estimated | Estimated |
| Bound param | Half separation | Full separation | Full separation |
| Drift sign | Positive only | Positive/negative | Positive/negative |
| Best for | High accuracy | Moderate accuracy | Any accuracy |
| Computation | Fast | Moderate | Slower |
| Parameters | 3-4 | 4-5 | 4-5 |

For tasks with low error rates, the `cswald` model provides equivalent
inferences with faster computation. When comparing results across model
versions, remember to multiply the simple version's `bound` by 2 to get the
equivalent total boundary separation.

# Fitting the Model with `bmm`

## Setup

```{r setup, message=FALSE, warning=FALSE}
library(bmm)
library(ggplot2)
```

## Generating Simulated Data

We'll simulate data from a diffusion process using `rcswald()`, which
internally uses `rtdists::rdiffusion()` to generate theoretically consistent
samples:

```{r}
# set seed for reproducibility
set.seed(42)

# simulate data for two conditions with different drift rates
n_per_condition <- 500

# Condition A: higher drift (faster, more accurate)
dat_A <- rcswald(n = n_per_condition, drift = 3, bound = 1.5, ndt = 0.3)
dat_A$condition <- "A"

# Condition B: lower drift (slower, less accurate)
dat_B <- rcswald(n = n_per_condition, drift = 1.5, bound = 1.5, ndt = 0.3)
dat_B$condition <- "B"

# combine data
dat <- rbind(dat_A, dat_B)
dat$condition <- factor(dat$condition)

# check error rates
aggregate(response ~ condition, data = dat, FUN = function(x) mean(x == 0))
```

Let's visualize the RT distributions:

```{r fig.width=8, fig.height=4}
ggplot(dat, aes(x = rt, fill = factor(response))) +
  geom_histogram(binwidth = 0.05, position = "identity", alpha = 0.7) +
  facet_wrap(~condition, labeller = labeller(
    condition = c(A = "Condition A (drift = 3)", B = "Condition B (drift = 1.5)")
  )) +
  scale_fill_manual(values = c("0" = "red", "1" = "blue"),
                    labels = c("0" = "Error", "1" = "Correct"),
                    name = "Response") +
  labs(x = "Response Time (s)", y = "Count",
       title = "Simulated RT distributions") +
  theme_classic()
```

## Specifying the Model

First, specify the model using `cswald()`:

```{r}
model <- cswald(
  rt = "rt",
  response = "response",
  version = "simple"
)
```

## Specifying the Formula

We want to estimate how drift rate varies by condition while keeping the
boundary and non-decision time constant:

```{r}
formula <- bmf(
  drift ~ 0 + condition,
  bound ~ 1,
  ndt ~ 1
)
```

## Fitting the Model

```{r, message=FALSE, results='hide'}
fit <- bmm(
  formula = formula,
  data = dat,
  model = model,
  cores = 4,
  backend = "cmdstanr",
  file = "assets/bmmfit_cswald_vignette.rds"
)
```

## Inspecting Results

```{r}
summary(fit)
```

The model uses link functions for the parameters:

- `drift` uses a log link, so estimated values are on the log scale
- `bound` uses a log link
- `ndt` uses a log link

To get the parameters on the original scale, we exponentiate. Note that for the
simple version, we multiply `bound` by 2 to get the total boundary separation
(since the simple version estimates the distance to one boundary only):

```{r}
# extract fixed effects
fixef_est <- brms::fixef(fit)

# transform to original scale
# Note: bound is multiplied by 2 to convert from half-separation to full DDM boundary
cat("Estimated parameters:\n")
cat("drift (Condition A):", exp(fixef_est["drift_conditionA", "Estimate"]), "\n")
cat("drift (Condition B):", exp(fixef_est["drift_conditionB", "Estimate"]), "\n")
cat("bound (total separation):", 2*exp(fixef_est["bound_Intercept", "Estimate"]), "\n")
cat("ndt:", exp(fixef_est["ndt_Intercept", "Estimate"]), "\n")
```

Compare with true values:

```{r}
cat("\nTrue parameters:\n")
cat("drift (Condition A): 3\n")
cat("drift (Condition B): 1.5\n")
cat("bound: 1.5\n")
cat("ndt: 0.3\n")
```

## Model Diagnostics

Check posterior predictive distributions:

```{r fig.width=8, fig.height=4, message=FALSE}
brms::pp_check(fit, type = "dens_overlay", ndraws = 50) +
  labs(title = "Posterior Predictive Check")
```

# References
