---
title: "The Diffusion Decision Model (ddm)"
output: bookdown::html_document2
author:
  - Gidon Frischkorn
  - Ven Popov
bibliography: REFERENCES.bib
header-includes:
  - \usepackage{amsmath}
vignette: >
  %\VignetteIndexEntry{The Diffusion Decision Model (ddm)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
pkgdown:
  as_is: true
---

<style type="text/css">
div.main-container {
max-width: 800px !important;
}

p {
margin-top: 1.5em ;
margin-bottom: 1.5em ;
}
.author{
    display: none;
}
</style>

```{r, include = FALSE}
options(crayon.enabled = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "jpeg",
  dpi = 100
)
fansi::set_knit_hooks(knitr::knit_hooks, which = c("output","message","error"))
```

# Introduction

The Diffusion Decision Model (DDM) is one of the most widely used cognitive
models for analyzing response time (RT) and accuracy data in two-alternative
forced-choice (2AFC) tasks. Originally developed by @ratcliffTheoryMemoryRetrieval1978,
the DDM has become a cornerstone of cognitive psychology and neuroscience
research on decision-making processes.

The DDM is ideal when:

- You have RT data from a two-choice decision task with both correct and error responses
- You want to decompose decision-making into separate cognitive processes
- You need to account for speed-accuracy tradeoffs
- You want to model trial-to-trial variability in decision processes
- You're interested in linking cognitive parameters to neural or behavioral measures

The `bmm` package provides three versions of the DDM with increasing complexity:

- **3-parameter version**: Simplest model with drift, boundary, and non-decision time
- **4-parameter version**: Adds starting point bias
- **7-parameter version**: Full model including trial-to-trial variability parameters

# Theoretical Background

## The Diffusion Process

The DDM models decision-making as a noisy accumulation of evidence over time
toward one of two decision boundaries. The evidence accumulation process is
described by the Wiener diffusion process:

$$
dX(t) = v \cdot dt + s \cdot dW(t)
$$

where:

- $X(t)$ is the accumulated evidence at time $t$
- $v$ is the drift rate (average rate of evidence accumulation)
- $s$ is the within-trial noise (standard deviation of the diffusion process, typically fixed at 0.1 or 1)
- $dW(t)$ is a Wiener process (Brownian motion)

The process starts at position $z \cdot a$ (where $z$ is the starting point
parameter and $a$ is the boundary separation) and continues until it reaches
one of two absorbing boundaries at 0 (lower/incorrect) or $a$ (upper/correct).

## Model Parameters

### Core Parameters (3-parameter version)

1. **Drift rate ($v$)**: The average rate at which evidence accumulates toward
   the correct boundary. Higher drift rates lead to faster and more accurate
   decisions.
   - Typical range: -5 to 5 (task-dependent)
   - Interpretation: Quality of evidence, task difficulty, or attention

2. **Boundary separation ($a$)**: The amount of evidence required before making
   a decision. Larger boundaries lead to slower but more accurate responses.
   - Typical range: 0.5 to 3
   - Interpretation: Response caution, speed-accuracy tradeoff

3. **Non-decision time ($t_0$ or ndt)**: Time for processes unrelated to the
   decision (e.g., stimulus encoding, motor response).
   - Typical range: 0.1 to 0.5 seconds
   - Interpretation: Perceptual and motor latencies

### Extended Parameters (4-parameter version)

4. **Starting point bias ($z_r$)**: Relative starting point of evidence
   accumulation (0 = lower boundary, 1 = upper boundary).
   - Default: 0.5 (unbiased)
   - Interpretation: A priori response bias

### Variability Parameters (7-parameter version)

5. **Drift rate variability ($s_v$ or sdrift)**: Between-trial variability in
   drift rate.
   - Typical range: 0 to 2
   - Accounts for variable evidence quality across trials

6. **Non-decision time variability ($s_{t_0}$ or sndt)**: Between-trial
   variability in non-decision time.
   - Typical range: 0 to 0.2
   - Accounts for variable encoding/motor times

7. **Starting point variability ($s_z$ or szr)**: Between-trial variability in
   starting point.
   - Typical range: 0 to 0.5
   - Accounts for variable response bias across trials

## Predictions

The DDM predicts complete RT distributions for both correct and error
responses, including:

- Mean RT and accuracy
- RT variability (standard deviation)
- Shape of RT distributions (skewness, leading edge)
- Speed-accuracy tradeoffs
- Effects of experimental manipulations on different cognitive processes

# Implementation in bmm

## Computational Efficiency Considerations

The DDM likelihood requires numerical solution of the Fokker-Planck equation,
which can be computationally demanding. @henrichSevenparameterDiffusionModel2024
provide detailed benchmarks of their Stan implementation (which `bmm` uses).

**Important**: Runtime scales approximately linearly with the number of trials.
The estimates below are based on typical experimental datasets with **1000-2000
trials per participant** (across all conditions). Smaller datasets will be faster;
larger datasets will take proportionally longer.

**Runtime per iteration** (from Henrich et al., 2024):

- **3-parameter version**: ~0.5-1 second per iteration
- **4-parameter version**: ~0.8-1.5 seconds per iteration  
- **7-parameter version**: ~2-4 seconds per iteration

**Total runtime for typical analysis** with 4 chains × 2000 iterations (1000
warmup, 1000 sampling) and ~1500 trials per participant:

- **3-parameter**: ~20-40 minutes
- **4-parameter**: ~30-60 minutes
- **7-parameter**: ~2-3 hours

For datasets with different sizes, scale these estimates proportionally:

- **500 trials**: ~30% of the time above
- **3000 trials**: ~200% of the time above
- **10,000 trials**: ~650% of the time above

### Effect of Model Complexity on Runtime

**Critical consideration**: Runtime depends not just on the total number of trials,
but also on the number of parameters being estimated. Hierarchical models with
many subjects or experimental conditions can be substantially slower than
single-subject models with the same total trial count.

**Example comparison** (same total of 10,000 trials):

- **1 subject, 10,000 trials**: Fast baseline
- **10 subjects, 1,000 trials each**: ~2-3× slower (adds random effects per subject)
- **50 subjects, 200 trials each**: ~5-10× slower (many random effects + correlation structure)
- **50 subjects × 2 conditions, 100 trials each**: ~8-15× slower (adds condition effects + interactions)

**Why hierarchical models are slower**:

1. **More parameters**: Each subject adds random intercepts/slopes (e.g., 3-7
   parameters per subject for random effects)
2. **Correlation matrices**: Stan must sample correlation matrices for random
   effects, which is computationally intensive
3. **Increased model complexity**: More parameters means longer gradient
   computations for each iteration
4. **Convergence challenges**: Complex hierarchical structures may require more
   warmup iterations

**Practical implications**:

- A single-subject model with 10,000 trials might take 1 hour
- A 50-subject hierarchical model (2 conditions × 100 trials = 10,000 total trials)
  might take 8-15 hours for the same DDM version
- Adding experimental predictors (e.g., drift ~ condition1 * condition2) increases
  runtime proportionally to the number of parameters

**Strategies for large hierarchical models**:

1. **Simplify random effects structure**: Start with random intercepts only
   before adding random slopes
2. **Consider non-centered parameterization**: May help with convergence (bmm
   handles this automatically for some models)
3. **Use more informative priors**: Helps constrain the posterior space and
   speeds convergence
4. **Parallel computing**: Use within-chain parallelization for large datasets
   (requires cmdstanr backend with threading support)
5. **Subset testing**: Fit to a subset of subjects first to verify model
   specification and estimate total runtime

**Recommendations for efficient model fitting**:

1. Start with the 3-parameter version to test your model specification
2. Use hierarchical priors to regularize parameter estimates
3. Consider parallelizing chains across cores (included in runtime estimates above)
4. For large datasets (>10,000 trials), consider within-chain parallelization or
   reduce chains to 2 with more iterations
5. Use informative priors when possible to speed convergence
6. Test on a subset of data first (e.g., one participant) to verify model
   specification before fitting the full dataset
7. The 7-parameter version is most useful when you have:
   - Large sample sizes per condition (>100 trials)
   - Theoretical interest in variability parameters
   - Evidence that simpler models don't capture the data well

## Example Dataset: Ratcliff & Rouder (1998)

The `rr98` dataset from the `rtdists` package provides a classic example of
choice response time data. The dataset contains responses from a brightness
discrimination task where participants judged whether pixel arrays were "high"
or "low" brightness.

```{r, eval=FALSE}
library(bmm)
library(ggplot2)
library(dplyr)

# Load the dataset from rtdists
data(rr98, package = "rtdists")

# Remove outliers as in the original analysis
rr98 <- rr98[!rr98$outlier, ]

# Examine structure
str(rr98[, c("id", "rt", "response_num", "instruction", "strength")])

# Check accuracy by condition
rr98 %>%
  mutate(correct = (response_num == 1 & strength <= 16) | 
                   (response_num == 2 & strength > 16)) %>%
  group_by(id, instruction) %>%
  summarise(accuracy = mean(correct), .groups = "drop")
```

The dataset includes:

- **rt**: Response time in seconds
- **response_num**: Binary response (1 = "dark", 2 = "light")
- **instruction**: Speed vs accuracy emphasis
- **strength**: Task difficulty (0-32, brightness level)
- **id**: Participant identifier

This dataset is ideal for DDM analysis because:

1. **Sufficient error trials**: Unlike tasks with near-ceiling accuracy
2. **Multiple conditions**: Speed vs accuracy emphasis
3. **Clear manipulation**: Brightness strength affects evidence quality
4. **Rich structure**: Individual differences and trial-level variability

## Preparing Data for DDM

The DDM requires:

1. A numeric RT variable (positive values)
2. A binary response variable coded as 0/1 or "lower"/"upper"

```{r, eval=FALSE}
# Prepare data for DDM analysis
# Focus on one participant and instruction condition for demonstration
demo_data <- rr98 %>%
  filter(id == 1, instruction == "accuracy") %>%
  mutate(
    # Recode response: 1 = "upper", 0 = "lower"
    response = ifelse(response_num == 1, 0, 1)
  ) %>%
  select(rt, response, strength)

# Check response coding
table(demo_data$response)

# Visualize RT distributions by response
ggplot(demo_data, aes(x = rt, fill = factor(response))) +
  geom_histogram(alpha = 0.6, bins = 30, position = "identity") +
  labs(x = "Response Time (s)", y = "Count", fill = "Response") +
  theme_minimal()
```

# Fitting DDM Models in bmm

## 3-Parameter DDM

The simplest DDM with drift, boundary, and non-decision time:

```{r, eval=FALSE}
# Define the model
model_3par <- ddm(rt = "rt", 
                  response = "response", 
                  version = "3par")

# Define the formula - let drift vary by strength
formula_3par <- bmf(
  drift ~ strength,
  bound ~ 1,
  ndt ~ 1
)

# Fit the model 
# Note: demo_data has ~260 trials, so this will take ~5-10 minutes
# For larger datasets (1500+ trials), expect 20-40 minutes
fit_3par <- bmm(
  formula = formula_3par,
  data = demo_data,
  model = model_3par,
  backend = "cmdstanr",
  cores = 4,
  chains = 4,
  iter = 2000,
  warmup = 1000
)

# Examine convergence
summary(fit_3par)

# Check diagnostics
plot(fit_3par)
```

**Interpreting the 3-parameter model**:

- **drift**: Higher values indicate faster evidence accumulation (easier trials)
- **bound**: Higher values indicate more cautious responding
- **ndt**: Time spent on non-decision processes

## 4-Parameter DDM with Starting Point Bias

Adding starting point bias allows modeling a priori response preferences:

```{r, eval=FALSE}
# Define the 4-parameter model
model_4par <- ddm(rt = "rt", 
                  response = "response", 
                  version = "4par")

# Define the formula
formula_4par <- bmf(
  drift ~ strength,
  bound ~ 1,
  ndt ~ 1,
  zr ~ 1  # starting point bias
)

# Fit the model
# With ~260 trials: ~8-15 minutes
# With 1500+ trials: ~30-60 minutes
fit_4par <- bmm(
  formula = formula_4par,
  data = demo_data,
  model = model_4par,
  backend = "cmdstanr",
  cores = 4
)

# Compare models
loo_3par <- loo(fit_3par)
loo_4par <- loo(fit_4par)
loo_compare(loo_3par, loo_4par)
```

**When to use starting point bias**:

- When you suspect a priori response preferences
- When conditions systematically favor one response
- When testing theoretical predictions about response bias

## 7-Parameter DDM with Variability

The full DDM includes between-trial variability in drift, non-decision time,
and starting point:

```{r, eval=FALSE}
# Define the 7-parameter model
model_7par <- ddm(rt = "rt", 
                  response = "response", 
                  version = "7par")

# Define the formula
formula_7par <- bmf(
  drift ~ strength,
  bound ~ 1,
  ndt ~ 1,
  zr ~ 1,
  sdrift ~ 1,   # drift variability
  sndt ~ 1,     # non-decision time variability
  szr ~ 1       # starting point variability
)

# Fit the model
# With ~260 trials: ~30-45 minutes
# With 1500+ trials: ~2-3 hours
fit_7par <- bmm(
  formula = formula_7par,
  data = demo_data,
  model = model_7par,
  backend = "cmdstanr",
  cores = 4
)
```

**When to use the 7-parameter version**:

- Large datasets with sufficient statistical power
- Theoretical interest in trial-to-trial variability
- When simpler models show systematic misfit
- When comparing to classic DDM literature (which often uses variability parameters)

# Hierarchical DDM Models

One of the key advantages of fitting DDMs in a Bayesian framework with `bmm`
is the ability to easily specify hierarchical models that account for
individual differences.

## Multi-Subject Analysis

```{r, eval=FALSE}
# Use multiple subjects from rr98
multi_subject_data <- rr98 %>%
  filter(instruction == "accuracy") %>%
  mutate(response = ifelse(response_num == 1, 0, 1)) %>%
  select(id, rt, response, strength)

# Hierarchical 3-parameter model
model_hier <- ddm(rt = "rt", 
                  response = "response", 
                  version = "3par")

# Formula with random effects
formula_hier <- bmf(
  drift ~ strength + (strength | id),  # random slopes for drift by strength
  bound ~ 1 + (1 | id),                # random intercepts for boundary
  ndt ~ 1 + (1 | id)                   # random intercepts for ndt
)

# Fit hierarchical model
fit_hier <- bmm(
  formula = formula_hier,
  data = multi_subject_data,
  model = model_hier,
  backend = "cmdstanr",
  cores = 4
)

# Extract individual-level parameters
# This gives you person-specific drift, bound, and ndt estimates
```

**Benefits of hierarchical modeling**:

1. **Partial pooling**: Borrows strength across subjects for better estimates
2. **Individual differences**: Quantifies between-subject variability
3. **Handles unbalanced data**: Works even with different trial counts per subject
4. **Shrinkage**: Extreme individual estimates are pulled toward group mean

## Condition Effects

Test how experimental manipulations affect different cognitive processes:

```{r, eval=FALSE}
# Compare speed vs accuracy emphasis
full_data <- rr98 %>%
  mutate(response = ifelse(response_num == 1, 0, 1)) %>%
  select(id, rt, response, instruction, strength)

# Model with instruction effects
formula_instruction <- bmf(
  drift ~ strength + (strength | id),
  bound ~ instruction + (instruction | id),  # boundary varies by instruction
  ndt ~ 1 + (1 | id)
)

fit_instruction <- bmm(
  formula = formula_instruction,
  data = full_data,
  model = model_hier,
  backend = "cmdstanr",
  cores = 4
)
```

**Theoretical predictions**:

- **Speed emphasis**: Lower boundary (faster, less accurate)
- **Accuracy emphasis**: Higher boundary (slower, more accurate)
- **Drift**: Should primarily reflect stimulus strength, not instruction

# Model Comparison and Selection

## Comparing DDM Versions

Use information criteria to compare model fit:

```{r, eval=FALSE}
# Compute LOO for each model
loo_3par <- loo(fit_3par)
loo_4par <- loo(fit_4par)
loo_7par <- loo(fit_7par)

# Compare models
loo_compare(loo_3par, loo_4par, loo_7par)

# Check for influential observations
plot(loo_7par)
```

**General findings** (from simulation studies):

- 3-parameter model often sufficient when variability is low
- 4-parameter model improves fit when response bias present
- 7-parameter model necessary when trial-to-trial variability is substantial
- Overfitting risks with 7-parameter model on small datasets (<100 trials)

## Posterior Predictive Checks

Assess model adequacy by comparing predictions to data:

```{r, eval=FALSE}
# Generate posterior predictions
pp_3par <- posterior_predict(fit_3par)

# Compare predicted vs observed RT distributions
library(bayesplot)
ppc_dens_overlay(demo_data$rt, pp_3par[1:50, ])

# Quantile-probability plots (QP plots)
# These are the gold standard for evaluating DDM fits
# Shows predicted vs observed RT quantiles for correct/error responses
```

# Hypothesis Testing

## Testing Parameter Differences

Use Bayesian hypothesis testing to compare parameters across conditions:

```{r, eval=FALSE}
# Test if drift differs between conditions
hypothesis(fit_instruction, "driftinstructionspeed < 0")

# Test if boundary differs between conditions  
hypothesis(fit_instruction, "boundinstructionspeed < boundinstructionaccuracy")

# Joint hypothesis about multiple parameters
hypothesis(fit_instruction, c(
  "driftstrength > 0",
  "boundinstructionaccuracy > boundinstructionspeed"
))
```

## Bayes Factors

Compare specific hypotheses using bridge sampling:

```{r, eval=FALSE}
# Fit constrained model (drift doesn't vary by strength)
formula_null <- bmf(
  drift ~ 1 + (1 | id),
  bound ~ instruction + (instruction | id),
  ndt ~ 1 + (1 | id)
)

fit_null <- bmm(
  formula = formula_null,
  data = full_data,
  model = model_hier,
  backend = "cmdstanr"
)

# Compute Bayes factor (requires bridgesampling package)
# bf <- bayes_factor(fit_instruction, fit_null)
```

# Advanced Topics

## Custom Priors

Specify informative priors based on prior research or pilot data:

```{r, eval=FALSE}
# Define custom priors
custom_priors <- c(
  prior(normal(2, 1), nlpar = "drift", coef = "Intercept"),
  prior(normal(1, 0.5), nlpar = "bound", coef = "Intercept"),
  prior(normal(-1, 0.3), nlpar = "ndt", coef = "Intercept")  # on log scale
)

# Fit with custom priors
fit_custom <- bmm(
  formula = formula_3par,
  data = demo_data,
  model = model_3par,
  prior = custom_priors,
  backend = "cmdstanr"
)
```

## Fixing Parameters

Fix parameters to specific values for hypothesis testing:

```{r, eval=FALSE}
# Fix starting point to 0.5 (unbiased) while estimating others
formula_fixed <- bmf(
  drift ~ strength,
  bound ~ 1,
  ndt ~ 1,
  zr = 0.5  # fixed parameter
)

fit_fixed <- bmm(
  formula = formula_fixed,
  data = demo_data,
  model = model_4par,  # uses 4par model but fixes zr
  backend = "cmdstanr"
)
```

## Link Functions

By default, `bmm` uses appropriate link functions:

- **drift**: identity (can be positive or negative)
- **bound**: log (must be positive)
- **ndt**: log (must be positive)
- **zr**: logit (must be between 0 and 1)
- **sdrift, sndt**: log (must be positive)
- **szr**: logit (must be between 0 and 1)

You can customize links if needed:

```{r, eval=FALSE}
# Use log link for drift (forces positive values)
model_custom_links <- ddm(
  rt = "rt",
  response = "response",
  version = "3par",
  links = list(drift = "log")
)
```

# Comparison with Other Models

## DDM vs Censored Shifted Wald (cswald)

**DDM advantages**:

- Models both correct and error responses
- Accounts for response bias and variability
- Richer theoretical interpretation
- Standard in the field

**cswald advantages**:

- Much faster to fit (minutes vs hours)
- Simpler interpretation with fewer parameters
- Adequate for high-accuracy tasks
- More stable with small samples

**Recommendation**: Use cswald for initial exploration or when errors are rare
(<10%); use DDM for complete analysis when computational resources allow.

## DDM vs EZ-DDM (ezdm)

The EZ-DDM provides closed-form parameter estimates from summary statistics
(mean RT, accuracy, RT variance). It's extremely fast but:

- Only provides point estimates (no uncertainty quantification)
- Cannot model individual differences hierarchically
- Cannot include covariates
- Assumes specific parameter values (e.g., $z = 0.5$, $s_z = 0$)

**Recommendation**: Use ezdm for quick initial estimates or when you need to
process many datasets rapidly; use full DDM for formal inference.

# Practical Guidelines

## Sample Size Recommendations

Based on simulation studies and practical experience:

- **Minimum**: 50 trials per condition (for 3-parameter model)
- **Adequate**: 100-200 trials per condition (for 4-parameter model)
- **Recommended**: 200+ trials per condition (for 7-parameter model)
- **Hierarchical models**: At least 5-10 subjects per group

## Convergence Diagnostics

Always check:

1. **Rhat < 1.01** for all parameters (indicates chain convergence)
2. **ESS > 400** (effective sample size, indicates sufficient sampling)
3. **No divergent transitions** (indicates numerical issues)
4. **Trace plots** look like "hairy caterpillars"
5. **Posterior predictive checks** show good fit

## Common Issues and Solutions

**Problem**: Divergent transitions
- **Solution**: Increase `adapt_delta` (e.g., 0.95 or 0.99)

**Problem**: Slow mixing / low ESS
- **Solution**: Increase iterations, use informative priors, check reparameterization

**Problem**: Boundary estimates at extremes
- **Solution**: Check data quality, consider more informative priors

**Problem**: Non-decision time longer than minimum RT
- **Solution**: Check for fast outliers, consider data trimming

**Problem**: Starting point estimates at 0 or 1
- **Solution**: May indicate model misspecification or insufficient data

# References

<div id="refs"></div>

# Session Info

```{r}
sessionInfo()
```
