---
title: "The Censored Shifted Wald Model (cswald)"
output: bookdown::html_document2
author:
  - Gidon Frischkorn
  - Ven Popov
bibliography: REFERENCES.bib
header-includes:
  - \usepackage{amsmath}
vignette: >
  %\VignetteIndexEntry{The Censored Shifted Wald Model (cswald)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
pkgdown:
  as_is: true
---

<style type="text/css">
div.main-container {
max-width: 800px !important;
}

p {
margin-top: 1.5em ;
margin-bottom: 1.5em ;
}
.author{
    display: none;
}
</style>

```{r, include = FALSE}
options(crayon.enabled = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "jpeg",
  dpi = 100
)
fansi::set_knit_hooks(knitr::knit_hooks, which = c("output","message","error"))
```

# Introduction

The Censored Shifted Wald (cswald) model is a measurement model for choice
response time (RT) tasks, particularly suited for tasks with high accuracy
(few errors). The model was introduced by @millerRelationCensoredShifted2018
as a simplified alternative to the full Wiener diffusion model
[@ratcliffDiffusionDecisionModel2008] that is easier to estimate while still
capturing essential features of the decision-making process.

The cswald model is ideal when:

- You have RT data from a two-choice decision task
- Accuracy is relatively high (error rates < 20%)
- You want to separate different components of the decision process
- You need a computationally efficient alternative to the full diffusion model

# Theoretical Background

## The Shifted Wald Distribution

The Wald distribution (also known as the inverse Gaussian distribution)
describes the first-passage time of a Wiener process (Brownian motion with
drift) to a single absorbing boundary. In the context of decision-making,
this models the time it takes for evidence to accumulate to a decision
threshold.

The shifted Wald distribution adds a non-decision time component to account
for processes like stimulus encoding and motor response execution:

$$
RT = T_{decision} + T_{ndt}
$$

where $T_{decision}$ follows a Wald distribution and $T_{ndt}$ is the
non-decision time.

The probability density function of the shifted Wald is:

$$
f(t | v, a, \tau) = \frac{a}{\sqrt{2\pi (t-\tau)^3}}
\exp\left(-\frac{(a - v(t-\tau))^2}{2(t-\tau)}\right)
$$

for $t > \tau$, where:

- $v$ is the drift rate (rate of evidence accumulation)
- $a$ is the boundary (decision threshold)
- $\tau$ is the non-decision time

## Censoring for Error Responses

In tasks with high accuracy, most responses are correct. The "censored"
aspect of the model refers to how error responses are handled. Instead of
modeling errors as arising from a separate accumulation process toward the
wrong boundary (as in the full diffusion model), errors are treated as
**censored** observations — trials where the correct response had not yet
been reached.

This censoring approach assumes that:

1. Correct responses follow the shifted Wald distribution
2. Error responses represent trials where the accumulation process was
   interrupted before reaching the correct boundary

The likelihood for a correct response is the Wald density:

$$
L_{correct}(t) = f_{Wald}(t - \tau | v, a)
$$

The likelihood for an error response is the survivor function (1 - CDF):

$$
L_{error}(t) = S_{Wald}(t - \tau | v, a) = 1 - F_{Wald}(t - \tau | v, a)
$$

## Relationship to the Diffusion Model

The cswald model can be seen as a simplification of the Wiener diffusion
model that makes certain assumptions:

1. **Single boundary**: Only the correct response boundary is explicitly
   modeled
2. **Unbiased starting point**: The starting point is assumed to be
   equidistant from both boundaries
3. **High accuracy**: The model works best when error rates are low

For tasks with substantial error rates (> 20%), the competing risks version
of the model (`version = "crisk"`) provides a more appropriate fit by
modeling both boundaries explicitly.

# Model Versions

The `bmm` package implements two versions of the cswald model:

## Simple Version (`version = "simple"`)

The simple version treats errors as censored correct responses. It estimates:

- **drift**: The drift rate (rate of evidence accumulation toward the
  correct response)
- **bound**: The distance from the starting point to the correct boundary
- **ndt**: Non-decision time
- **s**: Diffusion constant (usually fixed to 1 for scaling)

This version is best suited for tasks with error rates below 20%.

**Important note on boundary parameterization:** The `bound` parameter in the
simple version represents only the distance from the (unbiased) starting point
to the correct response boundary. This is **half** the total boundary
separation in the standard diffusion model parameterization. To convert the
simple version's `bound` estimate to the equivalent DDM or crisk boundary
separation, multiply by 2:

$$
a_{DDM} = 2 \times \text{bound}_{simple}
$$

## Competing Risks Version (`version = "crisk"`)

The competing risks version models both correct and error responses as
arising from racing accumulators toward opposite boundaries. It estimates:

- **drift**: The drift rate (can be positive or negative; negative drift
  indicates evidence accumulation toward the lower boundary)
- **bound**: Total boundary separation (full distance between both boundaries,
  consistent with DDM parameterization)
- **ndt**: Non-decision time
- **zr**: Relative starting point (proportion of boundary separation,
  0.5 = unbiased)
- **s**: Diffusion constant

This version is better suited for tasks with substantial error rates. The
`bound` parameter is directly comparable to the boundary separation in the
standard diffusion model.

## When to Use Each Version

### Use `version = "simple"` when:

- Error rates are low (< 20%)
- You want a computationally efficient model
- The focus is on correct response times
- Errors can reasonably be treated as censored observations

### Use `version = "crisk"` when:

- Error rates are substantial (> 20%)
- You want to model both response types explicitly
- Starting point bias might be present
- You need estimates comparable to full diffusion model parameters

## Comparison with the Wiener Diffusion Model

The `cswald` model in `bmm` is related to but distinct from the full Wiener
diffusion model available in `brms::wiener()`. Key differences:

| Feature | cswald (simple) | cswald (crisk) | brms::wiener |
|---------|-----------------|----------------|--------------|
| Boundaries | Single | Both | Both |
| Starting point | Fixed (unbiased) | Estimated | Estimated |
| Bound param | Half separation | Full separation | Full separation |
| Drift sign | Positive only | Positive/negative | Positive/negative |
| Best for | High accuracy | Moderate accuracy | Any accuracy |
| Computation | Fast | Moderate | Slower |
| Parameters | 3-4 | 4-5 | 4-5 |

For tasks with low error rates, the `cswald` model provides equivalent
inferences with faster computation. When comparing results across model
versions, remember to multiply the simple version's `bound` by 2 to get the
equivalent total boundary separation.

# Fitting the Model with `bmm`

## Setup

```{r setup, message=FALSE, warning=FALSE}
library(bmm)
library(ggplot2)
```

## Example Dataset: Ratcliff & Rouder (1998)

The `rr98` dataset from the `rtdists` package provides a classic example of
choice response time data suitable for evidence accumulation models. The
dataset contains responses from a brightness discrimination task where
participants judged whether pixel arrays were "high" or "low" brightness.

```{r}
# Load the dataset from rtdists (already a dependency of bmm)
data(rr98, package = "rtdists")

# Remove outliers as in the original analysis
rr98 <- rr98[!rr98$outlier, ]

# Examine structure
str(rr98[, c("id", "rt", "response_num", "instruction", "strength")])

# Check error rates by condition
with(rr98, tapply(response_num == 1, list(id, instruction), mean))
```

The dataset includes:

- **rt**: Response time in seconds
- **response_num**: Binary response (1 = "dark", 2 = "light")
- **instruction**: Speed vs accuracy emphasis
- **strength**: Task difficulty (0-32, brightness level)
- **id**: Participant identifier

This dataset is particularly well-suited for the cswald model because:

1. **High accuracy**: Most conditions have error rates below 20%, ideal for the
   simple version
2. **Multiple conditions**: Allows testing how task manipulations affect model
   parameters
3. **Trial-level data**: Contains the RT and response for every trial
4. **Classic reference**: Widely used in the evidence accumulation literature

For a complete analysis example using this dataset, see the `rtdists` package
vignette: `vignette("reanalysis_rr98", package = "rtdists")`.

## Generating Simulated Data with Hierarchical Structure

We'll simulate data from multiple participants with individual differences in
drift rates, demonstrating how to fit hierarchical models with `bmm`:

```{r}
# set seed for reproducibility
set.seed(42)

# simulate data for 3 participants in two conditions
n_subjects <- 5
n_per_condition <- 200  # fewer trials per condition to keep file size manageable

# Population-level parameters
pop_drift_A <- 3.0      # mean drift for condition A
pop_drift_B <- 1.8      # mean drift for condition B
drift_sd <- 0.5         # between-subject SD in drift
bound <- 1.5            # shared boundary
ndt <- 0.3              # shared non-decision time

# Generate subject-specific drift rates
subject_drifts_A <- rnorm(n_subjects, pop_drift_A, drift_sd)
subject_drifts_B <- rnorm(n_subjects, pop_drift_B, drift_sd)

# Simulate data for each subject and condition
dat_list <- list()
for (i in 1:n_subjects) {
  # Condition A
  dat_A <- rcswald(n = n_per_condition, 
                   drift = subject_drifts_A[i], 
                   bound = bound, 
                   ndt = ndt)
  dat_A$id <- paste0("S", i)
  dat_A$condition <- "A"
  
  # Condition B
  dat_B <- rcswald(n = n_per_condition, 
                   drift = subject_drifts_B[i], 
                   bound = bound, 
                   ndt = ndt)
  dat_B$id <- paste0("S", i)
  dat_B$condition <- "B"
  
  dat_list[[2*i-1]] <- dat_A
  dat_list[[2*i]] <- dat_B
}

# Combine all data
dat <- do.call(rbind, dat_list)
dat$id <- factor(dat$id)
dat$condition <- factor(dat$condition)

# Check error rates by subject and condition
error_rates <- aggregate(response ~ id + condition, data = dat, 
                        FUN = function(x) mean(x == 0))
print(error_rates)
```

Let's visualize the RT distributions by participant:

```{r fig.width=8, fig.height=6}
ggplot(dat, aes(x = rt, fill = factor(response))) +
  geom_histogram(binwidth = 0.05, position = "identity", alpha = 0.7) +
  facet_grid(id ~ condition, labeller = labeller(
    condition = c(A = "Condition A", B = "Condition B")
  )) +
  scale_fill_manual(values = c("0" = "red", "1" = "blue"),
                    labels = c("0" = "Error", "1" = "Correct"),
                    name = "Response") +
  labs(x = "Response Time (s)", y = "Count",
       title = "RT distributions by participant and condition") +
  theme_classic()
```

## Specifying the Model

First, specify the model using `cswald()`:

```{r}
model <- cswald(
  rt = "rt",
  response = "response",
  version = "simple"
)
```

## Specifying the Hierarchical Formula

We want to estimate:

1. **Population-level effects**: How drift rate varies by condition
2. **Individual differences**: Random effects for drift rate by participant
3. **Shared parameters**: Boundary and non-decision time constant across
   participants

The formula includes both fixed effects for condition and random effects
(varying intercepts) for participants:

```{r}
formula <- bmf(
  drift ~ 0 + condition + (0 + condition | id),  # fixed condition effects + random subject effects
  bound ~ 1,                          # shared boundary
  ndt ~ 1                             # shared non-decision time
)
```

The `(0 + condition | id)` term in the drift formula specifies random slopes
for each condition by participant. This allows the model to estimate:

- A population-level drift rate for each condition
- Individual deviations from these population means for each condition
- The variability (SD) and correlation of these individual differences across conditions

This hierarchical structure provides several advantages:

1. **Partial pooling**: Individual estimates are pulled toward the group mean,
   providing more stable estimates for participants with less data
2. **Uncertainty quantification**: The model quantifies both within- and
   between-subject variability
3. **Shrinkage**: Extreme individual estimates are regularized toward the group
   mean

## Fitting the Hierarchical Model

Now we fit the model with hierarchical structure. The random effects allow
each participant to have their own drift rate while estimating the
population-level mean drift for each condition:

```{r, message=FALSE, results='hide'}
fit <- bmm(
  formula = formula,
  data = dat,
  model = model,
  cores = 4,
  backend = "cmdstanr",
  # Reduce iterations to keep file size manageable
  iter = 1000,
  warmup = 500,
  control = list(adapt_delta = 0.9),
  file = "assets/bmmfit_cswald_hierarchical_vignette.rds"
)
```

## Inspecting Results

```{r}
summary(fit)
```

### Population-Level Effects

The model uses link functions for the parameters (log link for all parameters),
so we need to exponentiate to get values on the original scale:

```{r}
# Extract population-level (fixed) effects
fixef_est <- brms::fixef(fit)

# Transform to original scale
cat("Population-level estimates:\n")
cat("drift (Condition A):", exp(fixef_est["drift_conditionA", "Estimate"]), "\n")
cat("drift (Condition B):", exp(fixef_est["drift_conditionB", "Estimate"]), "\n")
cat("bound (half-separation):", exp(fixef_est["bound_Intercept", "Estimate"]), "\n")
cat("ndt:", exp(fixef_est["ndt_Intercept", "Estimate"]), "\n")

cat("\nTrue population-level parameters:\n")
cat("drift (Condition A):", pop_drift_A, "\n")
cat("drift (Condition B):", pop_drift_B, "\n")
cat("bound:", bound, "\n")
cat("ndt:", ndt, "\n")
```

### Individual Differences

The random effects capture individual differences in drift rate:

```{r}
# Extract random effects (deviations from population mean on log scale)
ranef_est <- brms::ranef(fit)

# Get subject-specific drift estimates by adding random effects to fixed effects
cat("\nSubject-specific drift estimates (Condition A):\n")
for (i in 1:n_subjects) {
  subj_id <- paste0("S", i)
  # Add random effect to fixed effect, then exponentiate
  drift_subj <- exp(fixef_est["drift_conditionA", "Estimate"] + 
                    ranef_est$id[subj_id, "Estimate", "drift_conditionA"])
  cat(subj_id, ": ", round(drift_subj, 2), 
      " (true: ", round(subject_drifts_A[i], 2), ")\n", sep = "")
}

cat("\nSubject-specific drift estimates (Condition B):\n")
for (i in 1:n_subjects) {
  subj_id <- paste0("S", i)
  drift_subj <- exp(fixef_est["drift_conditionB", "Estimate"] + 
                    ranef_est$id[subj_id, "Estimate", "drift_conditionB"])
  cat(subj_id, ": ", round(drift_subj, 2), 
      " (true: ", round(subject_drifts_B[i], 2), ")\n", sep = "")
}

# Between-subject variability
cat("\nEstimated SD of random effects (log scale, Cond A | Cond B):\n")
cat("sd(drift):", brms::VarCorr(fit)$id$sd[,"Estimate"], "\n")
cat("True SD (log scale):", drift_sd, "\n")
```

## Model Diagnostics

Check posterior predictive distributions. By default `pp_check()` shows the
response time distribution; `resp_var = "response"` checks the response
proportions and `resp_var = "signed_rt"` both jointly (see
`pp_check_vars(fit)` for all options):

```{r fig.width=8, fig.height=4, message=FALSE}
pp_check(fit, type = "dens_overlay", ndraws = 10) +
  labs(title = "Posterior Predictive Check")
```

## Hypothesis Testing

We can test whether drift rates differ between conditions using the
`brms::hypothesis()` function. This tests the difference on the log scale:

```{r}
# Test if drift rate is higher in Condition A than B
hyp_test <- brms::hypothesis(fit, "drift_conditionA > drift_conditionB")
print(hyp_test)
```

The output shows:

- **Estimate**: The posterior mean of the difference (on log scale)
- **Est.Error**: The standard error of the difference
- **CI**: The 95% credible interval for the difference
- **Evid.Ratio**: The evidence ratio (posterior odds) in favor of the hypothesis
- **Post.Prob**: The posterior probability that the hypothesis is true

To express the difference on the original scale, we can compute the ratio of
drift rates:

```{r}
# Extract posterior samples
posterior <- brms::as_draws_df(fit)

# Compute drift ratio (Condition A / Condition B) on original scale
drift_ratio <- exp(posterior$b_drift_conditionA - posterior$b_drift_conditionB)

cat("Drift rate ratio (A/B):\n")
cat("  Median:", round(median(drift_ratio), 2), "\n")
cat("  95% CI: [", round(quantile(drift_ratio, 0.025), 2), ", ",
    round(quantile(drift_ratio, 0.975), 2), "]\n", sep = "")
cat("  P(ratio > 1):", round(mean(drift_ratio > 1), 3), "\n")

cat("\nTrue drift ratio:", round(pop_drift_A / pop_drift_B, 2), "\n")
```

This shows that the model successfully recovers the difference between
conditions, with strong evidence that drift rates are higher in Condition A.

# References
