---
title: "The Diffusion Decision Model (ddm)"
output: bookdown::html_document2
author:
  - Gidon Frischkorn
  - Ven Popov
bibliography: REFERENCES.bib
header-includes:
  - \usepackage{amsmath}
vignette: >
  %\VignetteIndexEntry{The Diffusion Decision Model (ddm)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
pkgdown:
  as_is: true
---

<style type="text/css">
div.main-container {
max-width: 800px !important;
}

p {
margin-top: 1.5em ;
margin-bottom: 1.5em ;
}
.author{
    display: none;
}
</style>

```{r, include = FALSE}
options(crayon.enabled = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "jpeg",
  dpi = 100
)
fansi::set_knit_hooks(knitr::knit_hooks, which = c("output","message","error"))
```

# Introduction

The Diffusion Decision Model (DDM) is one of the most widely used cognitive
models for analyzing response time (RT) and accuracy data in two-alternative
forced-choice (2AFC) tasks. Originally developed by @Ratcliff_1978,
the DDM has become a cornerstone of cognitive psychology and neuroscience
research on decision-making processes.

The DDM is ideal when:

- You have RT data from a two-choice decision task with both correct and error responses
- You want to decompose decision-making into separate cognitive processes
- You need to account for speed-accuracy tradeoffs
- You want to model trial-to-trial variability in decision processes
- You're interested in linking cognitive parameters to neural or behavioral measures

The `bmm` package provides two versions of the DDM with increasing complexity:

- **3-parameter version**: Simplest model with drift, boundary, and non-decision time
- **4-parameter version**: Adds starting point bias

# Theoretical Background

## The Diffusion Process

The DDM models decision-making as a noisy accumulation of evidence over time
toward one of two decision boundaries. The evidence accumulation process is
described by the Wiener diffusion process:

$$
dX(t) = v \cdot dt + s \cdot dW(t)
$$

where:

- $X(t)$ is the accumulated evidence at time $t$
- $v$ is the drift rate (average rate of evidence accumulation)
- $s$ is the within-trial noise (standard deviation of the diffusion process, typically fixed at 0.1 or 1)
- $dW(t)$ is a Wiener process (Brownian motion)

The process starts at position $z \cdot a$ (where $z$ is the starting point
parameter and $a$ is the boundary separation) and continues until it reaches
one of two absorbing boundaries at 0 (lower/incorrect) or $a$ (upper/correct).

## Model Parameters

### Core Parameters (3-parameter version)

1. **Drift rate ($v$)**: The average rate at which evidence accumulates toward
   the correct boundary. Higher drift rates lead to faster and more accurate
   decisions.
   - Typical range: -5 to 5 (task-dependent)
   - Interpretation: Quality of evidence, task difficulty, or attention

2. **Boundary separation ($a$)**: The amount of evidence required before making
   a decision. Larger boundaries lead to slower but more accurate responses.
   - Typical range: 0.5 to 3
   - Interpretation: Response caution, speed-accuracy tradeoff

3. **Non-decision time ($t_0$ or ndt)**: Time for processes unrelated to the
   decision (e.g., stimulus encoding, motor response).
   - Typical range: 0.1 to 0.5 seconds
   - Interpretation: Perceptual and motor latencies

### Extended Parameters (4-parameter version)

4. **Starting point bias ($z_r$)**: Relative starting point of evidence
   accumulation (0 = lower boundary, 1 = upper boundary).
   - Default: 0.5 (unbiased)
   - Interpretation: A priori response bias

### Variability Parameters (7-parameter DDM)

The full DDM, as described by @ratcliffDiffusionDecisionModel2008 and
implemented by @henrichSevenparameterDiffusionModel2024, includes three
additional parameters that capture trial-to-trial variability in the core
parameters:

5. **Drift rate variability ($s_v$ or sdrift)**: Between-trial variability in
   drift rate, accounting for fluctuations in evidence quality across trials
   (typical range: 0 to 2).

6. **Non-decision time variability ($s_{t_0}$ or sndt)**: Between-trial
   variability in non-decision time, accounting for variable encoding/motor
   times (typical range: 0 to 0.2).

7. **Starting point variability ($s_z$ or szr)**: Between-trial variability in
   starting point, accounting for variable response bias across trials (typical
   range: 0 to 0.5).

**Why the 7-parameter version is not implemented in `bmm`**:

While these variability parameters are theoretically important and can improve
model fit, @henrichSevenparameterDiffusionModel2024 demonstrated that the
7-parameter likelihood has prohibitively long computation times, especially for
hierarchical Bayesian estimation. Key limitations include:

- **Single-subject feasibility only**: Even with 100-300 trials, model fitting
  can take hours. With 1500+ trials typical for robust DDM estimation, runtimes
  extend to many hours or days.

- **Hierarchical models infeasible**: The most common use case in cognitive
  research (multiple subjects, conditions, and trials) becomes computationally
  intractable without high-performance computing resources. A typical
  hierarchical study (20+ subjects, 2-3 conditions, 200+ trials per condition)
  would require weeks of computation time.

- **Limited practical benefit**: For most research questions, the 3- and
  4-parameter versions provide excellent fits and interpretable parameters,
  making the substantial computational cost of the 7-parameter version difficult
  to justify.

For researchers specifically interested in trial-to-trial variability
parameters, we recommend considering simulation-based inference methods (e.g.,
BayesFlow) or fitting separate single-subject models rather than hierarchical
approaches.

## Predictions

The DDM predicts complete RT distributions for both correct and error
responses, including:

- Mean RT and accuracy
- RT variability (standard deviation)
- Shape of RT distributions (skewness, leading edge)
- Speed-accuracy tradeoffs
- Effects of experimental manipulations on different cognitive processes

# Implementation in bmm

## Computational Efficiency Considerations

The DDM likelihood requires numerical solution of the Fokker-Planck equation, which can be computationally demanding for large datasets. Runtime scales approximately linearly with the number of trials. For typical datasets (100-300 trials), both 3-parameter and 4-parameter models fit quickly (< 20 seconds on modern hardware with 4 chains × 2000 iterations). Larger datasets (1500+ trials) may take several minutes for single-subject models and several hours for hierarchical models with many subjects.

**Practical recommendations**: Start with the 3-parameter version to test your model specification. Use the default `cores = 4` to parallelize chains efficiently. For hierarchical models, consider starting with random intercepts only before adding random slopes. Test on a subset of data first to verify model specification and estimate total runtime.

## Example Dataset: Ratcliff & Rouder (1998)

The `rr98` dataset from the `rtdists` package provides a classic example of
choice response time data. The dataset contains responses from a brightness
discrimination task where participants judged whether pixel arrays were "high"
or "low" brightness.

```{r, eval=FALSE}
library(bmm)
library(ggplot2)
library(dplyr)

# Load the dataset from rtdists
data(rr98, package = "rtdists")

# Remove outliers as in the original analysis
rr98 <- rr98[!rr98$outlier, ]

# Examine structure
str(rr98[, c("id", "rt", "response_num", "instruction", "strength")])

# Check accuracy by condition
rr98 %>%
  mutate(correct = (response_num == 1 & strength <= 16) | 
                   (response_num == 2 & strength > 16)) %>%
  group_by(id, instruction) %>%
  summarise(accuracy = mean(correct), .groups = "drop")
```

The dataset includes:

- **rt**: Response time in seconds
- **response_num**: Binary response (1 = "dark", 2 = "light")
- **instruction**: Speed vs accuracy emphasis
- **strength**: Task difficulty (0-32, brightness level)
- **id**: Participant identifier

This dataset is ideal for DDM analysis because:

1. **Sufficient error trials**: Unlike tasks with near-ceiling accuracy
2. **Multiple conditions**: Speed vs accuracy emphasis
3. **Clear manipulation**: Brightness strength affects evidence quality
4. **Rich structure**: Individual differences and trial-level variability

## Preparing Data for DDM

The DDM requires:

1. A numeric RT variable (positive values)
2. A binary response variable coded as 0/1 or "lower"/"upper"

```{r, eval=FALSE}
# Prepare data for DDM analysis
# Focus on one participant and instruction condition for demonstration
demo_data <- rr98 %>%
  dplyr::filter(id == "jf", instruction == "accuracy") %>%
  mutate(
    # Recode response: 1 = "upper", 0 = "lower"
    response = ifelse(response_num == 1, 0, 1)
  )

# Check response coding
table(demo_data$response)

# Visualize RT distributions by response
ggplot(demo_data, aes(x = rt, fill = factor(response))) +
  facet_wrap(~ source) +
  geom_histogram(alpha = 0.6, bins = 30, position = "identity") +
  labs(x = "Response Time (s)", y = "Count", fill = "Response") +
  theme_minimal()
```

# Fitting DDM Models in bmm

## 3-Parameter DDM

The simplest DDM with drift, boundary, and non-decision time:

```{r, eval=FALSE}
# Define the model
model_3par <- ddm(rt = "rt", 
                  response = "response", 
                  version = "3par")

# Define the formula - let drift vary by strength
formula_3par <- bmf(
  drift ~ 0 + source,
  bound ~ 1,
  ndt ~ 1
)

# Fit the model 
fit_3par <- bmm(
  formula = formula_3par,
  data = demo_data,
  model = model_3par,
  backend = "cmdstanr",
  cores = 4,
  chains = 4,
  iter = 2000,
  warmup = 1000,
  file = "assets/bmmfit_ddm_3par_vignette"
)

# Examine convergence
summary(fit_3par)

# Check diagnostics
plot(fit_3par)
```

**Interpreting the 3-parameter model**:

- **drift**: Higher values indicate faster evidence accumulation (easier trials)
- **bound**: Higher values indicate more cautious responding
- **ndt**: Time spent on non-decision processes

## 4-Parameter DDM with Starting Point Bias

Adding starting point bias allows modeling a priori response preferences:

```{r, eval=FALSE}
# Define the 4-parameter model
model_4par <- ddm(rt = "rt", 
                  response = "response", 
                  version = "4par")

# Define the formula
formula_4par <- bmf(
  drift ~ 0 + source,
  bound ~ 1,
  ndt ~ 1,
  zr ~ 1  # starting point bias
)

# Fit the model
fit_4par <- bmm(
  formula = formula_4par,
  data = demo_data,
  model = model_4par,
  backend = "cmdstanr",
  cores = 4,
  file = "assets/bmmfit_ddm_4par_vignette"
)

# Compare models
loo_3par <- loo(fit_3par)
loo_4par <- loo(fit_4par)
loo_compare(loo_3par, loo_4par)
```

**When to use starting point bias**:

- When you suspect a priori response preferences
- When conditions systematically favor one response
- When testing theoretical predictions about response bias

# Hierarchical DDM Models

One of the key advantages of fitting DDMs in a Bayesian framework with `bmm`
is the ability to easily specify hierarchical models that account for
individual differences.

## Multi-Subject Analysis

```{r, eval=FALSE}
# Use multiple subjects from rr98
multi_subject_data <- rr98 %>%
  filter(instruction == "accuracy") %>%
  mutate(response = ifelse(response_num == 1, 0, 1)) %>%
  select(id, rt, response, strength)

# Hierarchical 3-parameter model
model_hier <- ddm(rt = "rt", 
                  response = "response", 
                  version = "3par")

# Formula with random effects
formula_hier <- bmf(
  drift ~ 0 + source + (0 + source | id),  # random slopes for drift by strength
  bound ~ 1 + (1 | id),                # random intercepts for boundary
  ndt ~ 1 + (1 | id)                   # random intercepts for ndt
)

# Fit hierarchical model
fit_hier <- bmm(
  formula = formula_hier,
  data = multi_subject_data,
  model = model_hier,
  backend = "cmdstanr",
  cores = 4
)

# Extract individual-level parameters
# This gives you person-specific drift, bound, and ndt estimates
```

**Benefits of hierarchical modeling**:

1. **Partial pooling**: Borrows strength across subjects for better estimates
2. **Individual differences**: Quantifies between-subject variability
3. **Handles unbalanced data**: Works even with different trial counts per subject
4. **Shrinkage**: Extreme individual estimates are pulled toward group mean

## Condition Effects

Test how experimental manipulations affect different cognitive processes:

```{r, eval=FALSE}
# Compare speed vs accuracy emphasis
full_data <- rr98 %>%
  mutate(response = ifelse(response_num == 1, 0, 1)) %>%
  select(id, rt, response, instruction, strength)

# Model with instruction effects
formula_instruction <- bmf(
  drift ~ 0 + source + (0 + source| id),
  bound ~ instruction + (instruction | id),  # boundary varies by instruction
  ndt ~ 1 + (1 | id)
)

fit_instruction <- bmm(
  formula = formula_instruction,
  data = full_data,
  model = model_hier,
  backend = "cmdstanr",
  cores = 4
)
```

**Theoretical predictions**:

- **Speed emphasis**: Lower boundary (faster, less accurate)
- **Accuracy emphasis**: Higher boundary (slower, more accurate)
- **Drift**: Should primarily reflect stimulus strength, not instruction

# Model Comparison and Selection

## Comparing DDM Versions

Use information criteria to compare model fit:

```{r, eval=FALSE}
# Compute LOO for each model
loo_3par <- loo(fit_3par)
loo_4par <- loo(fit_4par)

# Compare models
loo_compare(loo_3par, loo_4par)

# Check for influential observations
plot(loo_4par)
```

**General findings** (from simulation studies):

- 3-parameter model often sufficient for most applications
- 4-parameter model improves fit when response bias is present
- Model selection should balance fit quality with interpretability

## Posterior Predictive Checks

Assess model adequacy by comparing predictions to data:

```{r, eval=FALSE}
# Generate posterior predictions
pp_3par <- posterior_predict(fit_3par)

# Compare predicted vs observed RT distributions
library(bayesplot)
ppc_dens_overlay(demo_data$rt, pp_3par[1:50, ])

# Quantile-probability plots (QP plots)
# These are the gold standard for evaluating DDM fits
# Shows predicted vs observed RT quantiles for correct/error responses
```

# Hypothesis Testing

## Testing Parameter Differences

Use Bayesian hypothesis testing to compare parameters across conditions:

```{r, eval=FALSE}
# Test if drift differs between conditions
hypothesis(fit_instruction, "driftinstructionspeed < 0")

# Test if boundary differs between conditions  
hypothesis(fit_instruction, "boundinstructionspeed < boundinstructionaccuracy")

# Joint hypothesis about multiple parameters
hypothesis(fit_instruction, c(
  "driftstrength > 0",
  "boundinstructionaccuracy > boundinstructionspeed"
))
```

## Bayes Factors

Compare specific hypotheses using bridge sampling:

```{r, eval=FALSE}
# Fit constrained model (drift doesn't vary by strength)
formula_null <- bmf(
  drift ~ 1 + (1 | id),
  bound ~ instruction + (instruction | id),
  ndt ~ 1 + (1 | id)
)

fit_null <- bmm(
  formula = formula_null,
  data = full_data,
  model = model_hier,
  backend = "cmdstanr"
)

# Compute Bayes factor (requires bridgesampling package)
# bf <- bayes_factor(fit_instruction, fit_null)
```

# Advanced Topics

## Custom Priors

Specify informative priors based on prior research or pilot data:

```{r, eval=FALSE}
# Define custom priors
custom_priors <- c(
  prior(normal(2, 1), nlpar = "drift", coef = "Intercept"),
  prior(normal(1, 0.5), nlpar = "bound", coef = "Intercept"),
  prior(normal(-1, 0.3), nlpar = "ndt", coef = "Intercept")  # on log scale
)

# Fit with custom priors
fit_custom <- bmm(
  formula = formula_3par,
  data = demo_data,
  model = model_3par,
  prior = custom_priors,
  backend = "cmdstanr"
)
```

## Fixing Parameters

Fix parameters to specific values for hypothesis testing:

```{r, eval=FALSE}
# Fix starting point to 0.5 (unbiased) while estimating others
formula_fixed <- bmf(
  drift ~ strength,
  bound ~ 1,
  ndt ~ 1,
  zr = 0.5  # fixed parameter
)

fit_fixed <- bmm(
  formula = formula_fixed,
  data = demo_data,
  model = model_4par,  # uses 4par model but fixes zr
  backend = "cmdstanr"
)
```

## Link Functions

By default, `bmm` uses appropriate link functions:

- **drift**: identity (can be positive or negative)
- **bound**: log (must be positive)
- **ndt**: log (must be positive)
- **zr**: logit (must be between 0 and 1)
- **sdrift, sndt**: log (must be positive)
- **szr**: logit (must be between 0 and 1)

You can customize links if needed:

```{r, eval=FALSE}
# Use log link for drift (forces positive values)
model_custom_links <- ddm(
  rt = "rt",
  response = "response",
  version = "3par",
  links = list(drift = "log")
)
```

# Comparison with Other Models

## DDM vs Censored Shifted Wald (cswald)

**DDM advantages**:

- Models both correct and error responses
- Accounts for response bias and variability
- Richer theoretical interpretation
- Standard in the field

**cswald advantages**:

- Much faster to fit (minutes vs hours)
- Simpler interpretation with fewer parameters
- Adequate for high-accuracy tasks
- More stable with small samples

**Recommendation**: Use cswald for initial exploration or when errors are rare
(<10%); use DDM for complete analysis when computational resources allow.

## DDM vs EZ-DDM (ezdm)

The EZ-DDM provides closed-form parameter estimates from summary statistics
(mean RT, accuracy, RT variance). It's extremely fast but:

- Only provides point estimates (no uncertainty quantification)
- Cannot model individual differences hierarchically
- Cannot include covariates
- Assumes specific parameter values (e.g., $z = 0.5$, $s_z = 0$)

**Recommendation**: Use ezdm for quick initial estimates or when you need to
process many datasets rapidly; use full DDM for formal inference.

# Practical Guidelines

## Sample Size Recommendations

Based on simulation studies and practical experience:

- **Minimum**: 50 trials per condition (for 3-parameter model)
- **Adequate**: 100-200 trials per condition (for 4-parameter model)
- **Recommended**: 200+ trials per condition for stable estimates
- **Hierarchical models**: At least 5-10 subjects per group

## Convergence Diagnostics

Always check:

1. **Rhat < 1.01** for all parameters (indicates chain convergence)
2. **ESS > 400** (effective sample size, indicates sufficient sampling)
3. **No divergent transitions** (indicates numerical issues)
4. **Trace plots** look like "hairy caterpillars"
5. **Posterior predictive checks** show good fit

## Common Issues and Solutions

**Problem**: Divergent transitions
- **Solution**: Increase `adapt_delta` (e.g., 0.95 or 0.99)

**Problem**: Slow mixing / low ESS
- **Solution**: Increase iterations, use informative priors, check reparameterization

**Problem**: Boundary estimates at extremes
- **Solution**: Check data quality, consider more informative priors

**Problem**: Non-decision time longer than minimum RT
- **Solution**: Check for fast outliers, consider data trimming

**Problem**: Starting point estimates at 0 or 1
- **Solution**: May indicate model misspecification or insufficient data

# References

<div id="refs"></div>

