---
title: "Extracting default priors, the generated Stan code and Stan data"
output: bookdown::html_document2
author: 
  - Ven Popov
  - Gidon Frischkorn
bibliography: REFERENCES.bib
header-includes:
  - \usepackage{amsmath}
vignette: >
  %\VignetteIndexEntry{Extracting default priors, the generated Stan code and Stan data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
pkgdown:
  as_is: true
---
<style type="text/css">
div.main-container {
max-width: 800px !important;
}

p {
margin-top: 1.5em ;
margin-bottom: 1.5em ;
}
.author{
    display: none;
}
</style>

```{r, include = FALSE}
options(crayon.enabled = TRUE, width = 800)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "jpeg",
  dpi = 100
)
fansi::set_knit_hooks(knitr::knit_hooks, which = c("output","message","error"))
```

# Default priors for models in the `bmm` package

Each model in `bmm` comes with default priors on all of its parameters. Unlike
in the `brms` package, the default priors in `bmm` are informative, based on
current expert knowledge in the domain of the model. These default priors help
with model identifiability and improve estimation. However, because the priors
are informed, it is even more important for you to understand what priors are
used when you estimate a model, or when you report the results of a model fit.

You can use the function `default_prior()` from the `brms` package to extract
the default priors for a model. The arguments to `default_prior` are the same as
for the `bmm()` function in `bmm`. For example, if you want to extract the
default priors for the SDM model ([see the online article for more information](https://venpopov.com/bmm/articles/bmm_sdm_simple.html)), where you have
a set_size categorical predictor of `c` and `kappa`, you can use the following
code:


```{r message=FALSE, warning=FALSE}
library(bmm)

default_prior(bmf(c ~ 0 + set_size, kappa ~ 0 + set_size), 
              data = oberauer_lin_2017,
              model = sdm(resp_error = 'dev_rad'))
```

In this case we used a formula of the type `~ 0 + factor`, which means that the
intercept is suppressed, and a separate parameter is estimated for each level of
the set_size factor variable. For the SDM model, both `kappa` and `c` have to be
positive, so they are defined in the model on the log scale, and exponentiated
afterwards. Thus, the parameters are sampled on the log scale, and the priors
are defined on the log scale as well. The default prior for `c` is a student-t
distribution with 5 degrees of freedom, a mean of 2, and a standard deviation of
0.75. This corresponds to the following prior distribution over the log scale,
with 80% of the prior mass between 0.9 and 3.10:

```{r fig.width=4.5, fig.height=4}
log_c <- seq(-2,6, 0.01)
y <- brms::dstudent_t(log_c, df = 5, mu = 2, sigma = 0.75)
plot(log_c, y, type = 'l', xlab = 'log(c)', ylab = 'Density', 
     main = 'Prior distribution for log(c)')
```

This corresponds to the following log-T prior over the native scale of `c`, with
a median of ~7.4, and 80% of the prior mass between 2.44 and 22.35:

```{r fig.width=4.5, fig.height=4}
c <- seq(0, 50, 0.01)
y <- brms::dstudent_t(log(c), df = 5, mu = 2, sigma = 0.75) / c
plot(c, y, type = 'l', xlab = 'c', ylab = 'Density', 
     main = 'Prior distribution for c')
```

The default prior for `kappa` is similar, with a lower mean, a student-t
distribution with 5 degrees of freedom, a mean of 1.75, and a standard deviation
of 0.75, which corresponds to a median of 3.5 on the native scale.

If we had retained the intercept in the formula, the default prior above would
be placed on the intercept, while the effects of each factor level relative to
the intercept would have a default prior of normal(0, 1):

```{r message=FALSE, warning=FALSE}
default_prior(bmf(c ~ 1 + set_size, kappa ~ 1 + set_size), 
              data = oberauer_lin_2017,
              model = sdm(resp_error = 'dev_rad'))
```

You can also see that in both cases, the last line is "constant(0)" on the
Intercept of the `mu` parameter, which is fixed to 0 by default in the model,
and is not estimated. You might wonder why it doesn't say `mu` in the `dpar`
column of that prior - this is because `brms` assumes `mu` is the default
parameter in all models, so it hides it in the output. If you wanted to estimate
`mu`, instead of leaving it fixed, the prior for it would change as well:

```{r message=FALSE, warning=FALSE}
default_prior(bmf(mu ~ 1 + set_size, c ~ 1, kappa ~ 1),
              data = oberauer_lin_2017,
              model = sdm(resp_error = 'dev_rad'))
```

The `mu` parameter uses a `tan_half` link function, which means that the
`student_t(1, 0, 1)` prior results in a uniform prior over the native scale of
`mu` from -pi to pi. You will also notice above that for the regression
coefficients on `mu`, the default prior is an improper flat prior - this is the
only parameter in `bmm` models which has a flat prior by default, and we
strongly recommend you set a prior on it, if you want to calculate Bayes Factors
or use other Bayesian inference methods.

All of the above examples make an important point - priors are always specified
on the scale at which the parameters are sampled. You can always check the
documentation for a given model to see the links for the parameters (e.g.
`?sdm`).

To overwrite the default priors and set your own, you can use the `set_prior`
function from `brms`. For more information, see `?brms::set_prior`.

# Extracting the Stan code

The Stan code used for fitting a model is generated together by `bmm` and
`brms`. `bmm` takes care of the code specific to the model, while `brms`
generates the code for the regression syntax, priors and everything else. If you
want to get the Stan code that would be used for fitting a model, so that you
can inspect it or modify it, you can use the `stancode()` function from `brms`:

```{r comment = "", collapse=FALSE}
stancode(bmf(c ~ 0 + set_size, kappa ~ 0 + set_size), 
         data = oberauer_lin_2017,
         model = sdm(resp_error = 'dev_rad'))
```


Alternatively, if you already have a fitted model object, you can just call
`stancode()` on that object, which will give you the same result:

```r
fit <- bmm(bmf(c ~ 0 + set_size, kappa ~ 0 + set_size), 
              data = oberauer_lin_2017,
              model = sdm(resp_error = 'dev_rad'))
stancode(fit)
```

# Extracting the Stan data

If you want to extract the data that would be used for fitting a model, you can
use the `standata()` function from `brms`. This function will return a list with
the data that would be passed to Stan for fitting the model.

```{r, message=F, warning=F}
sd <- standata(bmf(c ~ 0 + set_size, kappa ~ 0 + set_size), 
               data = oberauer_lin_2017,
               model = sdm(resp_error = 'dev_rad'))
str(sd)
```
