---
title: "Handling RT Contamination in Evidence Accumulation Models"
output: bookdown::html_document2
author:
  - Gidon Frischkorn
bibliography: REFERENCES.bib
header-includes:
  - \usepackage{amsmath}
vignette: >
  %\VignetteIndexEntry{Handling RT Contamination in Evidence Accumulation Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
pkgdown:
  as_is: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)
```

```{r load-packages, message = FALSE}
library(bmm)
library(dplyr)
library(ggplot2)
library(tidyr)
library(rtdists)

theme_set(theme_minimal(base_size = 12))
```

## Introduction

Evidence accumulation models assume that all observed RTs arise from a single cognitive process. In practice, some trials reflect something else entirely: anticipatory responses, attention lapses, or accidental button presses. Even a few such contaminant trials can bias parameter estimates and distort model fit [@ratcliff_EstimatingParametersDiffusion_2002].

The `bmm` package provides two functions for dealing with contamination, one for each type of model it supports:

| Model | Data type | Function | Role |
|-------|-----------|----------|------|
| EZDM | Aggregated stats | `ezdm_summary_stats()` | Required during aggregation |
| DDM / csWald | Trial-level | `flag_contaminant_rts()` | Optional preprocessing |

Both fit a mixture model via the EM algorithm, separating cognitive RTs from a uniform contaminant distribution [@ratcliff_EstimatingParametersDiffusion_2002]:

$$
f(RT) = (1 - \pi) \cdot f_{\text{cognitive}}(RT) + \pi \cdot f_{\text{uniform}}(RT)
$$

where $\pi$ is the contamination proportion. The cognitive component $f_{\text{cognitive}}$ can be ex-Gaussian (default), log-normal, or inverse Gaussian. For background on the EZ diffusion model, see @Wagenmakers_2007 and @wagenmakers_EZDoesIt_2008; for the full diffusion model, see @Ratcliff_McKoon_2008.

## Contamination handling for EZDM

EZDM estimates parameters from pre-computed moments of the reaction times (mean RT, variance) and proportion correct. Once data are aggregated, individual trials are gone, so contamination must be addressed during the aggregation step.

### Method options

```r
my_data |>
  group_by(subject, condition) |>
  reframe(ezdm_summary_stats(rt, response, method = "mixture"))
```

| Method | Description | Use when |
|--------|-------------|----------|
| `"simple"` | Standard mean/variance | Data is pre-cleaned |
| `"robust"` | Median + IQR/MAD-based variance [@wagenmakers_EZDoesIt_2008] | Moderate contamination |
| `"mixture"` | EM mixture modeling [@ratcliff_EstimatingParametersDiffusion_2002] | Default; provides contamination estimates |

### Example: method comparison

As an example, we use data from @Ratcliff_Rouder_1998, available in the `rtdists` package. First we show how the `ezdm_summary_stats()` method options work on a single participant's data from the accuracy instruction condition in this data. The histogram below shows a long tail of slow RTs, which are likely contaminants:

```{r load-rr98-data}
data(rr98)

rr98_subset <- rr98 |>
  filter(id == "jf", instruction == "accuracy") |>
  mutate(response = ifelse(response == "dark", "upper", "lower")) |>
  select(rt, response, strength, correct)

ggplot(rr98_subset, aes(x = rt)) +
  geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7) +
  labs(title = "RT Distribution (Ratcliff & Rouder, 1998)",
       subtitle = "One participant, accuracy condition",
       x = "Response Time (s)", y = "Count")
```

```{r compare-methods}
stats_simple <- ezdm_summary_stats(
  rr98_subset$rt, rr98_subset$response, method = "simple"
)
stats_robust <- ezdm_summary_stats(
  rr98_subset$rt, rr98_subset$response, method = "robust"
)
stats_mixture <- ezdm_summary_stats(
  rr98_subset$rt, rr98_subset$response,
  method = "mixture", contaminant_bound = c("min", "max")
)

bind_rows(
  stats_simple |> mutate(method = "simple"),
  stats_robust |> mutate(method = "robust"),
  stats_mixture |> mutate(method = "mixture")
) |>
  mutate(accuracy = n_upper / n_trials) |>
  select(method, mean_rt, var_rt, accuracy, contaminant_prop)
```

The mixture method also returns a contamination proportion estimate. The robust method is faster and makes no distributional assumptions, but does not quantify contamination.

### 3-parameter vs 4-parameter version

The `version` argument controls whether moments are computed jointly or separately by response boundary:

```{r version-comparison}
stats_3par <- ezdm_summary_stats(
  rr98_subset$rt, rr98_subset$response,
  method = "mixture", version = "3par"
)

stats_4par <- ezdm_summary_stats(
  rr98_subset$rt, rr98_subset$response,
  method = "mixture", version = "4par"
)

cat("3par - overall contamination:", round(stats_3par$contaminant_prop, 3), "\n")
cat("4par - upper:", round(stats_4par$contaminant_prop_upper, 3),
    "  lower:", round(stats_4par$contaminant_prop_lower, 3), "\n")
```

The `"4par"` version is useful when contamination rates might differ between response boundaries and users are interested in estimating response bias for one response over the other. In most cases, `"3par"` is more stable and sufficient.

### EZDM workflow

In a typical analysis, you would aggregate data for all subjects and conditions with `ezdm_summary_stats(method = "mixture")`, then fit the EZDM to those summary stats. If you want to adjust accuracy for contamination, use `adjust_ezdm_accuracy()` before fitting the model. This accounts for the fact that some contaminant trials are likely random guesses, which can bias accuracy estimates.

```{r ezdm-workflow, eval = FALSE}
# Step 1: Aggregate with contamination handling
ezdm_data <- raw_trial_data |>
  group_by(subject, condition) |>
  reframe(ezdm_summary_stats(rt, response, method = "mixture"))

# Step 2 (optional): Adjust accuracy for contamination
ezdm_data <- ezdm_data |>
  mutate(adjust_ezdm_accuracy(n_upper, n_trials, contaminant_prop))

# Step 3: Fit EZDM model
fit <- bmm(
  formula = bmf(drift ~ condition, bound ~ 1, ndt ~ 1),
  data = ezdm_data,
  model = ezdm(
    mean_rt = "mean_rt", var_rt = "var_rt",
    n_upper = "n_upper", n_trials = "n_trials"
  )
)
```

---

## Contamination handling for DDM/csWald

DDM and csWald work with individual trials, so contamination handling is optional. `flag_contaminant_rts()` implements the same mixture modelling approach as `ezdm_summary_stats(method = "mixture")` and returns a contamination probability for each trial; You can use these probabilities in various ways, such as filtering, probabilistic flagging, or weighting trials.

### Basic usage

```{r flag-basic-example}
flagged <- rr98_subset |>
  mutate(contam_prob = flag_contaminant_rts(rt))

head(flagged)
```

Diagnostics (convergence, parameters, log-likelihood) are available via `attr(result, "diagnostics")`:

```{r diagnostics-example}
probs <- flag_contaminant_rts(rr98_subset$rt)
attr(probs, "diagnostics")
```

### Grouped fitting

For obtaining the probability of contamination by subject, condition, or response, use `dplyr::group_by()`:

```{r grouping-example}
rr98_grouped <- rr98 |>
  filter(id %in% c("jf", "kr")) |>
  mutate(response = ifelse(response == "dark", "upper", "lower"))

flagged_grouped <- rr98_grouped |>
  group_by(id, instruction) |>
  mutate(contam_prob = flag_contaminant_rts(rt))

head(flagged_grouped)
```

### Filtering strategies

The simplest approach is a hard threshold: remove trials where $P(\text{contaminant} \mid RT) > 0.5$.

```r
clean_data <- my_data |>
  mutate(contam_prob = flag_contaminant_rts(rt)) |>
  filter(contam_prob <= 0.5)
```

Alternatively, you can probabilistically sample removal decisions proportional to each trial's contamination probability from a binomial distribution, which avoids the sharp cutoff:

```r
set.seed(123)
clean_data <- my_data |>
  mutate(contam_prob = flag_contaminant_rts(rt)) |>
  filter(rbinom(n(), 1, contam_prob) == 0)
```

### DDM/csWald workflow

In a full analysis, you would flag contaminated trials first, then either filter or weight them before fitting the DDM or csWald. The example below shows a simple filtering workflow:

```{r filtering-workflow, eval = FALSE}
# Step 1: Flag contaminated trials
flagged_data <- raw_data |>
  group_by(subject, condition) |>
  mutate(contam_prob = flag_contaminant_rts(rt)) |>
  ungroup()

# Step 2: Filter
clean_data <- flagged_data |> filter(contam_prob < 0.5)

# Step 3: Fit model
fit <- bmm(
  formula = bmf(drift ~ condition, bound ~ 1, ndt ~ 1),
  data = clean_data,
  model = ddm(resp1 = "rt", resp2 = "response")
)
```

### Validating fast guesses

In 2AFC tasks, if fast contaminants are truly random guesses, their accuracy should be around 50%. The helper function `validate_fast_guesses()` checks this with a Bayesian Beta-Binomial test (Savage-Dickey Bayes Factor):

```{r validate-fast-guesses, eval = FALSE}
my_data$is_contaminant <- flag_contaminant_rts(my_data$rt) > 0.5

validation <- validate_fast_guesses(
  contam_flag = my_data$is_contaminant,
  rt_data = my_data$rt,
  response = my_data$response
)

validation$bf_01        # BF for guessing hypothesis
validation$bf_evidence  # Jeffreys scale category
validation$guess_in_hdi # Is 0.5 in 95% HDI?
```

This function can also be used with grouped data to check for differences in fast guess rates across conditions or subjects.

## Case study: multi-subject analysis

The full @Ratcliff_Rouder_1998 dataset has three participants in a brightness discrimination task under speed and accuracy instructions. We walk through both workflows on these data.

### Data preparation

```{r prepare-case-study}
data(rr98)

case_study_data <- rr98 |>
  mutate(response = ifelse(response == "dark", "upper", "lower")) |>
  select(id, instruction, rt, response, correct, strength)

cat("Trials:", nrow(case_study_data),
    " Subjects:", n_distinct(case_study_data$id),
    " Instructions:", paste(unique(case_study_data$instruction), collapse = ", "), "\n")
```

```{r explore-case-study}
ggplot(case_study_data, aes(x = rt, fill = instruction)) +
  geom_histogram(bins = 50, alpha = 0.6, position = "identity") +
  scale_fill_manual(values = c("speed" = "coral", "accuracy" = "steelblue")) +
  labs(title = "RT Distributions by Instruction Condition",
       x = "Response Time (s)", y = "Count") +
  xlim(0, 3)
```

### Workflow A: EZDM

```{r ezdm-workflow-aggregate}
ezdm_data <- case_study_data |>
  group_by(id, instruction) |>
  reframe(ezdm_summary_stats(
    rt, response,
    method = "mixture", version = "4par",
    contaminant_bound = c("min", "max")
  ))

ezdm_data |>
  select(id, instruction, contaminant_prop_upper, contaminant_prop_lower)
```

```{r ezdm-compare-stats}
ezdm_simple <- case_study_data |>
  group_by(id, instruction) |>
  reframe(ezdm_summary_stats(rt, response, method = "simple"))

ezdm_mixture <- case_study_data |>
  group_by(id, instruction) |>
  reframe(ezdm_summary_stats(
    rt, response, method = "mixture",
    contaminant_bound = c("min", "max")
  ))

bind_rows(
  ezdm_simple |> mutate(method = "simple"),
  ezdm_mixture |> mutate(method = "mixture")
) |>
  group_by(instruction, method) |>
  summarize(grand_mean_rt = mean(mean_rt), grand_var_rt = mean(var_rt),
            .groups = "drop")
```

### Workflow B: DDM/csWald

```{r ddm-flag-trials}
flagged_data <- case_study_data |>
  group_by(id, instruction, response) |>
  mutate(contam_prob = flag_contaminant_rts(
    rt, contaminant_bound = c("min", "max")
  )) |>
  ungroup()
```

```{r ddm-viz-probabilities}
ggplot(flagged_data, aes(x = rt, y = contam_prob, color = instruction)) +
  geom_point(alpha = 0.1) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "red") +
  scale_color_manual(values = c("speed" = "coral", "accuracy" = "steelblue")) +
  xlim(0, 3) +
  labs(title = "Contamination Probability vs. RT",
       x = "Response Time (s)", y = "P(Contaminated)")
```

```{r ddm-filter}
clean_data <- flagged_data |> filter(contam_prob < 0.5)

case_study_data |>
  group_by(instruction) |>
  summarize(original_n = n(), .groups = "drop") |>
  left_join(
    clean_data |>
      group_by(instruction) |>
      summarize(filtered_n = n(), .groups = "drop"),
    by = "instruction"
  ) |>
  mutate(removed_pct = round(100 * (original_n - filtered_n) / original_n, 1))
```

Both workflows pick up on contamination differences between instruction conditions.

## Practical guidance

Group by factors that affect the RT distribution (subject, condition), but keep at least 50 to 100 trials per group so the mixture model has enough data to work with.

For contaminant bounds, 100--150ms is a sensible lower bound (faster than any real decision). The upper bound depends on your task; `c("min", "max")` adapts to the data and works well as a default.

The ex-Gaussian distribution is a safe starting point. Try log-normal if you have heavy tails, or inverse Gaussian if you want a closer match to the Wiener process. See `?ezdm_summary_stats` for details.

Check convergence with `attr(result, "diagnostics")$converged`. When the EM fails to converge, it usually means too few trials or bounds that do not fit the data well. `ezdm_summary_stats()` falls back to simple moments in that case.

If estimated contamination rates exceed 20%, first check that RTs are in seconds, not milliseconds. Then verify your bounds. If the rates still seem high, the data may genuinely be noisy.

A 0.5 filtering threshold means you remove trials that are more likely contaminant than not. Raise it to 0.7 to keep more borderline trials, or lower it to 0.3 to filter more aggressively.

## References

