---
title: "Signal Detection Theory models"
output: bookdown::html_document2
author:
  - Gidon Frischkorn
  - Ven Popov
bibliography: REFERENCES.bib
header-includes:
  - \usepackage{amsmath}
vignette: >
  %\VignetteIndexEntry{Signal Detection Theory models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
pkgdown:
  as_is: true
---
<style type="text/css">
div.main-container {
max-width: 800px !important;
}

p {
margin-top: 1.5em ;
margin-bottom: 1.5em ;
}
.author{
    display: none;
}
</style>

```{r, include = FALSE}
options(crayon.enabled = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  dev = "jpeg",
  dpi = 100,
  fig.width = 5,
  fig.asp = 0.8,
  out.width = "80%",
  fig.align = "center"
)
fansi::set_knit_hooks(knitr::knit_hooks, which = c("output", "message", "error"))
```

```{r setup, message = FALSE}
library(bmm)
library(ggplot2)
```

# Signal detection theory in bmm

Signal detection theory (SDT) separates an observer's *sensitivity* — how well
two classes of stimuli (signal vs. noise, old vs. new) can be told apart — from
their *response bias* — where they place the decision criterion
[@Green_Swets_1966; @DeCarlo_1998]. `bmm` ships four SDT measurement models, one
per response format. Each is a self-contained model with its own data
requirements; they are not modes of a single function.

| Constructor | Response format | Key parameters | ROC defined? |
|-------------|-----------------|----------------|--------------|
| `sdt_yn()` | counts of "signal"/"old" responses | `d`, `criterion`, `sdratio` | yes |
| `sdt_rating()` | counts per confidence-rating category | `d`, `criterion`, threshold params, `sdratio` | yes |
| `sdt_mafc()` | counts of correct m-AFC responses | `d` | no |
| `sdt_ranking()` | counts per rank position | `d`, `sdratio` (`dist = "normal"` only) | no |

All four assume a latent evidence variable whose noise distribution is selected
with the `dist` argument: `"normal"` (Gaussian SDT), `"logistic"`,
`"gumbel_min"` (extreme-value SDT), or `"gumbel_max"`.

Two of the models — `sdt_yn()` and `sdt_rating()` — have an explicit
response criterion, so a **receiver operating characteristic (ROC)** curve and
its **area under the curve (AUC)** are defined. The m-AFC and ranking models
have no criterion: accuracy already integrates over the decision rule, so
`roc_sdt()` and `auc_sdt()` deliberately error for them. The latent
decision-variable distributions (`latent_sdt()`) and posterior predictive checks
(`pp_check()`) are available for all four — for the criterion-free models the
latent plot simply shows the noise and signal densities without a boundary line.

This article walks through fitting each model and then through the
post-processing toolkit: `roc_sdt()`, `roc_observed()`, `auc_sdt()`,
`latent_sdt()`, `sdt_thresholds()`, `sdt_sensitivity()`, the `plot()` methods
(including the quantile-transformed ROC via `scale = "quantile"`), and
`pp_check()`.

## The bmm workflow for SDT models

Fitting any of the four models follows the same recipe as every `bmm` model;
only the ingredients are SDT-specific.

1. **Aggregate the data.** All four models consume *pre-aggregated counts*, not
   trial-level responses: one row per design cell (for example per subject and
   stimulus type), with the counts in the columns the constructor names. If
   your data hold one row per trial, aggregate them first (e.g. with
   `dplyr::count()` or `stats::aggregate()`); the Rating SDT section below
   shows a real reshaping example.
2. **Create the model object.** The constructor maps your column names onto the
   model's response format — for example `sdt_yn(response = "n_old",
   stimulus = "stimulus", n_trials = "n_trials")` — and selects the noise
   distribution via `dist`. Printing the model object lists its parameters and
   data requirements.
3. **Specify the formula and fit.** `bmf()` takes one regression formula per
   model parameter, each with full `brms` syntax: predictors for design effects
   and random effects such as `(1 | id)` for hierarchical estimation.
   Parameters you leave out of the formula are held fixed at their default —
   this is how `sdratio` stays at 0 (equal variance) until you add
   `sdratio ~ 1`. `bmm()` then compiles and samples the model through
   `brms`/Stan, and all `brms` arguments (`cores`, `iter`, `prior`, `file`,
   ...) pass through.
4. **Post-process.** The fitted object is a `brmsfit`, so the general `brms`
   toolkit applies (`summary()`, `hypothesis()`, `brms::fixef()`, ...). The
   SDT-specific extractors add the signal-detection views: ROC curves, AUC,
   latent distributions, and threshold locations.

Every example below follows these four steps.

# SDT in brief

All SDT models share one idea. On each trial the observer obtains a noisy sample
of *evidence* from a latent variable $X$. Noise items and signal items generate
evidence from two distributions that overlap; the observer cannot see which
distribution a sample came from, only its value. For the Gaussian (`"normal"`)
model the two distributions are

$$
X \mid \text{noise} \sim \mathrm{Normal}(0,\, 1),
\qquad
X \mid \text{signal} \sim \mathrm{Normal}(\delta,\, s^2),
$$

where $\delta$ is the mean separation and $s$ is the signal-to-noise SD ratio
(the `sdratio` parameter, fixed to $s = 1$ unless estimated). In a binary task
the observer responds "signal" whenever the evidence exceeds a **criterion**
$k$. The two response probabilities are then tail areas of the two
distributions:

$$
\text{FA} = P(X > k \mid \text{noise}) = 1 - \Phi(k),
\qquad
\text{Hit} = P(X > k \mid \text{signal}) = 1 - \Phi\!\left(\tfrac{k - \delta}{s}\right).
$$

**Sensitivity** is that separation expressed in standard-deviation units — but
the two distributions have *different* SDs as soon as $s \neq 1$, so which one
is the yardstick has to be stated. bmm reports

$$
d_a = \frac{\delta}{\sqrt{(1 + s^2)/2}},
$$

the separation in units of the root-mean-square of the two SDs, which weights
noise and signal equally. This is the `d` parameter in every formula below. When
$s = 1$ it reduces to the familiar $d' = \delta$, so nothing changes for
equal-variance models. Dividing by the noise SD instead gives $d_N$ (the
classical $d'$) and by the signal SD gives $d_S$; `sdt_sensitivity()` returns
all three from a fitted model. $d_a$ is the default because it is the only one
that remains comparable across conditions whose $s$ differs
[@Simpson_Fitter_1973; @Macmillan_Creelman_2005].

```{r sdt-distributions, echo = FALSE, fig.cap = "Latent evidence model. Noise and signal distributions overlap; responses above the criterion are labelled as signal. The shaded right tails are the false-alarm rate (noise) and the hit rate (signal); their separation is the sensitivity d-prime."}
crit <- 0.75
dpr  <- 1.5
x    <- seq(-3.5, 5.5, length.out = 600)
dens <- rbind(
  data.frame(x = x, density = dnorm(x, 0, 1),   dist = "Noise"),
  data.frame(x = x, density = dnorm(x, dpr, 1), dist = "Signal")
)
ggplot(dens, aes(x, density, color = dist, fill = dist)) +
  geom_area(data = subset(dens, x > crit), alpha = 0.3,
            position = "identity", color = NA) +
  geom_line(linewidth = 0.8) +
  geom_vline(xintercept = crit, linetype = "dashed") +
  annotate("segment", x = 0, xend = dpr, y = 0.43, yend = 0.43,
           arrow = arrow(ends = "both", length = unit(0.12, "cm"))) +
  annotate("text", x = dpr / 2 - 0.2, y = 0.465, label = "italic(d)*minute",
           parse = TRUE) +
  annotate("text", x = crit + 0.15, y = 0.55, label = "criterion (k)",
           hjust = 0) +
  scale_color_brewer(palette = "Set1") +
  scale_fill_brewer(palette = "Set1") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  coord_cartesian(ylim = c(0, 0.56)) +
  labs(x = "Latent evidence", y = "Density", color = NULL, fill = NULL) +
  theme_classic() +
  theme(legend.position = "top")
```

Sweeping the criterion from strict to lenient traces every (false-alarm, hit)
pair the observer could produce — the **receiver operating characteristic
(ROC)**. When the two distributions have equal variance ($s = 1$) the ROC is
symmetric about the minor diagonal; when the signal distribution is wider
($s > 1$, as is typical for recognition memory) the ROC becomes asymmetric. This
is why a single (hit, false-alarm) point cannot identify $s$: separating $d'$
from $s$ requires *several* operating points along the curve.

```{r sdt-roc-geometry, echo = FALSE, fig.cap = "ROC geometry. Equal-variance models give a symmetric curve; unequal variance (wider signal distribution) bends it asymmetrically. The dashed line is chance (d' = 0)."}
far <- seq(1e-4, 1 - 1e-4, length.out = 400)
roc <- function(far, d, s) 1 - pnorm((qnorm(1 - far) - d) / s)
curves <- rbind(
  data.frame(FA = far, Hit = roc(far, 1.5, 1.0), Model = "Equal variance (s = 1)"),
  data.frame(FA = far, Hit = roc(far, 1.5, 1.6), Model = "Unequal variance (s = 1.6)")
)
ggplot(curves, aes(FA, Hit, color = Model)) +
  geom_abline(linetype = "dashed", color = "grey60") +
  geom_line(linewidth = 1) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  scale_color_brewer(palette = "Dark2") +
  labs(x = "False-alarm rate", y = "Hit rate", color = NULL) +
  theme_classic() +
  theme(legend.position = "top")
```

The four response formats below reuse this latent model and differ only in the
decision rule mapped onto it. The `dist` argument selects the *shape* of the
latent distributions; Gaussian is the safe default, but it is not the only
option. The next section explains the alternatives and when a non-Gaussian shape
is worth choosing.

# Noise distributions

The Gaussian assumption in the model above is a modelling choice, not part of
SDT itself. Because the model never observes the latent samples — only the two
response probabilities — the shape of the noise distribution is exactly what
fixes the shape of the ROC. `bmm` exposes that choice through the `dist` argument, shared by
all four constructors: `"normal"`, `"logistic"`, `"gumbel_min"`, and
`"gumbel_max"`.

## Why the shape matters

Although the Gaussian model is by far the most common, signal detection theory is
not committed to it. The framework needs only two overlapping evidence
distributions and a criterion; the *shape* of those distributions is a separate,
substantive assumption. The Gaussian became the default for historical and
mathematical reasons — Fechner's and Thurstone's psychophysics, and the pull of
the central limit theorem — and @Green_Swets_1966 adopted it on frankly pragmatic
grounds (tractable mathematics) while conceding that other distributions would
serve just as well [@MeyerGrant_2025].

The choice is not cosmetic, because the shape of the distribution controls the
shape of the ROC. Empirical recognition ROCs are almost always *asymmetric*: read
off the z-axes, the operating points fall on a straight line with a slope below 1
(typically around 0.8), so the curve is pulled toward the upper-left rather than
being symmetric about the minor diagonal [@DeCarlo_1998]. A model has to
manufacture that asymmetry somehow, and there are two routes.

The first keeps the Gaussian distribution and spends a parameter on it: let the
signal distribution be *wider* than the noise distribution ($s > 1$, the `sdratio`
parameter). This is the unequal-variance Gaussian model fitted in the Binary SDT
section below, and its zROC slope is exactly $1/s$.

The second changes the *distribution* instead of adding a parameter. If the latent
evidence follows an extreme-value (Gumbel) distribution rather than a Gaussian, the
ROC comes out asymmetric *intrinsically* — with equal variance and one fewer
parameter. This is the extreme-value signal detection (EVSDT) model
[@MeyerGrant_2025], and it is why `sdt_ranking()` defaults to `"gumbel_min"`.

These two routes are easy to conflate, but ROC asymmetry and the variance ratio are
in fact *logically independent*: a skewed equal-variance distribution can produce an
asymmetric ROC, and unequal variances can produce a symmetric one
[@MeyerGrant_2025]. Nor are the routes equally clean. The unequal-variance Gaussian
buys the asymmetry at a conceptual cost: the extra variance it posits implies that
some study events *lower* an item's familiarity, its likelihood ratio is
non-monotonic (so it can predict below-chance hit rates at extreme criteria), and
the simple $d'$ must give way to $d_a$, which no longer guarantees that a larger
value means a uniformly higher ROC [@MeyerGrant_2025]. The extreme-value route
sidesteps all three.

## Where the distributions come from

The deeper reason to care about the noise distribution is that it encodes an
assumption about *how* an observer turns many noisy sensory or memory signals into a
single decision variable. Two aggregation rules lead to two limiting shapes. If the
evidence for an item is the *sum* or *average* of many small contributions, the
central limit theorem drives the decision variable toward a Gaussian. If instead it
is the *most extreme* contribution — the single strongest match, or the most glaring
mismatch — the extreme-value theorem drives it toward a Gumbel distribution, in the
same way and for the same mathematical reasons [@MeyerGrant_2025; @Robinson_2023].
Pooling gives the bell curve; a max (or min) rule gives the skew.

The parallel is exact: a signal-detection model with extreme-value noise and
a maximum decision rule is *formally equivalent* to the softmax (normalized
exponential) choice rule — that is, to Luce's choice axiom [@Yellott_1977;
@Robinson_2023]. Choosing the noise distribution is therefore also choosing a theory
of choice, and the link is visible in `bmm`: the m-AFC and ranking models decide by
comparing the maximum (or the rank order) of evidence across the `m` alternatives, so
an extreme-value distribution is their natural home — `sdt_ranking()` defaults to
`"gumbel_min"` for exactly this reason.

Whether human evidence is better described as pooled or as maximised is an empirical
question, and the answer can depend on the task. In recognition memory the
extreme-value model is principled and parsimonious, and can be grounded in a
behavioural invariance axiom rather than in distributional convenience
[@MeyerGrant_2025]. In visual working memory, by contrast, the Gaussian (pooling)
account reproduces a *stable* sensitivity across changes in the number of
alternatives, whereas the softmax (max-rule) account does not — favouring pooling in
that domain [@Robinson_2023]. `bmm` keeps the distribution switchable so the
data can answer this question.

Statistically, the choice of distribution is the choice of a *link function*.
Written as a generalized linear model, SDT pairs the probit link with the Gaussian
model, the logit link with the logistic, and complementary-log-log links with the
extreme-value models [@DeCarlo_1998] — the same correspondence `bmm` exposes through
`dist`.

## The four options

| `dist` | Latent shape | Equal-variance ROC | Typical motivation |
|--------|--------------|--------------------|--------------------|
| `"normal"` | symmetric (Gaussian) | symmetric, slope 1 | textbook default; pooling rule; comparability with the SDT literature |
| `"logistic"` | symmetric, heavier tails | very nearly symmetric | logit-link convenience; ≈ normal in criterion tasks |
| `"gumbel_min"` | skewed | asymmetric, slope < 1 | extreme-value / EVSDT; max-rule decisions |
| `"gumbel_max"` | mirror of `"gumbel_min"` | asymmetric, slope > 1 | completeness; opposite asymmetry, disfavoured for recognition |

Only two of these are live theoretical choices. `"normal"` and `"gumbel_min"` are
the SDT instantiations of the two aggregation rules above — pooling and the max
rule — and they are the *only* distributions `sdt_ranking()` offers, the model where
the choice bites hardest. The other two are best read as completeness options.
`"logistic"` is so close to `"normal"` once both are rescaled that the two are
nearly indistinguishable in the criterion-based models (its honest role is the logit
link for users coming from logistic regression, not a separate theory of evidence).
`"gumbel_max"` is the extreme-value family pointed the other way: it predicts the
*opposite* (slope > 1) asymmetry that recognition data rarely show, and the choice
rule it entails has been found empirically inadequate [@MeyerGrant_2025;
@Robinson_2023]. It is kept for the rare opposite-asymmetry case and for
non-recognition tasks, not as a co-equal recognition model.

## How the distribution shapes the ROC

Under the extreme-value model the signal and noise distributions are *skewed*
rather than symmetric (Figure \@ref(fig:noise-dist-latent)). Sweeping a criterion
across that skewed pair traces an asymmetric ROC — the same bend that unequal
variance produces, but built into the distribution.

```{r noise-dist-latent, echo = FALSE, fig.cap = "Latent evidence under the extreme-value (gumbel_min) model. Unlike the symmetric Gaussian distributions in the previous figure, both distributions are skewed (a long left tail). Sweeping the criterion over this skewed pair is what makes the ROC asymmetric, with no unequal-variance parameter."}
gamma <- -digamma(1)            # Euler-Mascheroni constant; centres the densities
d     <- 1.6
crit  <- 1.2
fY    <- function(y) exp(-y - exp(-y))   # standard gumbel_min density (.sdt_dists)
u     <- seq(-4, 5, length.out = 600)
dens  <- rbind(
  data.frame(x = u, density = fY(gamma - u),     dist = "Noise"),
  data.frame(x = u, density = fY(d + gamma - u), dist = "Signal")
)
ggplot(dens, aes(x, density, color = dist, fill = dist)) +
  geom_area(data = subset(dens, x > crit), alpha = 0.3,
            position = "identity", color = NA) +
  geom_line(linewidth = 0.8) +
  geom_vline(xintercept = crit, linetype = "dashed") +
  annotate("segment", x = gamma, xend = d + gamma, y = 0.42, yend = 0.42,
           arrow = arrow(ends = "both", length = unit(0.12, "cm"))) +
  annotate("text", x = gamma + d / 2, y = 0.455, label = "italic(d)*minute",
           parse = TRUE) +
  annotate("text", x = crit + 0.15, y = 0.49, label = "criterion (k)",
           hjust = 0) +
  scale_color_brewer(palette = "Set1") +
  scale_fill_brewer(palette = "Set1") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  coord_cartesian(ylim = c(0, 0.5)) +
  labs(x = "Latent evidence", y = "Density", color = NULL, fill = NULL) +
  theme_classic() +
  theme(legend.position = "top")
```

The consequence for the ROC is direct. An equal-variance Gaussian gives a
symmetric curve; an equal-variance `gumbel_min` gives an asymmetric one that lands
almost on top of an unequal-variance Gaussian (Figure
\@ref(fig:noise-dist-roc)). The same asymmetry, bought with the distribution
rather than with an extra parameter.

```{r noise-dist-roc, echo = FALSE, fig.cap = "Model-implied ROCs at d' = 1.6. The equal-variance Gaussian curve is symmetric; the equal-variance gumbel_min curve is asymmetric and nearly coincides with an unequal-variance Gaussian (s = 1.45). Hit(FA) = F((d' + F^-1(FA)) / s), matching bmm's roc_sdt()."}
far <- seq(0.001, 0.999, length.out = 400)
Fn  <- pnorm;                  Qn <- qnorm
Fg  <- function(x) exp(-exp(-x)); Qg <- function(p) -log(-log(p))
hit <- function(far, d, s, F, Q) F((d + Q(far)) / s)
curves <- rbind(
  data.frame(FA = far, Hit = hit(far, 1.6, 1.00, Fn, Qn), Model = "Normal, equal variance"),
  data.frame(FA = far, Hit = hit(far, 1.6, 1.00, Fg, Qg), Model = "Gumbel-min, equal variance"),
  data.frame(FA = far, Hit = hit(far, 1.6, 1.45, Fn, Qn), Model = "Normal, s = 1.45")
)
ggplot(curves, aes(FA, Hit, color = Model, linetype = Model)) +
  geom_abline(linetype = "dashed", color = "grey60") +
  geom_line(linewidth = 1) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  scale_color_manual(values = c("Normal, equal variance" = "#1b9e77",
                                "Gumbel-min, equal variance" = "#d95f02",
                                "Normal, s = 1.45" = "#7570b3")) +
  scale_linetype_manual(values = c("Normal, equal variance" = "solid",
                                   "Gumbel-min, equal variance" = "solid",
                                   "Normal, s = 1.45" = "22")) +
  labs(x = "False-alarm rate", y = "Hit rate", color = NULL, linetype = NULL) +
  theme_classic() +
  theme(legend.position = "top", legend.direction = "vertical")
```

## Worked example: extreme value vs Gaussian

The cleanest test holds everything fixed *except* the distribution. We fit the
base-rate recognition data introduced in the next section
(`broeder_schuetz_2009_e3`) twice, with the same equal-variance formula, changing
only `dist`. Neither model estimates `sdratio`, so they have an identical number
of parameters — any difference in ROC shape is the distribution's doing.

```{r}
ev_formula <- bmf(d ~ 1 + (1 | id), criterion ~ 0 + condition + (1 | id))

fit_normal <- bmm(
  formula = ev_formula,
  data    = broeder_schuetz_2009_e3,
  model   = sdt_yn(response = "n_old", stimulus = "stimulus",
                   n_trials = "n_trials", dist = "normal"),
  backend = "cmdstanr", cores = 4, refresh = 0, silent = 2,
  file = "assets/bmmfit_sdt_noise_normal_vignette"
)

fit_gumbel <- bmm(
  formula = ev_formula,
  data    = broeder_schuetz_2009_e3,
  model   = sdt_yn(response = "n_old", stimulus = "stimulus",
                   n_trials = "n_trials", dist = "gumbel_min"),
  backend = "cmdstanr", cores = 4, refresh = 0, silent = 2,
  file = "assets/bmmfit_sdt_noise_gumbel_vignette"
)
```

The post-processing helpers (`roc_sdt()`, `roc_observed()`, `auc_sdt()`, detailed
in the Binary SDT section) overlay each model-implied ROC on the observed operating
points.

```{r noise-roc-fit, fig.show = "hold", out.width = "49%", fig.cap = "Equal-variance fits to the base-rate recognition data. Left: Gaussian noise forces a symmetric ROC that misses the operating points. Right: gumbel_min reproduces the asymmetric curve through the same points, with no sdratio parameter."}
plot(roc_sdt(fit_normal), observed = roc_observed(fit_normal))
plot(roc_sdt(fit_gumbel), observed = roc_observed(fit_gumbel))
```

The equal-variance Gaussian is locked to a symmetric curve and systematically
misses the operating points; the equal-variance `gumbel_min` bends to follow them.
Its implied zROC slope (~0.65) is close to the unequal-variance Gaussian slope of
0.69 estimated for these data in the Binary SDT section — the extreme-value model
reaches the same place that the Gaussian reaches only by widening the signal
distribution. The two AUCs are similar (overall sensitivity is recovered either
way); the distribution shows up in the *shape* of the curve, not its area.

```{r}
attr(auc_sdt(fit_normal), "summary")
attr(auc_sdt(fit_gumbel), "summary")
```

## Choosing a noise distribution in practice

`"normal"` is the safe default and keeps results comparable with the bulk of the
SDT literature. Reach for `"gumbel_min"` when the ROC is asymmetric or when theory
motivates a max-rule process; it is the natural choice for ranking and forced-choice
tasks, and an equal-variance extreme-value model is often a more parsimonious
account of recognition asymmetry than an unequal-variance Gaussian. `"logistic"` is
mostly a numerical convenience, and `"gumbel_max"` covers the rare opposite
asymmetry. As always, let the data arbitrate: compare candidates with the
observed-versus-model ROC overlap and AUC shown above, with `pp_check()`, or with
`loo()` for a formal information criterion.

One caution on that comparison: the Gaussian model is the more *flexible* of the
two — flexible enough to mimic extreme-value data closely, to the point that in
simulations it wins on a clear majority of datasets that an extreme-value model
actually generated [@MeyerGrant_2025]. A better Gaussian fit is therefore weak
evidence on its own. The distributions are best told apart not by raw fit but by
parsimony, by whether their parameters stay invariant when the design changes (for
example across set sizes or numbers of alternatives) [@Robinson_2023], and by
whether the implied decision process is plausible for the task at hand.

# Binary SDT

In a binary old/new recognition task each cell of the design contributes a count
of "old" responses out of a number of trials, separately for old (`signal`) and
new (`noise`) items. The latent decision variable places "old" responses above a
criterion; sensitivity `d` is the distance between the signal and noise
distributions and `criterion` is the location of the boundary.

## A single condition: equal-variance SDT

We simulate one condition with `rsdt_yn()` and fit an equal-variance model.

```{r}
set.seed(123)
dat <- expand.grid(id = 1:30, stimulus = c(0L, 1L))
dat$n_trials <- 100L
dat$n_old <- rsdt_yn(nrow(dat), dat$n_trials, dat$stimulus,
                     d = 1.5, criterion = 0.2)

model <- sdt_yn(response = "n_old", stimulus = "stimulus",
                n_trials = "n_trials")

fit_ev <- bmm(
  formula = bmf(d ~ 1 + (1 | id), criterion ~ 1 + (1 | id)),
  data = dat,
  model = model,
  backend = "cmdstanr",
  cores = 4,
  refresh = 0,
  silent = 2,
  file = "assets/bmmfit_sdt_yn_ev_vignette"
)
summary(fit_ev)
```

The population estimates recover the simulated `d = 1.5` and
`criterion = 0.2`.

By default `sdratio` is fixed to 0, i.e. the signal and noise distributions have
equal variance (`exp(0) = 1`) — it appears in the model's parameter list but got
no formula, so it stays at its fixed default (step 3 of the workflow recipe).
With a single (hit, false-alarm) pair per subject the signal/noise variance
ratio is not identified, so leaving it fixed is the right default here.

## Multiple criteria: identifying the unequal-variance ratio

When the criterion is shifted across conditions — for example by a base-rate or
payoff manipulation — while sensitivity is held constant, the several
(false-alarm, hit) points trace out a *curve* rather than a single point. That
curvature is what identifies the unequal-variance ratio `sdratio` — the
signal-to-noise SD ratio $s$, whose reciprocal is the slope of the zROC
[@DeCarlo_1998]. The bundled `broeder_schuetz_2009_e3` data
[@Broeder_Schuetz_2009] has exactly this structure: 40 subjects tested under five
base-rate conditions (`br1` conservative … `br5` liberal).

```{r}
model <- sdt_yn(response = "n_old", stimulus = "stimulus",
                n_trials = "n_trials")

fit_uv <- bmm(
  formula = bmf(
    d    ~ 1 + (1 | id),
    criterion ~ 0 + condition + (1 | id),
    sdratio   ~ 1
  ),
  data = broeder_schuetz_2009_e3,
  model = model,
  backend = "cmdstanr",
  cores = 4,
  refresh = 0,
  silent = 2,
  file = "assets/bmmfit_sdt_yn_uv_vignette"
)
summary(fit_uv)
```

Here `d` and `sdratio` are shared across conditions while `criterion` varies
by `condition` — the five criteria are five operating points on a single ROC.
`sdratio` is estimated on the log scale, so exponentiate it for the SD ratio on
the natural scale:

```{r}
exp(brms::fixef(fit_uv)["sdratio_Intercept", c("Estimate", "Q2.5", "Q97.5")])
```

A value above 1 means the signal distribution is wider than the noise
distribution, the classic recognition-memory pattern.

### Which sensitivity index?

Once `sdratio` is free, "the" distance between the two distributions depends on
which standard deviation you measure it in. bmm estimates $d_a$, the separation
divided by the root-mean-square of the two SDs. The alternatives are $d_N$
(divide by the noise SD — the classical $d'$) and $d_S$ (divide by the signal
SD). `sdt_sensitivity()` returns any of them, converting draw by draw so the
intervals carry the joint uncertainty in `d` and `sdratio`:

```{r}
sdt_sensitivity(fit_uv)
```

Under equal variance all three are the same number, which is why the
distinction only appears once `sdratio` is estimated. $d_a$ is the default
because it is the only one of the three that stays comparable across conditions
that differ in `sdratio`: two conditions that are exactly equally discriminable
can show a large, confidently estimated $d_N$ difference purely because their
variance ratios differ. Report $d_N$ or $d_S$ when you need to line up with a
literature that uses them, but prefer $d_a$ for contrasts.

## ROC and AUC

`roc_sdt()` returns the model-implied ROC from the posterior. For a binary fit it
traces the smooth analytical curve from the posterior of `d` (and `sdratio`
when estimated). When it detects a predictor that varies the criterion only — as
`condition` does above — it additionally returns the model-implied operating
points (one per criterion level) in a `points` attribute. `roc_observed()` gives
the empirical operating points straight from the response counts, and the
`plot()` method overlays everything.

```{r}
roc <- roc_sdt(fit_uv)
obs <- roc_observed(fit_uv)

plot(roc, observed = obs)
```

The smooth ribbon is the posterior ROC, the filled points with crosshairs are the
model-implied operating points per base-rate condition, and the open circles are
the observed points. A curvilinear ROC that is asymmetric about the minor
diagonal is the signature of unequal variance (`sdratio > 1`).

`auc_sdt()` returns the posterior area under the curve. For *equal-variance*
Gaussian and Gumbel-min fits it uses the closed forms $\Phi(d'/\sqrt{2})$ and
$\mathrm{logistic}(g')$ respectively; every other case — unequal variance, the
logistic distribution, and all rating fits — is integrated numerically. The
returned AUC is always the area under the full curve (one value per curve, not
the area spanned by the discrete points).

```{r}
auc <- auc_sdt(fit_uv)
plot(auc)
attr(auc, "summary")
```

## Quantile-transformed ROC and latent distributions

Passing `scale = "quantile"` to the ROC `plot()` method reads the rates on the
noise distribution's quantile axis — the inverse CDF of the fitted noise
distribution applied to each rate ($\Phi^{-1}(\text{rate})$ for the Gaussian
model). The axis label names the transform: `z` for the normal (the classic
z-ROC, and `scale = "z"` is accepted as an alias), `logit` for the logistic, and
`loglog`/`cloglog` for the Gumbel distributions. On this axis the symmetric
(normal/logistic) model ROC is a straight line whose slope is
$1/\exp(\mathtt{sdratio})$ and whose intercept is
$\delta/\exp(\mathtt{sdratio})$, where $\delta$ is the separation in noise-SD
units (the $d_N$ that `sdt_sensitivity()` reports, not the fitted `d`), so
a slope below 1 is the visual signature of unequal variance and observed points
that bow away from a straight line flag misfit. (The Gumbel ROC is curved on this
axis; its natural linearising transform is the log-log power-ROC. The $(0,0)$ and
$(1,1)$ endpoints map to infinity and are dropped.)

```{r}
plot(roc, observed = obs, scale = "quantile")
```

`latent_sdt()` reconstructs the latent decision-variable distributions
themselves: the noise and signal evidence densities, separated by `d` and
scaled by `exp(sdratio)`, with the criterion drawn as a dashed line carrying a
credible band on its location. This is the picture the criterion and `d`
estimates describe. Here the criterion varies across the base-rate conditions but
`d`/`sdratio` do not, so the densities are identical across conditions:
`latent_sdt()` collapses them into one panel and overlays the several criteria,
colour-coded by condition (pass `collapse = FALSE` for one panel per condition).

```{r}
plot(latent_sdt(fit_uv))
```

## Posterior predictive checks

`sdt_yn()` uses a custom binomial family, so `pp_check()` delegates to
`brms`. The "old"-response counts span a wide range whose ceiling differs across
cells (the base-rate conditions have different numbers of trials), so the bar
view is cluttered and hard to read. An overlaid empirical CDF split by stimulus
type is a cleaner check that the model reproduces the response distributions for
signal and noise items — and the ROC plot above already checks the operating
points directly:

```{r}
pp_check(fit_uv, type = "ecdf_overlay", group = "stimulus", ndraws = 100)
```

# Rating SDT

In a confidence-rating task each item is judged on an ordered scale (e.g. 1 =
"sure new" … 6 = "sure old"). The data are counts per rating category, one column
per category. Internally `bmm` fits this with `brms`' native multinomial family;
the K-1 confidence criteria are parameterized by a `threshold_type`
[@Selker_2019; @Paulewicz_Blaut_2020].

We use a real recognition ROC dataset. The `roc6` object in the
[MPTinR](https://cran.r-project.org/package=MPTinR) package collects 12 published
old/new recognition experiments as per-subject counts of old and new items across
a 6-point confidence scale. Here we take @Pratte_2010 (97 subjects, 240 trials per
stimulus class — narrow posteriors, clear effects) and reshape it to the format
`sdt_rating()` expects: one row per (subject × stimulus), with categories
`r1` = "sure new" … `r6` = "sure old".

```{r}
stopifnot(requireNamespace("MPTinR", quietly = TRUE))
data("roc6", package = "MPTinR")
pratte <- roc6[roc6$exp == "Pratte_2010", ]

rating_data <- rbind(
  data.frame(id = pratte$id, stimulus = 1L,
             r1 = pratte$OLD_3new, r2 = pratte$OLD_2new, r3 = pratte$OLD_1new,
             r4 = pratte$OLD_1old, r5 = pratte$OLD_2old, r6 = pratte$OLD_3old),
  data.frame(id = pratte$id, stimulus = 0L,
             r1 = pratte$NEW_3new, r2 = pratte$NEW_2new, r3 = pratte$NEW_1new,
             r4 = pratte$NEW_1old, r5 = pratte$NEW_2old, r6 = pratte$NEW_3old)
)

model <- sdt_rating(response = paste0("r", 1:6), stimulus = "stimulus")

# Recognition ROCs are classically asymmetric, so estimate unequal variance
fit_rating <- bmm(
  formula = bmf(d ~ 1 + (1 | id), criterion ~ 1 + (1 | id),
                spacing ~ 1, sdratio ~ 1),
  data = rating_data,
  model = model,
  backend = "cmdstanr",
  cores = 4,
  refresh = 0,
  silent = 2,
  chains = 4,
  iter = 2000,
  file = "assets/bmmfit_sdt_rating_vignette"
)
summary(fit_rating)
```

## Threshold parameterizations

A rating model is binary SDT with several criteria at once: for a $K$-point scale
the $K-1$ thresholds carve the latent evidence axis into $K$ ordered response
bins. The probability of each rating is the area of the corresponding bin under
the signal or noise distribution.

```{r sdt-rating-thresholds, echo = FALSE, fig.cap = "A 6-point rating scale places K-1 = 5 thresholds on the evidence axis, splitting it into six ordered confidence bins. threshold_type governs how these thresholds are spaced."}
thr   <- c(-1.5, -0.6, 0.1, 0.9, 1.9)
dpr   <- 1.4
edges <- c(-3.2, thr, 3.2)
mids  <- (head(edges, -1) + tail(edges, -1)) / 2
x     <- seq(-3.5, 5.5, length.out = 600)
dens <- rbind(
  data.frame(x = x, density = dnorm(x, 0, 1),   dist = "Noise"),
  data.frame(x = x, density = dnorm(x, dpr, 1), dist = "Signal")
)
ggplot(dens, aes(x, density, color = dist)) +
  geom_line(linewidth = 0.8) +
  geom_vline(xintercept = thr, linetype = "dashed", color = "grey50") +
  annotate("text", x = mids, y = 0.45, label = paste0("r", 1:6), size = 3) +
  scale_color_brewer(palette = "Set1") +
  labs(x = "Latent evidence", y = "Density", color = NULL) +
  theme_classic() +
  theme(legend.position = "top")
```

The `threshold_type` argument controls how those $K-1$ criteria are placed and
therefore how many parameters they cost:

- `"parsimonious"` (default) and `"equidistant"`: two parameters
  (`criterion` + `spacing`) — the thresholds are evenly spaced around the
  criterion, differing only in their canonical anchor positions [@Selker_2019].
- `"log_distance"` and `"log_ratio"`: K-2 `delta` parameters that allow uneven
  spacing while guaranteeing the thresholds stay ordered [@Paulewicz_Blaut_2020].
- `"softmax"`: a shared `spacing` plus K-3 allocation `delta` parameters.

Add `sdratio ~ 1` to the formula to estimate unequal variance.

## ROC, AUC, and observed ROC

For a rating model the K-1 thresholds define K-1 ROC points per posterior draw.
`roc_sdt()` returns those points together with the smooth model-implied ROC curve
traced from the posterior of `d` (and `sdratio`); `plot()` shows that curve
as a credible-band ribbon with the K-1 thresholds overlaid as colour-coded
operating points (`c1`..`c(K-1)`) carrying crosshair credible intervals — the same
display as the binary ROC. `roc_observed()` gives the empirical points from the
pooled counts. The quantile axis (`scale = "quantile"`) works here too.

```{r}
roc <- roc_sdt(fit_rating)
plot(roc, observed = roc_observed(fit_rating))

attr(auc_sdt(fit_rating), "summary")
```

## Posterior predictive checks

Because the rating model uses a multinomial family, `pp_check()` produces a
response-proportion profile (observed bars vs. posterior-predictive
point-ranges). For SDT the signal and noise rating distributions should be
checked separately, so `pp_check()` **defaults to faceting by `stimulus`**:

```{r}
pp_check(fit_rating, ndraws = 100)
```

Pass `group = NA` to pool both stimulus types into a single profile, or
`group = "<predictor>"` to facet by another design variable.

The predicted point-ranges track the observed bars closely at the extreme
categories but **deviate systematically for the interior categories** (r3–r5):
the misses are smooth and one-sided rather than random scatter. That pattern is
the fingerprint of *threshold misfit*. The default `"parsimonious"`
parameterization buys its economy (two threshold parameters) by fixing the K-1
criteria at regularly spaced canonical positions, and when the true confidence
thresholds are unevenly spaced — as they usually are in recognition data — those
two parameters cannot reproduce the full rating profile.

## Improving fit with flexible thresholds

The `"log_ratio"` parameterization frees the spacing with K-2 `delta` parameters
(for a 6-point scale: `delta1`, `delta2`, `delta4`, `delta5`; `delta3` is the
reference midpoint), so the thresholds can sit wherever the data require. Refit
with it and re-run the same check:

```{r}
model_lr <- sdt_rating(response = paste0("r", 1:6), stimulus = "stimulus",
                       threshold_type = "log_ratio")

fit_rating_lr <- bmm(
  formula = bmf(d ~ 1 + (1 | id), criterion ~ 1 + (1 | id),
                delta1 ~ 1, delta2 ~ 1, delta4 ~ 1, delta5 ~ 1,
                sdratio ~ 1),
  data = rating_data,
  model = model_lr,
  backend = "cmdstanr",
  cores = 4,
  chains = 4,
  iter = 2000,
  refresh = 0,
  silent = 2,
  file = "assets/bmmfit_sdt_rating_logratio_vignette"
)

pp_check(fit_rating_lr, ndraws = 100)
```

The interior categories are now captured: the flexible thresholds absorb the
systematic deviation that the parsimonious model could not. The improvement is
not free — it costs four threshold parameters (the `delta`s) instead of two — so
it should be weighed against the gain in fit.

For these models the posterior predictive check is the most practical guide to
that trade-off: a parameterization that leaves systematic, one-sided deviations
is too rigid, while one whose predicted point-ranges sit on the observed bars (as
here) is adequate. A *formal* predictive comparison is tempting, but the usual
`loo()` is **not** appropriate here: each observation is an aggregated count over
many trials, so leaving one out removes a large, highly influential chunk of data
and the PSIS-LOO approximation becomes unreliable — it flags most observations
with a high Pareto-$k$, and (because that reflects heavy-tailed importance weights
rather than Monte Carlo noise) more posterior draws will not fix it. The
principled alternative for aggregated data is k-fold cross-validation via
`brms::kfold()`, which refits on held-out folds and avoids importance sampling
altogether. `"log_distance"` behaves like `"log_ratio"`, and `"softmax"` offers an
intermediate option with a shared spacing plus fewer free deltas.

## Extracting the thresholds

The `delta` parameters are only interpretable through the parameterization that
maps them to threshold locations. `sdt_thresholds()` performs that mapping: it
reconstructs the K-1 thresholds per posterior draw on the latent
decision-variable scale, so the confidence criteria can be reported directly,
whatever the `threshold_type`.

```{r}
attr(sdt_thresholds(fit_rating_lr), "summary")
```

`latent_sdt()` shows the same thresholds in context — the noise and signal
evidence densities with the K-1 confidence criteria cutting the axis into the
six rating bins, each threshold carrying the credible band on its location. The
uneven spacing the `log_ratio` fit recovered (and the parsimonious model could
not) is directly visible:

```{r}
plot(latent_sdt(fit_rating_lr))
```

# m-AFC SDT

In an m-alternative forced-choice task the observer picks the target among `m`
alternatives; the data are counts of correct responses. There is no criterion and
no stimulus column — accuracy alone identifies `d` [@DeCarlo_2012]. A correct
response occurs when the target's evidence exceeds that of all $m-1$ distractors;
`d` is the target--distractor separation and stays on the same scale as the `d`
of `sdt_yn()`: for $m = 2$ the Gaussian model has the closed form
$\Phi(d/\sqrt{2})$, and for $m \geq 3$ the Gaussian and logistic models evaluate
the accuracy by quadrature while both Gumbel models stay in closed form. The set size `m` may be a constant or a data column (for mixed set
sizes).

```{r}
# rsdt_mafc() is vectorised over its arguments; varying d by subject
# gives hierarchical data.
set.seed(123)
dat <- data.frame(id = 1:30, n_trials = 100L)
dat$n_correct <- rsdt_mafc(nrow(dat), dat$n_trials, m = 4,
                           d = rnorm(30, mean = 1.5, sd = 0.3))

model <- sdt_mafc(response = "n_correct", n_trials = "n_trials", m = 4)

fit_mafc <- bmm(
  formula = bmf(d ~ 1 + (1 | id)),
  data = dat,
  model = model,
  backend = "cmdstanr",
  cores = 4,
  refresh = 0,
  silent = 2,
  file = "assets/bmmfit_sdt_mafc_vignette"
)
summary(fit_mafc)
```

ROC curves are undefined here (`roc_sdt(fit_mafc)` errors), but `latent_sdt()`
still works: it shows the noise and signal evidence densities the estimated `d`
implies (the noise density stands for each of the $m-1$ distractors), now without
a criterion line because the decision is a max rule over the alternatives.
Setting `show_competitors = TRUE` overlays the density of the *maximum* of the
$m-1$ distractor samples — the effective competitor the target must beat — which
shifts rightward as $m$ grows and shows directly why accuracy falls with set
size.

```{r}
plot(latent_sdt(fit_mafc, show_competitors = TRUE))
```

Each subject contributes one correct-response count, so a bar per count value is
sparse and noisy; an overlaid empirical CDF of the correct-response counts is a
compact check that the model captures the spread of accuracy across subjects:

```{r}
pp_check(fit_mafc, type = "ecdf_overlay", ndraws = 100)
```

# Ranking SDT

In a ranking task the observer orders `m` items by evidence and the data are
counts per rank position assigned to the target [@MeyerGrant_2025]. As with
m-AFC there is no criterion: the rank assigned to the target is an *order
statistic* of the latent evidence, so the probability of each rank follows from
the evidence distribution directly. With `dist = "gumbel_min"` these rank
probabilities have a closed form (the extreme-value assumption that makes ranking
analytically tractable); `dist = "normal"` is also available and evaluated by
quadrature — prefer `gumbel_min` for speed (its closed form is exact), and
`normal` for Gaussian evidence comparable to the other SDT models (the classical
Thurstonian ranking assumption). The model uses the native multinomial family with one category per rank
position, and `m` may again be a constant or a column for mixed set sizes. The
bundled `meyer_grant_jakob_2025` data [@MeyerGrant_Jakob_2025] illustrate the
wide format.

```{r}
model <- sdt_ranking(
  response = c("rank1", "rank2", "rank3", "rank4", "rank5"),
  m = "set_size",
  dist = "gumbel_min"
)

fit_rank <- bmm(
  formula = bmf(d ~ 1 + (1 | id)),
  data = meyer_grant_jakob_2025,
  model = model,
  backend = "cmdstanr",
  cores = 4,
  refresh = 0,
  silent = 2,
  file = "assets/bmmfit_sdt_ranking_vignette"
)
summary(fit_rank)
```

ROC/AUC are undefined; `pp_check()` shows the rank-position profile. With mixed
set sizes, facet by the set-size column so each panel has a homogeneous
denominator:

```{r}
pp_check(fit_rank, group = "set_size", ndraws = 100)
```

# Post-processing reference

| Function | Models | Returns |
|----------|--------|---------|
| `roc_sdt()` | binary, rating | model-implied smooth ROC curve + operating points (criterion levels for multi-criteria binary; confidence thresholds for rating) |
| `roc_observed()` | binary, rating | empirical ROC points from the data |
| `auc_sdt()` | binary, rating | posterior AUC |
| `latent_sdt()` | all four | noise/signal latent densities + criterion/threshold locations (no boundary for m-AFC/ranking) |
| `sdt_thresholds()` | rating | K-1 threshold locations per draw on the latent scale, with posterior summary |
| `sdt_sensitivity()` | all four | sensitivity converted per draw to $d_a$, $d_N$, and $d_S$, with posterior summary |
| `plot()` | `bmm_sdt_roc` (incl. `scale = "quantile"`), `bmm_sdt_auc`, `bmm_sdt_latent` | `ggplot2` ROC / quantile-ROC / AUC / latent-distribution figures |
| `pp_check()` | all four | posterior predictive check |

# References
