---
title: "The Log-Normal Race Model (LNR)"
output: bookdown::html_document2
author:
  - Gidon Frischkorn
bibliography: REFERENCES.bib
header-includes:
  - \usepackage{amsmath}
vignette: >
  %\VignetteIndexEntry{The Log-Normal Race Model (LNR)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
pkgdown:
  as_is: true
---

<style type="text/css">
div.main-container {
max-width: 800px !important;
}

p {
margin-top: 1.5em ;
margin-bottom: 1.5em ;
}
.author{
    display: none;
}
</style>

```{r, include = FALSE}
options(crayon.enabled = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "jpeg",
  dpi = 100,
  fig.asp = 0.8,
  fig.width = 6,
  out.width = "80%",
  fig.align = "center"
)
fansi::set_knit_hooks(knitr::knit_hooks, which = c("output","message","error"))
```

```{r setup, message = FALSE}
library(bmm)
```

# Introduction to the model

The Log-Normal Race model [LNR; @rouderLognormalRaceCognitiveProcess2015] is
a race model for choice response time data. Each response option has its own
accumulator, and these accumulators race to completion. The first accumulator
to finish determines both the chosen response and the observed response time.

The LNR works with two or more response alternatives and has fewer parameters
than the Linear Ballistic Accumulator or the Diffusion Decision Model. This
makes it a good starting point for decomposing RT distributions, especially in
multi-alternative tasks.

## Model specification

Each response alternative $k$ has an accumulator whose finishing time $T_k$
follows a lognormal distribution:

$$
T_k \sim \text{Lognormal}(\mu_k, \sigma)
$$

where $\mu_k$ is the meanlog parameter (lower values produce faster finishing
times) and $\sigma$ is the sdlog parameter controlling variability, called `s`
in `bmm`. The observed response and response time are:

$$
\text{Response} = \arg\min_k T_k, \quad RT = \min_k T_k + \tau
$$

where $\tau$ is the non-decision time for stimulus encoding and motor
execution. Since the lognormal distribution is positively skewed and defined
on the positive reals, it fits the shape of empirical RT distributions well.

# Parametrization in the `bmm` package

The `bmm` package provides two versions of the LNR.

## Simple version (`version = "simple"`)

The simple version uses two meanlog parameters: one for correct responses
and one shared across all error responses.

| Parameter | Description | Link function |
|-----------|-------------|---------------|
| `correct` | Meanlog for the correct accumulator | identity |
| `error`   | Meanlog for all error accumulators | identity |
| `ndt`     | Non-decision time | log |
| `s`       | Sdlog, shared across accumulators | log |

Responses must be integer-coded: 1 for correct, 2 through K for errors. The
total number of alternatives is specified via `n_choices`. Use this
version for standard accuracy-based designs where the distinction between
correct and error processing is the primary interest.

## Custom version (`version = "custom"`)

The custom version assigns a separate meanlog parameter to each response
category. Category names are defined in the formula and matched to character
labels in the response column of the data:

| Parameter | Description | Link function |
|-----------|-------------|---------------|
| one per category | Meanlog for that category's accumulator(s) | identity |
| `ndt`     | Non-decision time | log |
| `s`       | Sdlog, shared across accumulators | log |

The custom version also supports per-category `accumulators` for tasks
where categories contain different numbers of items (e.g., 1 target, 3
similar lures, 5 unrelated distractors). This works analogously to the
`num_options` argument in the M3 model. Category names should not use
reserved internal parameter names such as `mu`, `ndt`, or `s`, should not be
Stan reserved words, should not contain an underscore, and should not end in
numbers. When supplied directly, `accumulators` must contain positive
integers; when supplied via columns, those columns must contain positive
integers on every trial.

## Identifiability

All four parameters are estimated by default, `s` among them. `s` is not a
scale constraint. That argument belongs to models whose accumulation is
measured in arbitrary evidence units — in the Linear Ballistic Accumulator,
multiplying the drift rates and the threshold by the same factor leaves the
likelihood unchanged, so one of them has to be pinned. Nothing of the sort
holds here: the finishing times of the LNR *are* the response times, observed
in seconds, and changing `s` changes both the shape of the predicted RT
distribution and the predicted accuracy.

The data are in fact quite informative about `s`. The table below profiles the
negative log-likelihood of 4000 simulated trials over a grid of `s`,
maximising the two meanlogs and the non-decision time at every grid point:

```{r}
set.seed(7)
prof_dat <- rlnr(n = 4000, m = c(-1, -0.5), s = 0.5, ndt = 0.3)
min_rt <- min(prof_dat$rt)

nll <- function(par, s, dat) {
  ndt <- min_rt * stats::plogis(par[3])
  -sum(dlnr(dat$rt, dat$response, m = par[1:2], s = s, ndt = ndt, log = TRUE))
}

profile_s <- as.data.frame(t(vapply(
  c(0.3, 0.4, 0.5, 0.6, 0.8, 1.0, 1.5),
  function(s) {
    opt <- stats::optim(c(-1, -0.5, 0), nll, s = s, dat = prof_dat)
    c(s = s, nll = opt$value, correct = opt$par[1], error = opt$par[2],
      ndt = min_rt * stats::plogis(opt$par[3]))
  },
  numeric(5)
)))

free <- stats::optim(c(-1, -0.5, 0, log(0.5)),
                     function(par) nll(par[1:3], exp(par[4]), prof_dat))

round(profile_s, 3)
```

The profile bottoms out at the generating value of `r profile_s$s[which.min(profile_s$nll)]`,
and the freely estimated `s` lands at `r round(exp(free$par[4]), 3)`. Holding
`s` at 1 instead is worse by
`r round(profile_s$nll[profile_s$s == 1] - free$value)` log-likelihood units —
this is not a flat direction in the likelihood.

The other parameters do not stay put while `s` is held at the wrong value.
Reading down the table, the non-decision time climbs with `s` until it is
pressed against the fastest observed response time
(`r round(min_rt, 3)` s here, against a generating value of 0.3), and both
meanlogs move with it. That is what makes a wrong `s` expensive: it is paid
for by the parameters one usually wants to interpret.

You can still fix `s` — to reproduce a published fit that held the sdlog at 1,
for instance — by writing `s = 0` in the formula:
`bmf(correct ~ 1, error ~ 1, ndt ~ 1, s = 0)`. Constants in `bmf()` are read on
the parameter's own link scale, and `s` uses a log link, so `s = 0` means
sdlog = 1.

# Data requirements

## Simple version

The data must contain:

- An RT column with response times in **seconds**
- A response column with integer values: 1 = correct, 2:K = errors

```{r}
set.seed(42)
m_true <- c(-1, 0)
s_true <- 0.5
ndt_true <- 0.2

dat_simple <- rlnr(n = 2000, m = m_true, s = s_true, ndt = ndt_true)
str(dat_simple)
```

For a four-choice task, simulate four accumulators. The simple version pools
responses 2 through 4 into the single `error` parameter, so they share a
meanlog here as well:

```{r}
set.seed(43)
dat_k4 <- rlnr(n = 2000, m = c(-1, -0.2, -0.2, -0.2), s = 0.5, ndt = 0.2)
table(dat_k4$response)
```

## Custom version

The data must contain:

- An RT column with response times in **seconds**
- A response column with **character labels** matching the formula parameters

```{r}
set.seed(44)
sim <- rlnr(n = 2000, m = c(-1, -0.5, -0.2), s = 0.5, ndt = 0.15)
dat_custom <- data.frame(
  rt = sim$rt,
  resp = c("recall", "similar", "other")[sim$response]
)
str(dat_custom)
table(dat_custom$resp)
```

# Fitting the model

## Simple version: two-choice task

```{r, message=FALSE, results='hide'}
model <- lnr(rt = "rt", response = "response", n_choices = 2)

formula <- bmf(
  correct ~ 1,
  error ~ 1,
  ndt ~ 1,
  s ~ 1
)

fit <- bmm(formula, dat_simple, model,
           cores = 4, backend = "cmdstanr",
           # Reduce iterations to keep file size manageable
           iter = 1000,
           warmup = 500,
           file = "assets/bmmfit_lnr_simple_vignette.rds")
```

```{r}
summary(fit)
```

`ndt` and `s` use a log link, so they have to be exponentiated before they are
compared with the values the data were simulated from:

```{r}
fixef_est <- brms::fixef(fit)

data.frame(
  parameter = c("correct", "error", "ndt", "s"),
  true = c(m_true, ndt_true, s_true),
  estimate = c(
    fixef_est["correct_Intercept", "Estimate"],
    fixef_est["error_Intercept", "Estimate"],
    exp(fixef_est["ndt_Intercept", "Estimate"]),
    exp(fixef_est["s_Intercept", "Estimate"])
  ),
  lower = c(
    fixef_est["correct_Intercept", "Q2.5"],
    fixef_est["error_Intercept", "Q2.5"],
    exp(fixef_est["ndt_Intercept", "Q2.5"]),
    exp(fixef_est["s_Intercept", "Q2.5"])
  ),
  upper = c(
    fixef_est["correct_Intercept", "Q97.5"],
    fixef_est["error_Intercept", "Q97.5"],
    exp(fixef_est["ndt_Intercept", "Q97.5"]),
    exp(fixef_est["s_Intercept", "Q97.5"])
  )
)
```

## Posterior predictive checks

A race model predicts two things at once, and a model can reproduce the
response time distribution while getting the choice proportions wrong, so both
are worth checking. `pp_check_vars()` lists the observables the LNR declares:

```{r}
pp_check_vars(fit)
```

The default check plots the response times. Passing `resp_var = "response"`
checks the winning category instead, as a bar plot of observed against
predicted response counts:

```{r, message=FALSE}
pp_check(fit)
```

```{r, message=FALSE}
pp_check(fit, resp_var = "response", type = "bars")
```

## Four-choice task

For a four-choice task, set `n_choices = 4`. Responses 2 through 4 all
share the `error` parameter:

```{r, message=FALSE, results='hide'}
model_k4 <- lnr(rt = "rt", response = "response", n_choices = 4)

fit_k4 <- bmm(bmf(correct ~ 1, error ~ 1, ndt ~ 1, s ~ 1),
              dat_k4, model_k4,
              cores = 4, backend = "cmdstanr",
              iter = 1000,
              warmup = 500,
              file = "assets/bmmfit_lnr_k4_vignette.rds")
```

```{r}
summary(fit_k4)
```

## Custom version: named categories

```{r, message=FALSE, results='hide'}
model_custom <- lnr(rt = "rt", response = "resp", version = "custom")

formula_custom <- bmf(
  recall ~ 1,
  similar ~ 1,
  other ~ 1,
  ndt ~ 1,
  s ~ 1
)

fit_custom <- bmm(formula_custom, dat_custom, model_custom,
                  cores = 4, backend = "cmdstanr",
                  iter = 1000,
                  warmup = 500,
                  file = "assets/bmmfit_lnr_custom_vignette.rds")
```

```{r}
summary(fit_custom)
```

## Per-category accumulator counts

When categories have different numbers of racing accumulators, pass a named
integer vector to `accumulators`:

```{r}
model_numalt <- lnr(
  rt = "rt", response = "resp", version = "custom",
  accumulators = c(recall = 1, similar = 3, other = 5)
)
```

If the number of items per category varies across trials, pass column names
instead of integers (e.g.,
`c(recall = "n_recall", similar = "n_similar", other = "n_other")`).

## Hierarchical models

Both versions can be specified with random effects. Earlier versions of this
article warned that hierarchical LNR fits were fragile, and that divergent
transitions appeared once the non-decision time came close to the fastest
observed RT. That was a symptom of the sdlog being held at 1 rather than a
property of the model: with `s` fixed above its true value, the only way the
model can reproduce the observed spread of response times is to push `ndt` up
against the fastest trial, and the sampler then has to work in a corner of the
parameter space that the data bound from above. With `s` estimated, `ndt`
stays where the data put it.

Hierarchical fits remain the more demanding case, simply because there are
more parameters and fewer trials informing each of them. Start with a simple
random-effects structure and slightly stricter sampler settings such as
`adapt_delta = 0.95`, and treat convergence diagnostics as a hard requirement
before interpreting the estimates.

```{r}
set.seed(45)
n_id <- 20
n_trials <- 60

dat_hier <- do.call(rbind, lapply(seq_len(n_id), function(id) {
  u_correct <- stats::rnorm(1, 0, 0.25)
  u_error <- stats::rnorm(1, 0, 0.25)
  u_ndt <- stats::rnorm(1, 0, 0.15)
  do.call(rbind, lapply(c("easy", "hard"), function(condition) {
    sim <- rlnr(
      n = n_trials,
      m = c(-1 + u_correct + 0.4 * (condition == "hard"), u_error),
      s = 0.5,
      ndt = 0.2 * exp(u_ndt)
    )
    data.frame(id = id, condition = condition,
               rt = sim$rt, response = sim$response)
  }))
}))
dat_hier$condition <- factor(dat_hier$condition, levels = c("easy", "hard"))
str(dat_hier)
```

```{r, message=FALSE, results='hide'}
model_hier <- lnr(rt = "rt", response = "response", n_choices = 2)

formula_hier <- bmf(
  correct ~ 1 + condition + (1 + condition | id),
  error ~ 1 + (1 | id),
  ndt ~ 1 + (1 | id),
  s ~ 1
)

fit_hier <- bmm(formula_hier, dat_hier, model_hier,
                cores = 4, backend = "cmdstanr",
                iter = 1000,
                warmup = 500,
                control = list(adapt_delta = 0.95),
                file = "assets/bmmfit_lnr_hierarchical_vignette.rds")
```

```{r}
summary(fit_hier)
```

# Default priors

| Parameter | Prior | Scale |
|-----------|-------|-------|
| `correct` | `normal(-1, 0.5)` | identity (meanlog) |
| `error` | `normal(0, 0.5)` | identity (meanlog) |
| `ndt` | `normal(-1.5, 0.5)` | log (median 0.22s on the natural scale) |
| `s` | `normal(0, 0.5)` | log (median 1, 95% between 0.37 and 2.7) |

For the custom version, category parameters default to `normal(0, 0.5)` on
the identity scale. The `ndt` prior is the one the **ddm** model uses:
multi-alternative choice tasks carry longer non-decision times than the
two-boundary models in `bmm` were first calibrated on, and a prior whose upper
range sits below the true value drags the meanlogs down with it. The prior on
`s` puts its median at a sdlog of 1 while leaving room for the values RT data
are usually fitted at.

Check the priors of a given model and formula with `default_prior()` and
adjust them as needed:

```{r}
default_prior(formula, dat_simple, model)
```

# References
