---
title: "The EZ-Diffusion Model (EZDM)"
output: bookdown::html_document2
author:
  - Gidon Frischkorn
  - Ven Popov
bibliography: REFERENCES.bib
header-includes:
  - \usepackage{amsmath}
vignette: >
  %\VignetteIndexEntry{The EZ-Diffusion Model (EZDM)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
pkgdown:
  as_is: true
---

<style type="text/css">
div.main-container {
max-width: 850px !important;
}

p {
margin-top: 1.5em ;
margin-bottom: 1.5em ;
}
.author{
    display: none;
}
</style>

```{r, include = FALSE}
options(crayon.enabled = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "jpeg",
  dpi = 100,
  fig.asp = 0.8,
  fig.width = 6,
  out.width = "80%",
  fig.align = "center"
)
fansi::set_knit_hooks(knitr::knit_hooks, which = c("output","message","error"))
```

# Introduction to the model

The diffusion model [@Ratcliff_1978] is a widely used model in speeded decision-making. It describes how participants accumulate noisy evidence over time until reaching one of two decision boundaries. Usually, the diffusion model is fit to trial-level reaction time distributions, which can be computationally intensive, especially for hierarchical Bayesian estimation.

The **EZ-Diffusion Model** [@Wagenmakers_2007] provides a simplified alternative building on closed-form equations that relate the parameters of the diffusion model to easily computed summary statistics: mean reaction time (MRT), reaction time variance (VRT), and accuracy (proportion correct, Pc). This makes it possible to calculate diffusion model parameters from aggregated data without fitting to individual trials.

## The Diffusion Model

In the standard diffusion model, evidence accumulates from a starting point toward one of two boundaries. The key parameters are:

- **Drift rate ($v$)**: The average rate of evidence accumulation, reflecting task difficulty or ability
- **Boundary separation ($a$)**: The distance between the two decision boundaries, reflecting the speed-accuracy tradeoff
- **Non-decision time ($T_{er}$)**: Time for processes outside the decision (encoding, motor response)
- **Starting point ($z$)**: A response bias indicating where evidence accumulation begins relative to the boundaries

```{r diffusion-illustration, echo=FALSE, fig.cap="Illustration of the diffusion model. Evidence accumulates from the starting point until reaching the upper (correct) or lower (error) boundary.", out.width="70%"}
# Simple illustration of diffusion process
set.seed(42)
n_paths <- 5
n_steps <- 100
drift <- 0.05
bound <- 1

# Generate sample paths
paths <- matrix(0, n_steps, n_paths)
for (j in 1:n_paths) {
  for (i in 2:n_steps) {
    paths[i, j] <- paths[i-1, j] + drift + rnorm(1, 0, 0.15)
    if (paths[i, j] > bound || paths[i, j] < -bound) {
      # Clamp to boundary and make remaining path invisible (absorbed)
      paths[i, j] <- pmax(pmin(paths[i, j], bound), -bound)
      if (i < n_steps) paths[(i+1):n_steps, j] <- NA
      break
    }
  }
}

# Plot
par(mar = c(4, 4, 2, 1))
plot(1:n_steps, paths[,1], type = "l", ylim = c(-1.2, 1.2),
     xlab = "Time", ylab = "Evidence", col = "steelblue", lwd = 1.5,
     main = "Diffusion Model: Evidence Accumulation")
for (j in 2:n_paths) {
  lines(1:n_steps, paths[,j], col = "steelblue", lwd = 1.5)
}
abline(h = bound, col = "darkgreen", lwd = 2, lty = 2)
abline(h = -bound, col = "darkred", lwd = 2, lty = 2)
abline(h = 0, col = "gray50", lty = 3)
text(5, bound + 0.1, "Upper boundary (correct)", adj = 0, col = "darkgreen")
text(5, -bound - 0.1, "Lower boundary (error)", adj = 0, col = "darkred")
text(n_steps*0.8, 0.1, "Starting point", adj = 0, col = "gray50")

# Add drift rate illustration (arrow showing average direction of accumulation)
# Arrow starts at step 0 and ends at the upper bound
drift_arrow_x_end <- bound / drift
arrows(x0 = 0, y0 = 0,
       x1 = drift_arrow_x_end, y1 = bound,
       col = "darkorange", lwd = 2.5, length = 0.12)
text((drift_arrow_x_end/2) + 25, (bound/2),
     expression(italic(v)~"(drift rate)"), adj = 0, col = "darkorange")
```

## The EZ Equations

@Wagenmakers_2007 provided closed-form solutions relating the diffusion model parameters to observable summary statistics. For the **3-parameter version** (assuming symmetric starting point, $z = a/2$):

Given the proportion correct ($P_c$), mean RT of correct responses ($MRT$), and variance of correct RTs ($VRT$), the parameters can be computed as:

$$
v = \text{sign}(P_c - 0.5) \cdot s \cdot \left[ L(P_c) \cdot \frac{P_c^2 L(P_c) - P_c \cdot L(P_c) + P_c - 0.5}{VRT} \right]^{1/4}
$$

$$
a = s^2 \cdot \frac{L(P_c)}{v}
$$

$$
T_{er} = MRT - \frac{a}{2v} \cdot \frac{1 - e^{-\frac{va}{s^2}}}{1 + e^{-\frac{va}{s^2}}}
$$

where $L(P_c) = \ln\left(\frac{P_c}{1 - P_c}\right)$ is the logit function and $s$ is the diffusion constant (typically fixed to 1 for identification).

## The 4-Parameter Extension

The original EZ model assumes a symmetric starting point. @Srivastava_2016 derived the moments of first-passage times for the Wiener diffusion process with arbitrary starting points, enabling a **4-parameter version** that can estimate starting point bias ($z_r$, the relative starting point). This 4-parameter version, however, requires separate summary statistics for responses to each boundary:

- Mean and variance of RTs for upper boundary responses (typically correct)
- Mean and variance of RTs for lower boundary responses (typically errors)

With sufficient responses at both the upper and lower bound, this allows estimation of response bias, which is important when participants favor one response over another.

## Bayesian Estimation

So far, studies using the EZ-Diffusion model have adopted step-wise analyses approaches. First, diffusion model parameters are calculated for summary statistics for each individual in each design cell. Second, the calculated parameters are submitted to statistical tests evaluating differences between experimental conditions. This approach has two downsides: 1) uncertainty in parameter estimation is not sufficiently propagated between analysis steps, and 2) all parameters of the diffusion model were allowed to vary over all design cells, bearing the risk of over fitting. To address these issues, @Chavez_Vandekerckhove_2025 introduced a Bayesian hierarchical approach to the EZ-diffusion model. This approach combines the derivation of summary statistics from diffusion model parameters with a probabilistic model of the sampling distributions of the summary statistics. Specifically, this approach:

1. Treats the summary statistics as **observed data** generated from a known sampling distribution
2. Uses **likelihood-based inference** to estimate parameters with proper uncertainty quantification
3. Enables **hierarchical modeling** with random effects across participants

@Chavez_Vandekerckhove_2025 shared JAGS code alongside their model. The downside of this implementation is that the model code has to be adapted to different experimental designs which likely leads to errors and increases the bar for using the implementation. The `bmm` package implements this Bayesian approach, allowing researchers to:

- Estimate uncertainty in parameter estimates
- Include random effects for participants
- Compare conditions using Bayesian hypothesis testing
- Leverage the full power of the `brms` modeling framework

# Parametrization in the `bmm` package

The `bmm` package implements two versions of the EZ-diffusion model:

## 3-Parameter Version (`version = "3par"`)

Estimates three parameters with a fixed symmetric starting point ($z_r = 0.5$):

| Parameter | Description | Link Function |
|-----------|-------------|---------------|
| `drift`   | Drift rate ($v$) - rate of evidence accumulation | identity |
| `bound`   | Boundary separation ($a$) - response caution | log |
| `ndt`     | Non-decision time ($T_{er}$) - encoding + motor time | log |

The diffusion constant `s` is fixed to 1 for identification.

## 4-Parameter Version (`version = "4par"`)

Adds estimation of starting point bias:

| Parameter | Description | Link Function |
|-----------|-------------|---------------|
| `drift` | Drift rate ($v$) | identity |
| `bound` | Boundary separation ($a$) | log |
| `ndt` | Non-decision time ($T_{er}$) | log |
| `zr` | Relative starting point (0-1 scale) | logit |

Note that `zr = 0.5` indicates no bias (symmetric starting point), `zr > 0.5` indicates bias toward the upper boundary, and `zr < 0.5` indicates bias toward the lower boundary.

# Data requirements

Unlike other models in `bmm` that work with trial-level data, the EZ-diffusion model requires **aggregated summary statistics**:

| Variable | Description |
|----------|-------------|
| `mean_rt` | Mean reaction time (in **seconds**) |
| `var_rt` | Variance of reaction times |
| `n_upper` | Number of responses hitting the upper boundary |
| `n_trials` | Total number of trials |

For the **3-parameter version**, `mean_rt` and `var_rt` should be computed from all correct responses pooled together.

For the **4-parameter version**, you need separate statistics for upper and lower boundary responses:
- `mean_rt_upper`, `var_rt_upper` for upper boundary (correct) responses
- `mean_rt_lower`, `var_rt_lower` for lower boundary (error) responses

# Preparing data with `ezdm_summary_stats()`

The `bmm` package provides the `ezdm_summary_stats()` function to compute the required aggregated statistics from trial-level RT data in the correct format. RT contaminants (fast guesses, lapses, attentional failures) can severely distort mean and variance estimates, leading to biased parameter estimates [@wagenmakers_EZDoesIt_2008]. The function offers three methods to address this:

| Method | Description | Use when |
|--------|-------------|----------|
| `"simple"` | Standard mean/variance | Data is pre-cleaned |
| `"robust"` | Median + IQR/MAD-based variance | Moderate contamination |
| `"mixture"` | EM mixture modeling (default) | General use; provides contamination estimates |

For detailed coverage of RT contamination handling, including mixture model theory, distribution options, contaminant bounds, and diagnostics, see `vignette("bmm_rt_contamination")`.

## Example with Simulated Data

To illustrate the impact of contamination, we generate trial-level data with known parameters and introduce 5% contaminant trials:

```{r trial-level-simulation, message=FALSE}
library(bmm)
library(rtdists)
library(dplyr)

set.seed(42)

# Generate clean trial-level data with known parameters
n_trials <- 200
true_drift <- 2.5
true_bound <- 1.2
true_ndt <- 0.25

# Simulate diffusion process
trials_clean <- rdiffusion(
  n = n_trials,
  a = true_bound,
  v = true_drift,
  t0 = true_ndt,
  z = 0.5 * true_bound
)

# Add contaminants (5% fast guesses and lapses)
n_contaminants <- round(n_trials * 0.05)
contaminant_idx <- sample(n_trials, n_contaminants)

trials_contaminated <- trials_clean
trials_contaminated$rt[contaminant_idx] <- runif(n_contaminants, 0.1, 3)
trials_contaminated$response[contaminant_idx] <- sample(c("upper", "lower"),
                                                         n_contaminants,
                                                         replace = TRUE)

trials_contaminated <- trials_contaminated |>
  mutate(
    correct = as.numeric(response == "upper"),
    subject = 1,
    condition = "standard"
  )
```

We can now compare the three methods. The function takes `rt` and `response` as vectors. For grouped operations, use `dplyr::group_by()` with `reframe()`:

```{r method-comparison}
summary_simple <- ezdm_summary_stats(
  trials_contaminated$rt, trials_contaminated$correct,
  method = "simple", version = "3par"
)

summary_robust <- ezdm_summary_stats(
  trials_contaminated$rt, trials_contaminated$correct,
  method = "robust", version = "3par"
)

summary_mixture <- ezdm_summary_stats(
  trials_contaminated$rt, trials_contaminated$correct,
  method = "mixture", version = "3par"
)

comparison <- data.frame(
  Method = c("Simple", "Robust", "Mixture", "TRUE"),
  Mean_RT = c(
    summary_simple$mean_rt,
    summary_robust$mean_rt,
    summary_mixture$mean_rt,
    mean(trials_clean$rt[trials_clean$response == "upper"])
  ),
  Var_RT = c(
    summary_simple$var_rt,
    summary_robust$var_rt,
    summary_mixture$var_rt,
    var(trials_clean$rt[trials_clean$response == "upper"])
  ),
  Contaminant = c(NA, NA, summary_mixture$contaminant_prop, 0.05)
)

print(comparison)
```

The mixture method identifies contaminants and provides more accurate moment estimates.

## Accuracy Adjustment

When using the mixture method, the `adjust_ezdm_accuracy()` function can adjust accuracy counts by removing estimated contaminants. This assumes contaminants are random guesses (50% accuracy by default) and is applied as a separate step after computing summary statistics:

```{r accuracy-adjustment}
summary_no_adjust <- ezdm_summary_stats(
  trials_contaminated$rt, trials_contaminated$correct, version = "3par"
)

summary_adjust <- ezdm_summary_stats(
  trials_contaminated$rt, trials_contaminated$correct, version = "3par"
) |>
  mutate(adjust_ezdm_accuracy(n_upper, n_trials, contaminant_prop))

cat("Unadjusted accuracy:",
    round(summary_no_adjust$n_upper / summary_no_adjust$n_trials, 3), "\n")
cat("Adjusted accuracy:",
    round(summary_adjust$n_upper_adj / summary_adjust$n_trials_adj, 3), "\n")
cat("True accuracy (no contaminants):",
    round(mean(trials_clean$response == "upper"), 3), "\n")
```

## Working with the 4-Parameter Version

The 4-parameter version estimates starting point bias (`zr`) and requires separate statistics for upper and lower boundary responses:

```{r fourpar-example}
# Generate data with response bias (zr = 0.6, favoring upper boundary)
trials_biased <- rdiffusion(
  n = 200,
  a = 1.2,
  v = 2.0,
  t0 = 0.25,
  z = 0.7 * 1.2
)

trials_biased <- trials_biased |>
  mutate(
    correct = as.numeric(response == "upper"),
    subject = 1
  )

# Compute summary statistics for 4-parameter model
summary_4par <- ezdm_summary_stats(
  trials_biased$rt, trials_biased$correct, version = "4par"
)

summary_4par
```

Notice the output now includes:
- `mean_rt_upper`, `var_rt_upper`: Statistics for upper boundary responses
- `mean_rt_lower`, `var_rt_lower`: Statistics for lower boundary responses

This format is required for fitting the 4-parameter EZ-DM model.

## Complete Workflow: From Trial-Level to Model Fitting

Here's the complete pipeline from trial-level data to parameter estimation:

```{r complete-workflow, eval=FALSE}
# Step 1: Generate or load trial-level data
trial_data <- rdiffusion(
  n = 100 * 20,  # 20 subjects, 100 trials each
  a = 1.5, v = 2.5, t0 = 0.3, z = 0.75
) |>
  mutate(
    subject = rep(1:20, each = 100),
    correct = as.numeric(response == "upper")
  )

# Step 2: Compute summary statistics with robust estimation
summary_data <- trial_data |>
  group_by(subject) |>
  reframe(ezdm_summary_stats(
    rt, correct,
    method = "mixture", distribution = "exgaussian", version = "3par"
  ))

# Step 3: Inspect contaminant proportions (quality control)
summary(summary_data$contaminant_prop)

# Step 4: Fit model with bmm
model <- ezdm("mean_rt", "var_rt", "n_upper", "n_trials", version = "3par")
formula <- bmf(drift ~ 1, bound ~ 1, ndt ~ 1)

fit <- bmm(
  formula = formula,
  data = summary_data,
  model = model,
  backend = "cmdstanr"
)

# Step 5: Examine results
summary(fit)
```

This workflow ensures robust parameter estimation by properly handling RT contaminants throughout the analysis pipeline.

# Fitting the model with `bmm`

We start by loading the `bmm` package:

```{r setup, message=FALSE, warning=FALSE}
library(bmm)
```

## Generating simulated data

For illustration, we will generate simulated data with known parameters using the `rezdm()` function. As for all other models, `bmm` provides a density and random generation function for all implemented models. In the case of the `ezdm` there is no single dependent variable for the density function, which makes efficient illustration challenging. However, the `rezdm` function allows for evaluating parameter recovery simulations and predicitions based on the `ezdm`.

```{r}
set.seed(123)

# Define true parameters for two conditions
true_params <- data.frame(
  condition = c("easy", "hard"),
  drift = c(3.0, 1.5),
  bound = c(1.5, 1.5),
  ndt = c(0.3, 0.3)
)

# Number of subjects and trials per condition
n_subjects <- 20
n_trials <- 100

# Generate data
sim_data <- data.frame()
for (subj in 1:n_subjects) {
  for (cond in 1:2) {
    # Add some random variation for each subject
    drift_subj <- true_params$drift[cond] + rnorm(1, 0, 0.3)
    bound_subj <- true_params$bound[cond] + rnorm(1, 0, 0.1)
    ndt_subj <- true_params$ndt[cond] + rnorm(1, 0, 0.02)

    # Generate summary statistics
    stats <- rezdm(
      n = 1,
      n_trials = n_trials,
      drift = max(drift_subj, 0.1),
      bound = max(bound_subj, 0.5),
      ndt = max(ndt_subj, 0.1),
      version = "3par"
    )

    stats$subject <- subj
    stats$condition <- true_params$condition[cond]
    sim_data <- rbind(sim_data, stats)
  }
}

# Convert condition to factor
sim_data$condition <- factor(sim_data$condition, levels = c("easy", "hard"))

head(sim_data)
```

## Specifying the formula

Next, we specify the model formula using `bmmformula()` (or `bmf()`). For this example, we want to estimate separate drift rates for each condition while keeping boundary and non-decision time constant:

```{r}
model_formula <- bmf(
  drift ~ 0 + condition,
  bound ~ 1,
  ndt ~ 1
)
```

## Specifying the model

Next, we specify the `bmmodel`object for the EZDM model by calling `ezdm` and passing the respective variable names from our dataset:

```{r}
model <- ezdm(
  mean_rt = "mean_rt",
  var_rt = "var_rt",
  n_upper = "n_upper",
  n_trials = "n_trials",
  version = "3par"
)
```

As for all other `bmmodel` this provides the necessary information for `bmm` to properly set up the model for estimation.

## Fitting the model

Now we can fit the model using `bmm()`:

```{r, eval=FALSE}
fit <- bmm(
  formula = model_formula,
  data = sim_data,
  model = model,
  cores = 4,
  backend = "cmdstanr",
  file = "assets/bmmfit_ezdm_vignette"
)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# For the vignette, we'll fit a quick model or load a cached one
fit <- bmm(
  formula = model_formula,
  data = sim_data,
  model = model,
  cores = 4,
  iter = 1000,
  backend = "cmdstanr",
  file = "assets/bmmfit_ezdm_vignette"
)
```

## Inspecting results

First, check the model summary:

```{r, comment=NA}
summary(fit)
```

To provide stable sampling all parameters use log link functions, so we need to exponentiate to get estimates on the natural scale:

```{r}
# Extract fixed effects
fixef_est <- brms::fixef(fit)

# Transform to natural scale
cat("Drift rate (easy):", exp(fixef_est["drift_conditioneasy", "Estimate"]), "\n")
cat("Drift rate (hard):", exp(fixef_est["drift_conditionhard", "Estimate"]), "\n")
cat("Boundary:", exp(fixef_est["bound_Intercept", "Estimate"]), "\n")
cat("Non-decision time:", exp(fixef_est["ndt_Intercept", "Estimate"]), "\n")
```

As you can see, these match the generating values well:

```{r}
print(true_params)
```

## Visualizing posterior distributions

```{r, message=FALSE, warning=FALSE, fig.width=7, fig.height=5}
library(tidybayes)
library(dplyr)
library(tidyr)
library(ggplot2)

# Extract posterior draws
draws <- tidybayes::tidy_draws(fit) |>
  select(starts_with("b_")) |>
  mutate(across(everything(), exp))

# Rename for plotting
colnames(draws) <- gsub("b_", "", colnames(draws))

# True values for comparison
true_vals <- data.frame(
  par = c("drift_conditioneasy", "drift_conditionhard", "bound_Intercept", "ndt_Intercept"),
  value = c(3.0, 1.5, 1.5, 0.3)
)

# Plot
draws |>
  pivot_longer(everything(), names_to = "par", values_to = "value") |>
  filter(par != "Intercept") |>
  ggplot(aes(x = value, y = par)) +
  stat_halfeye(normalize = "groups", fill = "steelblue", alpha = 0.7) +
  geom_point(data = true_vals, aes(x = value, y = par),
             color = "red", shape = 18, size = 4) +
  labs(x = "Parameter value", y = "",
       title = "Posterior distributions of EZDM parameters",
       subtitle = "Red diamonds indicate true generating values") +
  coord_cartesian(xlim = c(0,5)) +
  theme_minimal()
```

# Comparing conditions

As `bmm` feeds seamlessly into `brms`, we can use Bayesian hypothesis testing via Savage-Dickey density ratios to compare drift rates between conditions:

```{r}
# Test if drift rate is higher in easy vs hard condition
brms::hypothesis(fit, "exp(drift_conditioneasy) > exp(drift_conditionhard)")
```

# Adding random effects

For hierarchical models with random effects across subjects:

```{r, eval=FALSE}
# Formula with random effects
model_formula_re <- bmf(
  drift ~ 0 + condition + (0 + condition | subject),
  bound ~ 1 + (1 | subject),
  ndt ~ 1 + (1 | subject)
)

fit_re <- bmm(
  formula = model_formula_re,
  data = sim_data,
  model = model,
  cores = 4,
  backend = "cmdstanr"
)
```

# References
