---
title: "Dual-process and meta-d′ signal detection models"
output: bookdown::html_document2
author:
  - Gidon Frischkorn
  - Ven Popov
bibliography: REFERENCES.bib
header-includes:
  - \usepackage{amsmath}
vignette: >
  %\VignetteIndexEntry{Dual-process and meta-d′ signal detection models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
pkgdown:
  as_is: true
---
<style type="text/css">
div.main-container {
max-width: 800px !important;
}

p {
margin-top: 1.5em ;
margin-bottom: 1.5em ;
}
.author{
    display: none;
}
</style>

```{r, include = FALSE}
options(crayon.enabled = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  dev = "jpeg",
  dpi = 100,
  fig.width = 5,
  fig.asp = 0.8,
  out.width = "80%",
  fig.align = "center"
)
fansi::set_knit_hooks(knitr::knit_hooks, which = c("output", "message", "error"))
```

```{r setup, message = FALSE}
library(bmm)
library(ggplot2)
```

# Beyond familiarity: two process models

The [confidence-rating SDT model](bmm_sdt.html) treats recognition as a single
continuous process: an item generates a sample of memory *strength* (or
*familiarity*), and the observer maps that one number onto an ordered confidence
scale. A persistent finding in recognition memory is that the resulting ROC is
**asymmetric** — on z-transformed axes its slope is below one. The core SDT
article reads that asymmetry as *unequal variance*: the old-item distribution is
simply wider than the new-item distribution.

Unequal variance is not the only explanation. Two influential models keep the
familiarity process but add a **second process** that also leaves its mark on the
ROC:

- **Dual-process SDT** [@Yonelinas_1994] adds a *recollection* process: besides
  graded familiarity, an item can be consciously recollected, which supports a
  confident, all-or-none response. Recollection is the textbook account of the
  recognition ROC's shape.
- **Meta-d′** [@Maniscalco_Lau_2012] adds a *metacognitive* read-out: it asks how
  well an observer's confidence tracks whether their decision was actually
  correct, separately from how well they discriminate old from new.

Both are *versions of the same rating model* — they consume identical data
(counts per confidence category, with a 0/1 `stimulus` column) and reduce to the
standard model when their extra parameter is switched off. They are selected with
the `version` argument of `sdt_rating()`:

| `version` | extra parameter(s) | captures | reduces to standard when |
|-----------|--------------------|----------|--------------------------|
| `"standard"` (default) | — | familiarity / strength only | always |
| `"dpsdt"` | `Ro`, `Rn` | recollection of old / new items | `Ro`, `Rn` → 0 |
| `"metad"` | `logmratio` | metacognitive efficiency (log M-ratio) | `logmratio` = 0 |

This article assumes the [core SDT article](bmm_sdt.html): the four-step
fitting workflow, the standard rating model, confidence thresholds, and the
`roc_sdt()` / `latent_sdt()` / `sdt_thresholds()` / `pp_check()` toolkit. We
focus here on what the second process adds — and because both versions are
plain `bmm` models, everything from that workflow (hierarchical formulas,
priors, post-processing) carries over unchanged.

# Dual-process SDT

## The model

Dual-process theory holds that a recognition decision can draw on two separable
sources [@Yonelinas_1994; @Wixted_Mickes_2010]. *Familiarity* is a graded
signal-detection process: old and new items generate overlapping Gaussian
evidence, exactly as in the standard model. *Recollection* is a high-threshold
process: with some probability the observer retrieves specific details that
settle the decision with high confidence, and otherwise recollection fails and
they fall back on familiarity.

`bmm` parameterizes recollection with two probabilities, both on the logit
scale:

- `Ro` — *recall-to-accept*: the probability $\mathrm{logit}^{-1}(\texttt{Ro})$
  that an **old** item is recollected as old, producing the most-confident "old"
  rating.
- `Rn` — *recall-to-reject*: the probability $\mathrm{logit}^{-1}(\texttt{Rn})$
  that a **new** item is recollected as new (its absence from the studied set is
  recollected), producing the most-confident "new" rating.

Classic recognition models keep only recall-to-accept; the symmetric
recall-to-reject term is motivated by change-detection and working-memory
paradigms, where recollecting that an item was *not* studied is itself
diagnostic [@Yonelinas_2023]. The probability of each confidence category is then
a mixture: a recollection component that piles mass on the extreme category, and
a $(1 - R)$-weighted familiarity component spread across the scale by the
thresholds.

```{r dpsdt-densities, echo = FALSE, fig.cap = "Dual-process model. Old and new items generate equal-variance familiarity evidence (curves); on a fraction Ro of old trials recollection succeeds and the item is reported with maximal 'old' confidence (right spike), and on a fraction Rn of new trials recall-to-reject yields maximal 'new' confidence (left spike). The recollection mass sits on top of the familiarity process, not on the evidence axis."}
dpr <- 1.2
x   <- seq(-3.5, 4.8, length.out = 600)
dens <- rbind(
  data.frame(x = x, density = dnorm(x, -dpr / 2, 1), dist = "Noise (new)"),
  data.frame(x = x, density = dnorm(x,  dpr / 2, 1), dist = "Signal (old)")
)
ggplot(dens, aes(x, density, color = dist, fill = dist)) +
  geom_line(linewidth = 0.8) +
  annotate("segment", x = 3.6, xend = 3.6, y = 0, yend = 0.30,
           color = "#377EB8", linewidth = 1.1,
           arrow = arrow(length = unit(0.12, "cm"))) +
  annotate("text", x = 3.6, y = 0.34, label = "italic(R[o])", parse = TRUE,
           color = "#377EB8") +
  annotate("segment", x = -2.7, xend = -2.7, y = 0, yend = 0.18,
           color = "#E41A1C", linewidth = 1.1,
           arrow = arrow(length = unit(0.12, "cm"))) +
  annotate("text", x = -2.7, y = 0.22, label = "italic(R[n])", parse = TRUE,
           color = "#E41A1C") +
  scale_color_brewer(palette = "Set1") +
  scale_fill_brewer(palette = "Set1") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(x = "Latent familiarity", y = "Density", color = NULL, fill = NULL) +
  theme_classic() +
  theme(legend.position = "top")
```

Because recollection adds mass at the *high-confidence* ends, its fingerprint is
a **curved** z-transformed ROC: a pure unequal-variance Gaussian produces a
*straight* zROC (slope $1/s$), whereas a recollection component bows it
[@Yonelinas_1994]. We return to this contrast below.

`Ro` and `Rn` are **fixed off by default** (a constant near $-\infty$ on the
logit scale, so recollection probability $\approx 0$ and the model is exactly the
standard rating model). You switch recollection on by adding it to the formula —
`Ro ~ 1` for the one-sided model (recollection of old items only) or
`Ro ~ 1, Rn ~ 1` for the two-sided model.

## Worked example: recognition memory

We use the same @Pratte_2010 recognition data as the core article (from the
`roc6` collection in the [MPTinR](https://cran.r-project.org/package=MPTinR)
package): 97 subjects judging old and new items on a 6-point confidence scale.

```{r}
stopifnot(requireNamespace("MPTinR", quietly = TRUE))
data("roc6", package = "MPTinR")
pratte <- roc6[roc6$exp == "Pratte_2010", ]

rating_data <- rbind(
  data.frame(id = pratte$id, stimulus = 1L,
             r1 = pratte$OLD_3new, r2 = pratte$OLD_2new, r3 = pratte$OLD_1new,
             r4 = pratte$OLD_1old, r5 = pratte$OLD_2old, r6 = pratte$OLD_3old),
  data.frame(id = pratte$id, stimulus = 0L,
             r1 = pratte$NEW_3new, r2 = pratte$NEW_2new, r3 = pratte$NEW_1new,
             r4 = pratte$NEW_1old, r5 = pratte$NEW_2old, r6 = pratte$NEW_3old)
)
```

The dual-process model attributes the ROC asymmetry to recollection, so we keep
the familiarity process *equal-variance* (we do not estimate `sdratio`) and free
both recollection terms:

```{r}
model_dp <- sdt_rating(response = paste0("r", 1:6), stimulus = "stimulus",
                       version = "dpsdt")

fit_dp <- bmm(
  formula = bmf(dprime ~ 1 + (1 | id), criterion ~ 1 + (1 | id),
                spacing ~ 1, Ro ~ 1, Rn ~ 1),
  data = rating_data,
  model = model_dp,
  backend = "cmdstanr",
  cores = 4,
  chains = 4,
  iter = 2000,
  refresh = 0,
  silent = 2,
  file = "assets/bmmfit_sdt_dpsdt_vignette"
)
summary(fit_dp)
```

The recollection terms are reported on the logit scale. `latent_sdt()` converts
them to probabilities and prints them alongside the familiarity distributions:

```{r}
latent_dp <- latent_sdt(fit_dp)
attr(latent_dp, "extra")
```

Recollection of old items (`Ro`) is credibly positive — it contributes to
high-confidence "old" responses over and above familiarity, the dual-process
signature. Recall-to-reject (`Rn`), by contrast, is estimated at essentially
zero. That is exactly what we should expect for standard old/new recognition:
recollecting that an item was *not* studied has little to work with here. Freeing
`Rn` does no harm — the two-sided fit simply recovers the classic one-sided model
— and it becomes diagnostic only in paradigms where the absence of an item is
itself recollectable, such as change detection and visual working memory
[@Yonelinas_2023].

## Two accounts of the same asymmetry

The core article fit these same data with an *unequal-variance* Gaussian
(`sdratio ~ 1`, no recollection). Both models reproduce the asymmetric
recognition ROC, but they tell different stories: unequal variance says the
old-item *strength distribution* is wider, while dual-process says a separate
*recollection process* adds high-confidence hits. We fit the unequal-variance
model here too, for a direct comparison:

```{r}
model_uv <- sdt_rating(response = paste0("r", 1:6), stimulus = "stimulus")

fit_uv <- bmm(
  formula = bmf(dprime ~ 1 + (1 | id), criterion ~ 1 + (1 | id),
                spacing ~ 1, sdratio ~ 1),
  data = rating_data,
  model = model_uv,
  backend = "cmdstanr",
  cores = 4,
  chains = 4,
  iter = 2000,
  refresh = 0,
  silent = 2,
  file = "assets/bmmfit_sdt_dp_uv_vignette"
)
```

On probability axes both ROCs bend through the operating points. The
**z-transformed** ROC (`scale = "z"`) is the diagnostic that separates them: the
unequal-variance Gaussian is a straight line, whereas the dual-process curve is
bowed by the recollection mass.

```{r dpsdt-zroc, fig.show = "hold", out.width = "49%", fig.cap = "The same recognition asymmetry, two accounts, on z-transformed ROC axes. Left: unequal-variance Gaussian — a straight zROC with slope below one. Right: dual-process — recollection bows the zROC. Points are the observed operating points."}
plot(roc_sdt(fit_uv), observed = roc_observed(fit_uv), scale = "z") +
  ggplot2::ggtitle("Unequal-variance Gaussian")
plot(roc_sdt(fit_dp), observed = roc_observed(fit_dp), scale = "z") +
  ggplot2::ggtitle("Dual-process (recollection)")
```

In practice the two are hard to separate from a single ROC: recognition zROCs
are only gently curved, and an unequal-variance Gaussian mimics the dual-process
shape closely (the same flexibility caveat that applies to the extreme-value vs
Gaussian comparison in the core article; @MeyerGrant_2025). The choice is best
driven by theory and by whether parameters stay invariant across conditions —
for instance whether recollection, but not familiarity, responds to a
manipulation [@Yonelinas_2023] — rather than by ROC fit alone. The continuous
dual-process model [@Wixted_Mickes_2010] and finite-mixture SDT [@DeCarlo_2002]
are further variations on the same theme of a second component shaping the ROC.

## Posterior predictive checks

As for the standard rating model, `pp_check()` compares the observed
rating-category proportions against the posterior predictive distribution,
faceted by `stimulus`:

```{r}
pp_check(fit_dp, ndraws = 100)
```

If recollection is doing real work, the dual-process model should track the
high-confidence categories (r1 and r6) more closely than an equal-variance
familiarity-only model, whose single strength axis cannot put enough mass at the
extremes.

# Meta-d′

## The model

A confidence-rating task answers two different questions at once. *Type-1*: how
well does the observer tell old from new? That is ordinary sensitivity,
`dprime`. *Type-2*: how well does the observer's **confidence** track whether
each decision was actually correct? That is *metacognitive* sensitivity, and it
need not match type-1 sensitivity — an observer can discriminate well yet place
confidence poorly, or vice versa [@Fleming_Lau_2014].

Meta-d′ [@Maniscalco_Lau_2012] expresses type-2 sensitivity in the same units as
type-1 `dprime`. It is the sensitivity that a metacognitively *ideal*
signal-detection observer would need in order to produce the observed confidence
data, **holding the type-1 decision fixed** — that is, the old/new response rates
are constrained to match what `dprime` already implies, so meta-d′ is identified
purely from the *spread of confidence* within each response. The natural summary
is the

$$
\text{M-ratio} = \frac{\text{meta-}d'}{\texttt{dprime}},
$$

the **metacognitive efficiency**: a value of 1 means confidence is as
informative as the decision (ideal), below 1 means confidence carries less
information than the decision used (the common case), and above 1 means
confidence draws on information the decision did not.

Rather than estimating meta-d′ directly, `bmm` estimates the **log M-ratio**,
`logmratio` $= \log(\text{meta-}d'/d')$, and recovers meta-d′ as
`exp(logmratio) * dprime`. This is the parameterization used by the HMeta-d
toolbox [@Fleming_2017]: the M-ratio is the quantity researchers actually report,
so making it the estimated parameter places the prior, hierarchical
between-subject SD, and any group contrasts directly on metacognitive efficiency.
It also keeps meta-d′ positive and regularizes it toward `dprime`, and it anchors
the ideal point — perfect metacognition, meta-d′ = `dprime` — at `logmratio = 0`,
which recovers the standard rating model. Add `logmratio` to the formula to
estimate it; extract the M-ratio posterior with `mratio()`.

```{r metad-resolution, echo = FALSE, fig.cap = "Metacognitive efficiency. Both panels have the same type-1 sensitivity (old/new discrimination) and the same old/new criterion (solid line). When metacognition is efficient (left, M-ratio = 1) the confidence thresholds (dashed) are well spread, so confidence separates likely-correct from likely-incorrect responses. When it is inefficient (right, M-ratio < 1) the thresholds bunch toward the criterion: confidence varies little and is weakly diagnostic of accuracy."}
make_panel <- function(metad, label) {
  dpr <- 1.6
  x   <- seq(-3.5, 5, length.out = 500)
  thr <- c(-1.4, -0.6, 0.6, 1.4) * (metad / dpr)  # type-2 criteria scale with metad
  d <- rbind(
    data.frame(x = x, density = dnorm(x, -dpr / 2, 1), dist = "Noise (new)",
               panel = label),
    data.frame(x = x, density = dnorm(x,  dpr / 2, 1), dist = "Signal (old)",
               panel = label)
  )
  list(dens = d, thr = data.frame(thr = thr, panel = label))
}
p1 <- make_panel(1.6, "Efficient (M-ratio = 1)")
p2 <- make_panel(0.8, "Inefficient (M-ratio = 0.5)")
dens <- rbind(p1$dens, p2$dens)
thr  <- rbind(p1$thr, p2$thr)
ggplot(dens, aes(x, density, color = dist)) +
  geom_line(linewidth = 0.8) +
  geom_vline(data = thr, aes(xintercept = thr), linetype = "dashed",
             color = "grey55") +
  geom_vline(xintercept = 0, linewidth = 0.7, color = "grey20") +
  facet_wrap(~ panel) +
  scale_color_brewer(palette = "Set1") +
  labs(x = "Latent evidence", y = "Density", color = NULL) +
  theme_classic() +
  theme(legend.position = "top")
```

## Worked example: metacognitive efficiency in recognition memory

We stay with the same @Pratte_2010 recognition data reshaped above
(`rating_data`). No new response format is needed: in a confidence-rating task
the 6-point scale is exactly the *joint* type-1/type-2 response — the side of
the scale is the old/new decision and the distance from the middle is
confidence — so the same counts that identified `dprime` also identify
metacognitive efficiency. Only the `version` changes:

```{r}
model_md <- sdt_rating(response = paste0("r", 1:6), stimulus = "stimulus",
                       version = "metad")

fit_md <- bmm(
  formula = bmf(dprime ~ 1 + (1 | id), criterion ~ 1 + (1 | id),
                spacing ~ 1, logmratio ~ 1 + (1 | id)),
  data = rating_data,
  model = model_md,
  backend = "cmdstanr",
  cores = 4,
  chains = 4,
  iter = 2000,
  refresh = 0,
  silent = 2,
  file = "assets/bmmfit_sdt_metad_pratte_vignette"
)
summary(fit_md)
```

The M-ratio is the headline quantity, so `bmm` provides it directly: `mratio()`
reconstructs meta-d′ $= \exp(\texttt{logmratio}) \times \texttt{dprime}$ from the
fit and returns a posterior summary of both the M-ratio and meta-d′ (with the
full per-draw posteriors kept in the `draws` attribute for plotting or contrasts).

```{r}
mratio(fit_md)
```

The population M-ratio sits a little below 1, the typical finding for recognition
memory: confidence tracks accuracy well but not perfectly, carrying slightly less
information than the old/new decision itself. The `(1 | id)` term on `logmratio`
lets metacognitive efficiency vary across subjects, so this single fit also gives
a posterior for *individual differences* in metacognition — the question that
motivates most meta-d′ analyses.

Two practical notes. First, recognition ROCs are classically asymmetric, which is
usually modelled as unequal variance; adding `sdratio ~ 1` to the formula fits
**unequal-variance meta-d′**, separating a genuine variance asymmetry from a
metacognitive one. Second, the model recovers known M-ratios from simulated
data: the package's development sources include a parameter-recovery script
that fits under- and over-confident observers generated by an independent code
path.

## Interpreting meta-d′

Meta-d′ is most useful when the substantive question is about *metacognition* —
whether confidence is well-calibrated to performance — rather than about
discrimination itself. Expressing type-2 sensitivity on the `dprime` scale makes
the M-ratio comparable across tasks and observers with different type-1
performance, which is why it has become the standard index of metacognitive
efficiency [@Fleming_Lau_2014; @Fleming_2017].

Two cautions carry over from the rest of SDT. First, like all the rating models
here, meta-d′ is fit to *aggregated counts*, so the type-2 estimate is only as
informative as the number of trials per confidence category. Second, meta-d′
inherits the standard model's assumptions about how confidence criteria are
placed; systematic `pp_check()` deviations flag when those assumptions strain.

### Confidence-threshold parameterization

In the canonical meta-d′ model the type-2 confidence criteria are *free*: their
spacing is exactly what makes the metacognitive fit informative. In `bmm` those
criteria are the rating thresholds, so the `threshold_type` you choose
([core SDT article](bmm_sdt.html)) governs how freely they can move. The
delta-based parameterizations (`"log_distance"`, `"log_ratio"`) leave each
threshold distance free and stay closest to the textbook meta-d′ estimator. The
two-parameter `"parsimonious"` (default) and `"equidistant"` types instead tie
the thresholds to a single spacing parameter: this regularizes the fit and is
often desirable with few trials per category, but it constrains the type-2 ROC
and can mask genuine asymmetries in how confidence is used for "old" versus "new"
responses. If the metacognitive question hinges on that asymmetry, prefer a
delta-based `threshold_type`. Whatever the parameterization, `sdt_thresholds()`
reconstructs the fitted confidence criteria on the latent scale — it works for
all three versions, so the type-2 criterion placements can be inspected
directly.

### How the type-1 criterion anchors confidence

Meta-d′ models must fix the type-2 confidence scale relative to the type-1
criterion. `bmm` uses the **absolute** convention — the confidence criteria are
anchored at the same boundary as the old/new decision (meta-$c = c$), as in the
HMeta-d toolbox [@Fleming_2017]. The literature also describes a **relative**
convention that scales the anchor with metacognitive efficiency
(meta-$c = M \cdot c$) [@Maniscalco_Lau_2012; @Maniscalco_Lau_2014]; `bmm` does
not currently expose that toggle. The two rarely diverge materially when the
criterion sits near the centre of the rating scale, but with a strongly biased
criterion the choice can shift meta-d′, so it is worth keeping in mind when
comparing estimates against toolboxes that use the relative anchor.

## Posterior predictive checks

The rating-profile check works as for the other rating models, faceted by
`stimulus`. For meta-d′ the categories to watch are the *interior* ones: the
model constrains the total "old"/"new" rates to match `dprime`, so any misfit
shows up in how confidence spreads within each response side. Systematic
one-sided deviations there suggest the confidence thresholds are too
constrained — see the threshold-parameterization note above.

```{r}
pp_check(fit_md, ndraws = 100)
```

# Continuous dual-process SDT

The dual-process model above treats recollection as a *threshold* process: an
old item is either recollected — landing in the most-confident category — or it
is not. The **continuous dual-process** (CDP) model [@Wixted_Mickes_2010] keeps
the dual-process intuition but makes recollection *continuous*. Every item
carries a familiarity signal $F$ and a recollection signal $R$, and the old/new
confidence rating is driven by their sum, the aggregate memory strength
$S = F + R$. What makes it a *dual*-process model is that the **Remember/Know
judgment is read off $R$ directly**: an "old" response is called *Remember* when
recollection clears a criterion $r$ on the $R$ axis, and *Know* otherwise.

This needs a richer response than the rating versions above — not just a
confidence rating, but a Remember/Know judgment on every "old" response — so bmm
provides it as a dedicated model, `sdt_cdp()`, rather than a `version` of
`sdt_rating()`. For a target, $F \sim \mathrm{Normal}(\mu_F, 1)$ and
$R \sim \mathrm{Normal}(\mu_R, \sigma_R)$ (lures fix both means to 0 and
$\sigma_R$ to 1); `dprimef` and `dprimer` are the two sensitivities and `sigmar`
is $\log \sigma_R$. Old/new confidence partitions the strength axis $S$ with
ordered thresholds, exactly as in the rating model, while the Remember/Know split
is the probability that $R$ exceeds `rcrit` *within* each confidence bin — a
region of the two-dimensional $(F, R)$ plane that bmm computes exactly with the
bivariate-normal CDF.

## Worked example: Remember/Know in recognition memory

We fit the model to the group Remember/Know ROC data of @Rotello_2005 (neutral
condition): a 6-point scale where rating 1 is a confident "new" response and
ratings 2--6 are "old" responses, each accompanied by a Remember or Know
judgment. We enter the data in **tidy long** form — one row per combination of
stimulus class, judgment, and confidence level, with a `count` column — since
that is how Remember/Know data typically leave an experiment (trial-level data
work the same way, just without `count`):

```{r}
rotello_long <- rbind(
  data.frame(stimulus = 1L, judgment = "new",      confidence = 1L,  count = 290),
  data.frame(stimulus = 1L, judgment = "know",     confidence = 2:6, count = c(122, 106, 101, 92, 59)),
  data.frame(stimulus = 1L, judgment = "remember", confidence = 2:6, count = c(26, 38, 61, 109, 433)),
  data.frame(stimulus = 0L, judgment = "new",      confidence = 1L,  count = 745),
  data.frame(stimulus = 0L, judgment = "know",     confidence = 2:6, count = c(199, 124, 84, 45, 13)),
  data.frame(stimulus = 0L, judgment = "remember", confidence = 2:6, count = c(46, 29, 47, 51, 55))
)
head(rotello_long)
```

Like `sdt_rating()`, the model itself is fit to **aggregated counts** — one row
per cell and one count column per response category.
`aggregate_sdt_cdp_data()` pivots the long rows into these canonical columns;
every column other than `judgment`, `confidence`, and `count` (here just
`stimulus`) defines the cells and is carried over:

```{r}
rotello <- aggregate_sdt_cdp_data(rotello_long, judgment = "judgment",
                                  confidence = "confidence", count = "count")
rotello
```

The 22 long rows collapse to one row per stimulus class, with columns `new1`
(the single "new" level) and `know2`...`know6` / `remember2`...`remember6` on
the unified confidence scale. This is also the format `rsdt_cdp()` simulates
directly, and if your own wide columns carry a shared prefix (say `cdp_new1`),
you declare it once via the constructor's `response` argument. The scale is
asymmetric — one "new" level and five "old" levels — which the constructor is
told via `n_new` and `n_old` (the two need not match). We free `sigmar` so
recollection can add variance to the
old-item strength distribution (the canonical Wixted--Mickes result) and keep
the familiarity--recollection correlation `rho` at its default of zero, the
classic independent-process model:

```{r}
model_cdp <- sdt_cdp(stimulus = "stimulus", n_new = 1, n_old = 5)

fit_cdp <- bmm(
  formula = bmf(dprimef ~ 1, dprimer ~ 1, criterion ~ 1, spacing ~ 1,
                rcrit ~ 1, sigmar ~ 1),
  data = rotello, model = model_cdp,
  backend = "cmdstanr", cores = 4, chains = 4, iter = 2000,
  refresh = 0, silent = 2,
  file = "assets/bmmfit_sdt_cdp_vignette"
)
summary(fit_cdp)
```

Recollection (`dprimer` ≈ 0.81) and familiarity (`dprimef` ≈ 0.78) contribute
comparable sensitivities, the Remember criterion sits well into the recollection
distribution (`rcrit` ≈ 0.92), and `exp(sigmar)` ≈ 1.48 puts the recollection SD
half again above one — the model's account of the recognition-ROC asymmetry as
*added recollection variance* rather than a uniformly wider strength
distribution. Despite using the two-parameter `parsimonious` thresholds rather
than five free criteria, these estimates reproduce the maximum-likelihood values
@Wixted_Mickes_2010 report for exactly these data (their Table 3:
$\mu_F = 0.78$, $\mu_R = 0.81$, $\sigma_R = 1.48$, $r = 0.92$).

Two further options extend the model, each opt-in. The
familiarity--recollection correlation is freed with `rho ~ 1` (and can itself be
predicted, e.g. `rho ~ condition`); it is bounded to $(-1, 1)$ internally and is
a *within-item* correlation, distinct from a between-subject correlation of
random effects (which you would write as `(1 |p| id)`). And when the task also
collects an explicit *Guess* response, include `"guess"` judgments in the long
data: `aggregate_sdt_cdp_data()` turns them into `guess` count columns, whose
presence switches on the three-way Remember/Know/Guess model with a familiarity
criterion `kcrit` separating Know from Guess.

# Choosing and comparing the models

The first three models share the same confidence-rating response interface and
the same familiarity core, so the choice is about which *additional question* the
data are meant to answer:

- Reach for **`version = "dpsdt"`** when the question is about *recollection* —
  whether a threshold retrieval process contributes beyond graded familiarity,
  as in recognition-memory and change-detection paradigms.
- Reach for **`version = "metad"`** when the question is about *metacognition* —
  whether confidence is efficiently calibrated to accuracy.
- Stay with **`version = "standard"`** when a single strength process suffices;
  both extensions reduce to it when their extra parameter is switched off
  (`Ro`, `Rn` → 0; `logmratio` = 0, i.e. meta-d′ = `dprime`).
- Reach for **`sdt_cdp()`** when the data carry explicit *Remember/Know*
  judgments: it models that richer response directly — recollection as a
  continuous second axis — rather than collapsing it to a confidence rating.

Because each extension nests the standard model, the most direct evidence is the
**posterior of the extra parameter**: is `Ro` credibly greater than zero, is
`logmratio` credibly different from zero (M-ratio different from 1)? That,
together with `pp_check()` and the
ROC/zROC fit, is the practical guide. A *formal* predictive comparison via
`loo()` is **not** appropriate for these aggregated-count models — leaving out one
multinomial observation removes a large, influential block of data and the
PSIS-LOO approximation becomes unreliable. The principled alternative is k-fold
cross-validation with `brms::kfold()`, which refits on held-out folds. As always,
let theory and parameter invariance across designs — not ROC fit alone — arbitrate
between accounts that can mimic one another [@MeyerGrant_2025; @Robinson_2023].

# References
