---
title: "The EZ-Diffusion Model (EZDM)"
output: bookdown::html_document2
author:
  - Gidon Frischkorn
  - Ven Popov
bibliography: REFERENCES.bib
header-includes:
  - \usepackage{amsmath}
vignette: >
  %\VignetteIndexEntry{The EZ-Diffusion Model (EZDM)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
pkgdown:
  as_is: true
---

<style type="text/css">
div.main-container {
max-width: 850px !important;
}

p {
margin-top: 1.5em ;
margin-bottom: 1.5em ;
}
.author{
    display: none;
}
</style>

```{r, include = FALSE}
options(crayon.enabled = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "jpeg",
  dpi = 100,
  fig.asp = 0.8,
  fig.width = 6,
  out.width = "80%",
  fig.align = "center"
)
fansi::set_knit_hooks(knitr::knit_hooks, which = c("output","message","error"))
```

# Introduction to the model

The diffusion model [@Ratcliff_1978] is one of the most influential models of speeded decision-making. It describes how participants accumulate noisy evidence over time until reaching one of two decision boundaries. Despite its theoretical elegance, the full diffusion model requires fitting to trial-level reaction time distributions, which can be computationally intensive and requires specialized software.

The **EZ-Diffusion Model** [@Wagenmakers_2007] provides a simpler alternative by establishing closed-form equations that relate the three core parameters of the diffusion model to easily computed summary statistics: mean reaction time (MRT), reaction time variance (VRT), and accuracy (proportion correct, Pc). This makes it possible to estimate diffusion model parameters from aggregated data without fitting to individual trials.

## The Diffusion Model

In the standard diffusion model, evidence accumulates from a starting point toward one of two boundaries. The key parameters are:

- **Drift rate ($v$)**: The average rate of evidence accumulation, reflecting task difficulty or ability
- **Boundary separation ($a$)**: The distance between the two decision boundaries, reflecting the speed-accuracy tradeoff
- **Non-decision time ($T_{er}$)**: Time for processes outside the decision (encoding, motor response)
- **Starting point ($z$)**: Where evidence accumulation begins relative to the boundaries

```{r diffusion-illustration, echo=FALSE, fig.cap="Illustration of the diffusion model. Evidence accumulates from the starting point until reaching the upper (correct) or lower (error) boundary.", out.width="70%"}
# Simple illustration of diffusion process
set.seed(42)
n_paths <- 5
n_steps <- 100
drift <- 0.05
bound <- 1

# Generate sample paths
paths <- matrix(0, n_steps, n_paths)
for (j in 1:n_paths) {
  for (i in 2:n_steps) {
    paths[i, j] <- paths[i-1, j] + drift + rnorm(1, 0, 0.15)
    if (paths[i, j] > bound || paths[i, j] < -bound) {
      paths[i:n_steps, j] <- paths[i, j]
      break
    }
  }
}

# Plot
par(mar = c(4, 4, 2, 1))
plot(1:n_steps, paths[,1], type = "l", ylim = c(-1.2, 1.2),
     xlab = "Time", ylab = "Evidence", col = "steelblue", lwd = 1.5,
     main = "Diffusion Model: Evidence Accumulation")
for (j in 2:n_paths) {
  lines(1:n_steps, paths[,j], col = "steelblue", lwd = 1.5)
}
abline(h = bound, col = "darkgreen", lwd = 2, lty = 2)
abline(h = -bound, col = "darkred", lwd = 2, lty = 2)
abline(h = 0, col = "gray50", lty = 3)
text(5, bound + 0.1, "Upper boundary (correct)", adj = 0, col = "darkgreen")
text(5, -bound - 0.1, "Lower boundary (error)", adj = 0, col = "darkred")
text(5, 0.1, "Starting point", adj = 0, col = "gray50")
```

## The EZ Equations

@Wagenmakers_2007 derived closed-form solutions relating the diffusion model parameters to observable summary statistics. For the **3-parameter version** (assuming symmetric starting point, $z = a/2$):

Given the proportion correct ($P_c$), mean RT of correct responses ($MRT$), and variance of correct RTs ($VRT$), the parameters can be computed as:

$$
v = \text{sign}(P_c - 0.5) \cdot s \cdot \left[ L(P_c) \cdot \frac{P_c^2 L(P_c) - P_c \cdot L(P_c) + P_c - 0.5}{VRT} \right]^{1/4}
$$

$$
a = s^2 \cdot \frac{L(P_c)}{v}
$$

$$
T_{er} = MRT - \frac{a}{2v} \cdot \frac{1 - e^{-\frac{va}{s^2}}}{1 + e^{-\frac{va}{s^2}}}
$$

where $L(P_c) = \ln\left(\frac{P_c}{1 - P_c}\right)$ is the logit function and $s$ is the diffusion constant (typically fixed to 1 for identification).

## The 4-Parameter Extension

The original EZ model assumes a symmetric starting point. @Srivastava_2016 derived the moments of first-passage times for the Wiener diffusion process with arbitrary starting points, enabling a **4-parameter version** that can estimate starting point bias ($z_r$, the relative starting point).

The 4-parameter version requires separate summary statistics for responses to each boundary:
- Mean and variance of RTs for upper boundary responses (typically correct)
- Mean and variance of RTs for lower boundary responses (typically errors)

This allows estimation of response bias, which is important when participants favor one response over another.

## Bayesian Estimation

@Chavez_Vandekerckhove_2025 introduced a Bayesian hierarchical approach to the EZ-diffusion model. Rather than using the closed-form equations to compute point estimates, this approach:

1. Treats the summary statistics as **observed data** generated from a known sampling distribution
2. Uses **likelihood-based inference** to estimate parameters with proper uncertainty quantification
3. Enables **hierarchical modeling** with random effects across participants

The `bmm` package implements this Bayesian approach, allowing researchers to:
- Estimate uncertainty in parameter estimates
- Include random effects for participants
- Compare conditions using Bayesian hypothesis testing
- Leverage the full power of the `brms` modeling framework

# Parametrization in the `bmm` package

The `bmm` package implements two versions of the EZ-diffusion model:

## 3-Parameter Version (`version = "3par"`)

Estimates three parameters with a fixed symmetric starting point ($z_r = 0.5$):

| Parameter | Description | Link Function |
|-----------|-------------|---------------|
| `drift` | Drift rate ($v$) - rate of evidence accumulation | log |
| `bound` | Boundary separation ($a$) - response caution | log |
| `ndt` | Non-decision time ($T_{er}$) - encoding + motor time | log |

The diffusion constant `s` is fixed to 1 for identification.

## 4-Parameter Version (`version = "4par"`)

Adds estimation of starting point bias:
| Parameter | Description | Link Function |
|-----------|-------------|---------------|
| `drift` | Drift rate ($v$) | log |
| `bound` | Boundary separation ($a$) | log |
| `ndt` | Non-decision time ($T_{er}$) | log |
| `zr` | Relative starting point (0-1 scale) | logit |

Note that `zr = 0.5` indicates no bias (symmetric starting point), `zr > 0.5` indicates bias toward the upper boundary, and `zr < 0.5` indicates bias toward the lower boundary.

# Data requirements

Unlike other models in `bmm` that work with trial-level data, the EZ-diffusion model requires **aggregated summary statistics**:

| Variable | Description |
|----------|-------------|
| `mean_rt` | Mean reaction time (in **seconds**) |
| `var_rt` | Variance of reaction times |
| `n_upper` | Number of responses hitting the upper boundary |
| `n_trials` | Total number of trials |

For the **3-parameter version**, `mean_rt` and `var_rt` should be computed from all correct responses pooled together.

For the **4-parameter version**, you need separate statistics for upper and lower boundary responses:
- `mean_rt_upper`, `var_rt_upper` for upper boundary (correct) responses
- `mean_rt_lower`, `var_rt_lower` for lower boundary (error) responses

# Preparing data with `ezdm_summary_stats()`

The `bmm` package provides the `ezdm_summary_stats()` function to convert trial-level data to the required aggregated format. This function also implements robust estimation via mixture modeling to handle contaminant responses (fast guesses, lapses).

## Basic usage

```{r, eval=FALSE}
library(bmm)

# Basic usage: compute statistics grouped by subject and condition
summary_data <- ezdm_summary_stats(
  data = trial_data,
  rt = "reaction_time",
  response = "correct",
  .by = c("subject", "condition")
)
```

## Robust estimation

By default, `ezdm_summary_stats()` uses mixture modeling to separate true responses from contaminants:

```{r, eval=FALSE}
# Robust estimation with mixture model
summary_data <- ezdm_summary_stats(
  data = trial_data,
  rt = "rt",
  response = "accuracy",
  .by = c("subject", "condition"),
  method = "mixture",              # Use mixture modeling (default)
  distribution = "exgaussian",     # RT distribution for true responses
  contaminant_bound = c(0.1, 3.0)  # Bounds for contaminant uniform distribution
)
```

The function returns the estimated contaminant proportion, which can be useful for quality control.

# Fitting the model with `bmm`

Begin by loading the `bmm` package:

```{r setup, message=FALSE, warning=FALSE}
library(bmm)
```

## Generating simulated data

For illustration, we will generate simulated data with known parameters using the `rezdm()` function:

```{r}
set.seed(123)

# Define true parameters for two conditions
true_params <- data.frame(
  condition = c("easy", "hard"),
  drift = c(3.0, 1.5),
  bound = c(1.5, 1.5),
  ndt = c(0.3, 0.3)
)

# Number of subjects and trials per condition
n_subjects <- 20
n_trials <- 100

# Generate data
sim_data <- data.frame()
for (subj in 1:n_subjects) {
  for (cond in 1:2) {
    # Add some random variation for each subject
    drift_subj <- true_params$drift[cond] + rnorm(1, 0, 0.3)
    bound_subj <- true_params$bound[cond] + rnorm(1, 0, 0.1)
    ndt_subj <- true_params$ndt[cond] + rnorm(1, 0, 0.02)

    # Generate summary statistics
    stats <- rezdm(
      n = 1,
      n_trials = n_trials,
      drift = max(drift_subj, 0.1),
      bound = max(bound_subj, 0.5),
      ndt = max(ndt_subj, 0.1),
      version = "3par"
    )

    stats$subject <- subj
    stats$condition <- true_params$condition[cond]
    sim_data <- rbind(sim_data, stats)
  }
}

# Convert condition to factor
sim_data$condition <- factor(sim_data$condition, levels = c("easy", "hard"))

head(sim_data)
```

## Specifying the formula

We specify the model formula using `bmmformula()` (or `bmf()`). For this example, we want to estimate separate drift rates for each condition while keeping boundary and non-decision time constant:

```{r}
model_formula <- bmf(
  drift ~ 0 + condition,
  bound ~ 1,
  ndt ~ 1
)
```

## Specifying the model

Next, we specify the EZDM model with the variable names from our dataset:

```{r}
model <- ezdm(
  mean_rt = "mean_rt",
  var_rt = "var_rt",
  n_upper = "n_upper",
  n_trials = "n_trials",
  version = "3par"
)
```

## Fitting the model

Now we can fit the model using `bmm()`:

```{r, eval=FALSE}
fit <- bmm(
  formula = model_formula,
  data = sim_data,
  model = model,
  cores = 4,
  backend = "cmdstanr",
  file = "assets/bmmfit_ezdm_vignette"
)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# For the vignette, we'll fit a quick model or load a cached one
fit <- bmm(
  formula = model_formula,
  data = sim_data,
  model = model,
  cores = 4,
  iter = 1000,
  backend = "cmdstanr",
  file = "assets/bmmfit_ezdm_vignette"
)
```

## Inspecting results

First, check the model summary:

```{r, comment=NA}
summary(fit)
```

All parameters use log link functions, so we need to exponentiate to get estimates on the natural scale:

```{r}
# Extract fixed effects
fixef_est <- brms::fixef(fit)

# Transform to natural scale
cat("Drift rate (easy):", exp(fixef_est["drift_conditioneasy", "Estimate"]), "\n")
cat("Drift rate (hard):", exp(fixef_est["drift_conditionhard", "Estimate"]), "\n")
cat("Boundary:", exp(fixef_est["bound_Intercept", "Estimate"]), "\n")
cat("Non-decision time:", exp(fixef_est["ndt_Intercept", "Estimate"]), "\n")
```
Compare these to the true generating values:

```{r}
print(true_params)
```

## Visualizing posterior distributions

```{r, message=FALSE, warning=FALSE, fig.width=7, fig.height=5}
library(tidybayes)
library(dplyr)
library(tidyr)
library(ggplot2)

# Extract posterior draws
draws <- tidybayes::tidy_draws(fit) %>%
  select(starts_with("b_")) %>%
  mutate(across(everything(), exp))

# Rename for plotting
colnames(draws) <- gsub("b_", "", colnames(draws))

# True values for comparison
true_vals <- data.frame(
  par = c("drift_conditioneasy", "drift_conditionhard", "bound_Intercept", "ndt_Intercept"),
  value = c(3.0, 1.5, 1.5, 0.3)
)

# Plot
draws %>%
  pivot_longer(everything(), names_to = "par", values_to = "value") %>%
  ggplot(aes(x = value, y = par)) +
  stat_halfeye(normalize = "groups", fill = "steelblue", alpha = 0.7) +
  geom_point(data = true_vals, aes(x = value, y = par),
             color = "red", shape = 18, size = 4) +
  labs(x = "Parameter value", y = "",
       title = "Posterior distributions of EZDM parameters",
       subtitle = "Red diamonds indicate true generating values") +
  theme_minimal()
```

# Comparing conditions

We can use Bayesian hypothesis testing to compare drift rates between conditions:

```{r}
# Test if drift rate is higher in easy vs hard condition
hypothesis(fit, "exp(drift_conditioneasy) > exp(drift_conditionhard)")
```

# Adding random effects

For hierarchical models with random effects across subjects:

```{r, eval=FALSE}
# Formula with random effects
model_formula_re <- bmf(
  drift ~ 0 + condition + (0 + condition | subject),
  bound ~ 1 + (1 | subject),
  ndt ~ 1 + (1 | subject)
)

fit_re <- bmm(
  formula = model_formula_re,
  data = sim_data,
  model = model,
  cores = 4,
  backend = "cmdstanr"
)
```

# References
