---
title: "The Log-Normal Race Model (LNR)"
output: bookdown::html_document2
author:
  - Gidon Frischkorn
bibliography: REFERENCES.bib
header-includes:
  - \usepackage{amsmath}
vignette: >
  %\VignetteIndexEntry{The Log-Normal Race Model (LNR)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
pkgdown:
  as_is: true
---

<style type="text/css">
div.main-container {
max-width: 800px !important;
}

p {
margin-top: 1.5em ;
margin-bottom: 1.5em ;
}
.author{
    display: none;
}
</style>

```{r, include = FALSE}
options(crayon.enabled = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "jpeg",
  dpi = 100,
  fig.asp = 0.8,
  fig.width = 6,
  out.width = "80%",
  fig.align = "center"
)
fansi::set_knit_hooks(knitr::knit_hooks, which = c("output","message","error"))
```

# Introduction to the model

The Log-Normal Race model [LNR; @rouderLognormalRaceCognitiveProcess2015] is
a race model for choice response time data. Each response option has its own
accumulator, and these accumulators race to completion. The first accumulator
to finish determines both the chosen response and the observed response time.

The LNR works with two or more response alternatives and has fewer parameters
than the Linear Ballistic Accumulator or the Diffusion Decision Model. This
makes it a good starting point for decomposing RT distributions, especially in
multi-alternative tasks.

## Model specification

Each response alternative $k$ has an accumulator whose finishing time $T_k$
follows a lognormal distribution:

$$
T_k \sim \text{Lognormal}(\mu_k, \sigma)
$$

where $\mu_k$ is the meanlog parameter (lower values produce faster finishing
times) and $\sigma$ is the sdlog parameter controlling variability. The
observed response and response time are:

$$
\text{Response} = \arg\min_k T_k, \quad RT = \min_k T_k + \tau
$$

where $\tau$ is the non-decision time for stimulus encoding and motor
execution. Since the lognormal distribution is positively skewed and defined
on the positive reals, it fits the shape of empirical RT distributions well.

# Parametrization in the `bmm` package

The `bmm` package provides two versions of the LNR.

## Simple version (`version = "simple"`)

The simple version uses two meanlog parameters: one for correct responses
and one shared across all error responses.

| Parameter | Description | Link function |
|-----------|-------------|---------------|
| `correct` | Meanlog for the correct accumulator | identity |
| `error`   | Meanlog for all error accumulators | identity |
| `ndt`     | Non-decision time | log |
| `s`       | Sdlog, shared across accumulators (fixed to 1) | log |

Responses must be integer-coded: 1 for correct, 2 through K for errors. The
total number of alternatives is specified via `n_alternatives`. Use this
version for standard accuracy-based designs where the distinction between
correct and error processing is the primary interest.

## Custom version (`version = "custom"`)

The custom version assigns a separate meanlog parameter to each response
category. Category names are defined in the formula and matched to character
labels in the response column of the data:

| Parameter | Description | Link function |
|-----------|-------------|---------------|
| one per category | Meanlog for that category's accumulator(s) | identity |
| `ndt`     | Non-decision time | log |
| `s`       | Sdlog, shared across accumulators (fixed to 1) | log |

The custom version also supports per-category `num_alternatives` for tasks
where categories contain different numbers of items (e.g., 1 target, 3
similar lures, 5 unrelated distractors). This works analogously to the
`num_options` argument in the M3 model.

## Identifiability

Like other race and diffusion models, the LNR needs a scale constraint. By
default, `s` (sdlog) is fixed to 1. You can unfix it if you constrain another
parameter instead, but the default is appropriate for most applications.

# Data requirements

## Simple version

The data must contain:

- An RT column with response times in **seconds**
- A response column with integer values: 1 = correct, 2:K = errors

```{r}
library(bmm)

set.seed(42)
dat_simple <- rlnr(n = 500, m = c(-1, 0), s = c(1, 1), ndt = 0.2)
str(dat_simple)
```

## Custom version

The data must contain:

- An RT column with response times in **seconds**
- A response column with **character labels** matching the formula parameters

```{r}
set.seed(42)
sim <- rlnr(n = 500, m = c(-1.5, -0.5, 0.5), s = 1, ndt = 0.15)
dat_custom <- data.frame(
  rt = sim$rt,
  resp = c("recall", "similar", "other")[sim$response]
)
str(dat_custom)
table(dat_custom$resp)
```

# Fitting the model

## Simple version: two-choice task

```{r, eval=FALSE}
model <- lnr(rt = "rt", response = "response", n_alternatives = 2)

formula <- bmf(
  correct ~ 1,
  error ~ 1,
  ndt ~ 1
)

fit <- bmm(formula, dat_simple, model,
           cores = 4, backend = "cmdstanr")
```

For a four-choice task, set `n_alternatives = 4`. Responses 2 through 4 all
share the `error` parameter:

```{r, eval=FALSE}
model_k4 <- lnr(rt = "rt", response = "response", n_alternatives = 4)
fit_k4 <- bmm(bmf(correct ~ 1, error ~ 1, ndt ~ 1),
               dat_k4, model_k4,
               cores = 4, backend = "cmdstanr")
```

## Custom version: named categories

```{r, eval=FALSE}
model_custom <- lnr(rt = "rt", response = "resp", version = "custom")

formula_custom <- bmf(
  recall ~ 1,
  similar ~ 1,
  other ~ 1,
  ndt ~ 1
)

fit_custom <- bmm(formula_custom, dat_custom, model_custom,
                   cores = 4, backend = "cmdstanr")
```

## Per-category accumulator counts

When categories have different numbers of racing accumulators, pass a named
integer vector to `num_alternatives`:

```{r, eval=FALSE}
model_numalt <- lnr(
  rt = "rt", response = "resp", version = "custom",
  num_alternatives = c(recall = 1, similar = 3, other = 5)
)
```

If the number of items per category varies across trials, pass column names
instead of integers (e.g.,
`c(recall = "n_recall", similar = "n_similar", other = "n_other")`).

## Hierarchical models

Both versions support random effects. For hierarchical models, we recommend
setting `adapt_delta = 0.95` or higher, because divergent transitions can
occur when the non-decision time parameter gets close to the fastest observed
RT.

```{r, eval=FALSE}
model_hier <- lnr(rt = "rt", response = "response", n_alternatives = 2)

formula_hier <- bmf(
  correct ~ 1 + condition + (1 + condition | id),
  error ~ 1 + (1 | id),
  ndt ~ 1 + (1 | id)
)

fit_hier <- bmm(formula_hier, dat_hier, model_hier,
                cores = 4, backend = "cmdstanr",
                control = list(adapt_delta = 0.95))
```

# Default priors

| Parameter | Prior | Scale |
|-----------|-------|-------|
| `correct` | `normal(-1, 0.5)` | identity (meanlog) |
| `error` | `normal(0, 0.5)` | identity (meanlog) |
| `ndt` | `normal(-2, 0.3)` | log (~0.14s on the natural scale) |
| `s` | `normal(0, 0.3)` | log (~1.0 on the natural scale) |

For the custom version, category parameters default to `normal(0, 0.5)` on
the identity scale. These priors are weakly informative and appropriate for
RT data in seconds. Check them with `default_prior()` and adjust as needed.

# References
