---
title: "Visual working memory change detection"
output: bookdown::html_document2
author:
  - Ven Popov
  - Gidon Frischkorn
bibliography: REFERENCES.bib
header-includes:
  - \usepackage{amsmath}
vignette: >
  %\VignetteIndexEntry{Visual working memory change detection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
pkgdown:
  as_is: true
---
<style type="text/css">
div.main-container {
max-width: 800px !important;
}

p {
margin-top: 1.5em ;
margin-bottom: 1.5em ;
}
.author{
    display: none;
}
</style>

```{r, include = FALSE}
options(crayon.enabled = TRUE, width = 800)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "jpeg",
  dpi = 100
)
fansi::set_knit_hooks(knitr::knit_hooks, which = c("output", "message", "error"))
```

# Overview

`bmm` provides four model constructors for visual working memory change
detection:

- `sdm_cd()` for the signal discrimination model
- `mixture2p_cd()` for the two-parameter mixture model
- `mixture3p_cd()` for the three-parameter mixture model with non-targets
- `imm_cd()` for the full interference measurement model

These constructors replace the older `task = "cd"` entry points. The legacy
calls are still accepted during the transition, but they now issue a
deprecation warning.

# Choosing a model

Use `sdm_cd()` when you want the leanest model for binary same/change judgments.
It is usually the fastest place to start and gives interpretable estimates of
memory strength (`c`) and precision (`kappa`).

Use `mixture2p_cd()` when you want a discrete-memory account with a target
memory probability and a guessing component.

Use `mixture3p_cd()` when non-target intrusions are theoretically important.
The distribution helpers accept interpretable probability inputs
`p_target` and `p_nontarget`, while the fitted model continues to estimate the
corresponding mixture-weight parameters through the `bmm` formula interface.

Use `imm_cd()` when interference from non-target features and distances is a
core scientific question. It is the richest and typically the slowest of the
four models.

# Required data

All change-detection models expect:

- a binary response coded `0 = same`, `1 = change`
- a `probe` feature in radians
- a `target` feature in radians

`mixture3p_cd()` and `imm_cd()` additionally require non-target features that
are centered relative to the target. `imm_cd()` also requires positive
non-target distances.

# Model specification

The examples below show the constructor surface for each model family.

```{r eval = FALSE}
library(bmm)

sdm_model <- sdm_cd(
  response = "resp",
  probe = "probe",
  target = "target"
)

mix2_model <- mixture2p_cd(
  response = "resp",
  probe = "probe",
  target = "target"
)

mix3_model <- mixture3p_cd(
  response = "resp",
  probe = "probe",
  target = "target",
  nt_features = c("nt1", "nt2"),
  set_size = "set_size"
)

imm_model <- imm_cd(
  response = "resp",
  probe = "probe",
  target = "target",
  nt_features = c("nt1", "nt2"),
  nt_distances = c("d1", "d2"),
  set_size = "set_size"
)
```

For likelihood inspection or simulation, the public distribution helpers are
vectorized over observations and accept matrix-like non-target inputs:

```{r}
library(bmm)

dmixture3p_cd(
  response = c(0, 1),
  probe = c(0, pi / 2),
  nt_features = matrix(c(1, -1, 0.5, -0.5), nrow = 2, byrow = TRUE),
  lure_idx = matrix(c(1, 1, 1, 0), nrow = 2, byrow = TRUE),
  kappa = c(5, 6),
  p_target = c(0.6, 0.7),
  p_nontarget = c(0.25, 0.15)
)
```

# Diagnostics and runtime

Short pilot fits are a good idea before committing to long hierarchical runs.
`cd_fit_diagnostics()` extracts the core NUTS diagnostics for a fitted
change-detection model:

```{r eval = FALSE}
fit <- bmm(
  formula = bmf(c ~ 1, kappa ~ 1),
  data = dat,
  model = sdm_cd(response = "resp", probe = "probe", target = "target"),
  backend = "rstan"
)

cd_fit_diagnostics(fit)
```

For side-by-side runtime checks across all four models, use the internal
benchmark script in `tests/internal/benchmark_vwm_cd_models.R`.

# Practical guidance

Start with a short run, inspect divergences and effective sample sizes, and
only then scale up the number of chains and iterations. In most applications,
`sdm_cd()` or `mixture2p_cd()` are sensible baselines, while `mixture3p_cd()`
and `imm_cd()` should be reserved for cases where non-target structure is
substantively important.
